---
title: "Causal simulation"
date: "2025-09-13"
---

```{r}
#| include: false

# lmtp's print methods indent every line of their output. knitr then indents the
# opening ::: cell-output container, and Pandoc renders that marker as literal
# text rather than as a div, so ":::" appears in the published page and PDF.
# Dedenting the captured output keeps the container flush left.
show_lmtp <- function(x) {
  out  <- capture.output(print(x))
  keep <- nzchar(trimws(out))
  if (any(keep)) {
    pad <- min(nchar(sub("[^[:space:]].*$", "", out[keep])))
    if (pad > 0) out <- sub(sprintf("^[[:space:]]{1,%d}", pad), "", out)
  }
  cat(out, sep = "\n")
}
```




Evans and Didelez (2024), "Parameterizing and simulating from causal models," address the inverse problem: given a causal model, what joint distribution does it imply? The usual direction is observational data to causal structure to parameter estimates; this paper starts from the structure and asks what data it would generate.

Consider the joint distribution $p(x,a,y)$, where $x$ is covariates, $a$ is treatment, and $y$ is outcome.


 
$$
\begin{aligned}
p(x,a,y) &= p(x,a)p(y|x,a) \\
         &= p(x,a)p_a(y|x) \\
         &= p(x,a)\frac{p_a(x,y)}{p_a(x)} \\
         &= p(x,a)\frac{p_a(x)p_a(y)c(x,y|a)}{p_a(x)} \\
         &= p(x,a)p_a(y)c(x,y|a)
\end{aligned}
$$ {#eq-causal-simulation-1}


The joint distribution factorizes into the joint of $(x,a)$, the interventional marginal $p_a(y) = p(y \mid do(a))$, and the copula $c(x,y \mid a)$. The second line is where the causal content enters, and it is not an algebraic identity: replacing the observational conditional $p(y \mid x,a)$ by the interventional $p_a(y \mid x)$ requires the causal model to make them agree -- consistency together with exchangeability given $x$, or the equivalent structural condition. For an arbitrary observed joint distribution the two differ, and the rest of the factorization inherits that assumption. The derivation starts from $p(y \mid x,a)$ and absorbs the $x$-dependence into the copula, leaving only the marginal $p_a(y)$. Each piece can be specified separately: $(x,a)$ as the "past," $p_a(y)$ as the marginal structural model, and the copula as the dependence structure between $x$ and $y$ given $a$.

A copula links a joint distribution to its marginals:

$$ p(x,y) = f(x)g(y)c(F(x),G(y)) $$ {#eq-causal-simulation-2}

where $F(x)$ and $G(y)$ are the marginal distribution functions of $x$ and $y$. Keep the two objects apart: the copula $C$ is a distribution function on $[0,1]^2$ with uniform marginals, and $c$ in @eq-causal-simulation-2 is its *density*, which is what multiplies $f(x)g(y)$.  It can be used to generate joint distributions from marginal distributions.

Here $x$, $a$ and $y$ can all be vectors.

## examples

What do we use this for?  We can use this to simulate data from a causal model, and then fit different models to the simulated data.  This allows to compare the performance of different models.  

### example 1

Suppose we have a causal graph like this:

```{r}
#| label: dag1
#| include: true
#| warning: false
#| cache: true
#| message: false
#| echo: false

library(ggplot2)
library(ggdag)

g <- dagify(
  Y ~ A,
  Y ~ X,
  A ~ X,
  exposure = "A",
  outcome = "Y",
  coords = list(x = c(A = 1,  X = 2 , Y = 3),
                y = c(A = 1,  X = 2, Y = 1))
)

ggdag(g) + 
  theme_dag() 
```

We need to specify $P(y|do(a)) = \sum_x P(x) P(y|x,a)$, where $P(x)$ is the marginal distribution of $x$, and $P(y|x,a)$ is the conditional distribution of $y$ given $x$ and $a$.  This is the "marginal structural model" (Robins, 2000).  Also the "g-formula".  

We also need to specify $P(x,a)$, the "past".

Finally we need to specify the copula $c(x,y|a)$, which captures the dependence between $x$ and $y$ given $a$. Depending on different situations, different copula can be used.

In the "causl" package example, we specify:
$$ X \sim N(0,1) $$ {#eq-causal-simulation-3}
$$ A | X=x  \sim N(x/2, 1) $$ {#eq-causal-simulation-4}
$$ Y | do(A=a) \sim N({(a-1)/2}, 1) $$ {#eq-causal-simulation-5}
and Gaussian copula with correlation $\rho = 2 expit(1)-1$.

```{r}
#| label: sim0
#| include: true
#| warning: false
#| cache: true
#| message: false
#| echo: true

library(causl)
# formulae corresponding to covariates, treatments, outcomes and the dependence
forms <- list(X ~ 1, A ~ X, Y ~ A, ~ 1)
# vector of model families (3=gamma/exponential, 1=normal/Gaussian)
fam <- c(1, 1, 1, 1)
# list of parameters, including 'beta' (regression params) and 'phi' dispersion
pars <- list(X = list(beta=0, phi=1),   
             A = list(beta=c(0,0.5), phi=1),
             Y = list(beta=c(-0.5,0.5), phi=1),
             cop = list(beta=1))

## now create a `causl_model` object
cm <- causl_model(formulas=forms, family=fam, pars=pars,method="inversion")

# now simulate 1000 observations
set.seed(123456)
data <- rfrugal(n=1000, causl_model=cm)
head(data)
```

In this example, first we specify the structure of the model using a list of formulas.  The first formula is for the covariates, the second for the treatments, the third for the outcomes, and the fourth for the dependence structure (copula).  Note that outcome $Y$ only depends on $A$ in this interventional distribution.

Then we specify the families of the random variables.  Here we use normal distribution for $X$, normal distribution for $A$ and $Y$, and Gaussian copula for the dependence structure.

Finally we specify the parameters of the model, including regression coefficients and dispersion parameters.  For example, $Y \sim A$ has two coefficients, intercept -.5 and slope .5.  

The `causl_model` object is then created using these components.

In this simple example, if we run a regression with $A$, $X$ in the model, on the simulated data, we would get the correct effect back:

```{r}
#| label: sim1
#| include: true
#| warning: false
#| cache: true
#| message: false
#| echo: true

lm1 <- lm(Y ~ A*X, data=data)
summary(lm1)
```

The true interventional model is $Y \mid do(A=a) \sim N((a-1)/2, 1)$, so the
true intercept is $-0.5$ and the true coefficient on $A$ is 0.5. The naive
regression of $Y$ on $A$ and $X$ gives $A = 0.493$ and an intercept of $-0.492$,
with the interaction $A{:}X$ at $-0.009$ and insignificant ($p = 0.68$), which is
correct — the DGP has no such interaction.

Note this regression *also* includes $X$, which is the adjustment the g-formula
calls for. Its coefficient of 0.468 is not a causal quantity; it is there to
close the backdoor.

We can also use MLE:

```{r}
#| label: sim2
#| include: true
#| warning: false
#| cache: true
#| message: false
#| echo: true

out <- fit_causl(data, formulas = list(X ~ 1, Y ~ A, ~ 1), family = c(1, 1, 1))
out
```

`fit_causl` recovers the same structure directly from the frugal
parameterisation: intercept $-0.491$ and $A$ at 0.494, against the true $-0.5$
and 0.5, with a copula parameter of 0.998 against the specified 1. The sandwich
standard errors barely differ from the model-based ones, which is what should
happen when the model is correctly specified.

### Plasmode simulation

In reality, we have a data set and we have a causal model.  We want to simulate data from the causal model, but we want the simulated data to have the same distribution as the original data.  This is called "plasmode simulation".  For example, we don't want to use the real outcome variable.  Instead, we simulate X's from the original data, Then simulate Y and A from the causal model.  We can then test which method works better in estimating the causal effect of $A$ on $Y$.

```{r}
#| label: plasmode
#| echo: false

library(aciccomp2016)
library(dplyr)
(dat <- as_tibble(input_2016)) # show 10 rows of first few variables

```

```{r}
#| label: sim3
#| include: true
#| warning: false
#| cache: true
#| message: false
#| echo: true


# Model for the causal effect of smoking on birthweight
forms <- list(list(),
              list(A ~ x_1 + x_3 + x_4),
              list(Y ~ A),
              list(~ 1))
# fams <- list(integer(0), 5, 1, 1)
fams <- list(integer(0), "binomial", "gaussian", 1)
pars <- list(A = list(beta=c(-1.5,0.03,0.02,0.05)),
             Y = list(beta=c(3200, -500), phi=400^2),
             cop = list(beta=-1))

cm2 <- causl_model(formulas=forms, family=fams, pars=pars, dat = dat)
 
set.seed(123456)
data2 <- rfrugal(causl_model=cm2)
head(data2)

# we can also use rfrugalParam(), which is older version of rfrugal().
#datAY <- rfrugalParam(formulas=forms, family=fams, pars=pars, dat=dat)

```

In this specification, we have a binary treatment $A$ and a continuous outcome $Y$.  The covariates are $x_1$, $x_3$ and $x_4$.  As in the first example above, the copula's `beta` parameter maps to the actual Gaussian correlation via $\rho = 2\,\text{expit}(\beta) - 1$; with `beta=-1` this gives $\rho = 2\,\text{expit}(-1) - 1 \approx -0.46$, not $-1$ (a correlation of exactly $-1$ would require $\beta \to -\infty$ and would make the joint distribution singular).  We can then simulate the data.

## a more complicated model

Let's simulate a more complicated model:
```{r}
#| label: dag2
#| include: true
#| warning: false
#| cache: false
#| message: false
#| echo: true

library(ggplot2)
library(ggdag)

g <- dagify(
  A1 ~ L1,
  L2 ~ L1,
  L2 ~ A1,
  A2 ~ A1,
  A2 ~ L2,
  Y ~ A2,
  Y ~ L2,
  Y ~ A1,
  exposure = "A2",
  outcome = "Y",
  coords = list(x = c(L1 = 1,  L2 = 2, Y = 3, A1 = 1.5, A2 = 2.5),
                y = c(L1 = 1,  L2 = 1, Y = 1, A1 = 1.5, A2 = 1.5))
)

ggdag(g) + 
  theme_dag() 
```
Let's see how to write a causal model:

```{r}
#| label: sim4
#| include: true
#| warning: false
#| cache: true
#| message: false
#| echo: true

# formulae corresponding to covariates, treatments, outcomes and the dependence
forms <- list(list(L1  ~ 1, L2 ~ L1*A1),   # covariates
              list(A1 ~ L1, A2 ~ L2*A1), # treatments
              Y ~ A1*A2*L2,             # outcome
              ~ 1)

# vector of model families (3=gamma/exponential, 1=normal/Gaussian)

fams <- list(c(1, 1), c(5,5), 1, 1)
pars <- list(L1 = list(beta=0, phi=1),
             L2 = list(beta=c(0.3,0.5,-0.2,-0.1), phi=1),
             A1 = list(beta=c(-0.3,0.4), phi=1),
             A2 = list(beta=c(0.5,0.3,0.1,0), phi=1),
             Y = list(beta=c(0,1,2,0.5,1,0.2,0,0.8), phi=1), 
             cop = list(beta = 0.5))

# we can use rfrugalParam() to simulate data from the model, or rfrugal().
#dat <- rfrugalParam(n=1e4, formulas=forms, family = fams, pars=pars)
#head(dat)

cm3 <- causl_model(formulas=forms, family=fams, pars=pars)
 
set.seed(123456)
data3 <- rfrugal(n=1e4,causl_model=cm3)
head(data3)


```

If we have the correct specification for the outcome model, we'll get it right by linear model:

```{r}
#| label: sim5
#| include: true
#| warning: false
#| cache: true
#| message: false
#| echo: true

m1 <- glm(Y ~ A1*A2*L2, data = data3)
summary(m1)

library(marginaleffects)
avg_comparisons(
    model = m1,
    variables = "A2")
```

The fitted model has $A_2 = 2.023$, $A_1{:}A_2 = 0.983$ and
$A_1{:}A_2{:}L_2 = 0.835$, so the effect of $A_2$ depends on both $A_1$ and
$L_2$ and cannot be read off any single coefficient. `avg_comparisons` averages
over the observed distribution and returns 2.51 with a standard error of 0.020.

So the ATE is 2.5.

Let's use "lmtp" package to estimate the treatment effect of $A_2$ on $Y$.
```{r}
#| label: lmtp
#| include: true
#| warning: false
#| cache: true
#| message: false
#| echo: true


library(lmtp)

d1 <- function(data, trt) {
  rep(1, nrow(data))
}
A <- "A2"
Y <- "Y"
W <- c("L1", "L2", "A1")

set.seed(34465)

treat <- lmtp_tmle(
  data = data3, 
  trt = "A2", 
  outcome = "Y", 
  baseline = W, 
  outcome_type = "continuous", 
  shift = d1, 
  folds = 1, 
  learners_trt = "SL.glm", 
  learners_outcome = "SL.glm"
)

show_lmtp(treat)

d2 <- function(data, trt) {
  rep(0, nrow(data))
}
control <- lmtp_tmle(
  data = data3, 
  trt = "A2", 
  outcome = "Y", 
  baseline = W, 
  outcome_type = "continuous", 
  shift = d2, 
  folds = 1, 
  learners_trt = "SL.glm", 
  learners_outcome = "SL.glm"
)

show_lmtp(control)

show_lmtp(lmtp_contrast(treat, ref = control))

```


