---
title: "Correlated Random Effect"
date: "2025-04-15"
---



## Panel data

For panel data, the usual set up is:

$$y_{it} = \alpha + X_{it} \beta + v_i + \epsilon_{it} $$

## Fixed effect

A fixed effect model can be done with OLS on 

$$ (y_{it} - \bar y_i + \bar {\bar y}) = \alpha + (X_{it} - \bar X 
+ \bar {\bar X}) \beta + (\epsilon_{it} - \bar \epsilon_i + \bar v) + \bar {\bar \epsilon}$$
This is basically OLS on de-meaned $Y$ and $X$.

## Random effect

The random effect looks at this at different angle:  it treated $v_i + \epsilon_{it}$ as the error term.  There are two components of the error term.  Suppose they are estimated as $\hat \sigma^2_e$ (idiosyncratic component), and $\hat \sigma^2_u$ (individual component).  Then we can do GLS transformation:

$z^*_{it} = z_{it} - \hat \theta_i \bar z_i$

and $\hat \theta_i = 1 - \sqrt{\frac{\hat \sigma^2_e}{T_i \hat \sigma^2_u + \hat \sigma^2_e}}$

where $T_i$ is the number of observations for individual $i$.

Given estimates of  $\hat \sigma^2_e$  and $\hat \sigma^2_u$, we can run OLS on transformed variables (including $y$ and all $X$'s).  We can iterate the process.

## Correlated random effect

Correlated random effect (CRE) can be done by running a random effect model of $y_{it}$ on $X_{it}$, $\bar X_i$, $z_i$ and a constant.  

The shortcoming of RE model is that it has to assume $v_i$ and $X_{it}$ are uncorrelated to have a consistent estimator.  If that is not true (most eocnomists don't think it is), then we have an inconsistent estimator.  Meanwhile, FE estimator is consistent, because $v_i$ is not in the error term, it gets cancelled out.

The disadvantage of FE model is that it cannot include any time-invariant covariate, such as race, gender, etc.

The benefit of CRE is that it can include $z_i$'s which are time-invariant, while remain consistent.  In fact, Mundlak and Wooldridge pointed out that CRE estimates on $X_{it}$ are the same as FE estimates.

There is also a Mundlak test to choose between RE or CRE.


## Example

Stata 19 implemented CRE in the `xtreg` command, with "cre" option.  The following example is from Stata's website, using the `nlswork` dataset.

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#| include: false

library(Statamarkdown)
stataexe <- find_stata()
#stataexe <- "/usr/local/bin/stata"
knitr::opts_chunk$set(engine.path=list(stata=stataexe))
```

```{stata}
*| label: stata1
*| echo: true
*| cache: true
*| collectcode: true

webuse nlswork

xtreg ln_wage tenure age i.race, cre vce(cluster idcode)
```

To compare with a fixed effect model:

```{stata}
*| label: stata2
*| echo: true
*| cache: true
*| collectcode: true

webuse nlswork

xtreg ln_wage tenure age i.race, fe vce(cluster idcode)
```

We see the coefficient estimates are the same for "tenure" and "age", but CRE model allows you to estimate the effect of "race".

We can also manually do it by using a RE model on $X$, $\bar X$ and $z$:

```{stata}
*| label: stata3
*| echo: true
*| cache: true
*| collectcode: true

webuse nlswork
egen age_mean = mean(age), by(idcode)
egen tenure_mean = mean(tenure), by(idcode)
xtreg ln_wage tenure tenure_mean age age_mean i.race, vce(cluster idcode)
```

This is what stata's "cre" option is doing behind the scene.  





