---
title: "Correlated Random Effect"
date: "2025-04-15"
---



## Panel data

The standard panel-data setup is:

$$y_{it} = \alpha + X_{it} \beta + v_i + \epsilon_{it} $$ {#eq-correlated-random-effect-1}

## Fixed effect

The fixed-effect estimator runs OLS on demeaned data:

$$ (y_{it} - \bar y_i) = (X_{it} - \bar X_i) \beta + (\epsilon_{it} - \bar \epsilon_i)$$ {#eq-correlated-random-effect-2}
Both $\alpha$ and $v_i$ drop out after demeaning by unit.

## Random effect

Random effects treat $v_i + \epsilon_{it}$ as a composite error with two variance components: $\hat \sigma^2_e$ (idiosyncratic) and $\hat \sigma^2_u$ (individual). The GLS transformation is

$z^*_{it} = z_{it} - \hat \theta_i \bar z_i$

with $\hat \theta_i = 1 - \sqrt{\hat \sigma^2_e / (T_i \hat \sigma^2_u + \hat \sigma^2_e)}$, where $T_i$ is the number of observations for unit $i$. Given estimated variance components, we run OLS on the transformed variables and iterate.

## Correlated random effect

The correlated random-effect (CRE) model runs a random-effects regression of $y_{it}$ on $X_{it}$, $\bar X_i$, time-invariant covariates $w_i$, and a constant.

RE requires $v_i$ and $X_{it}$ to be uncorrelated — an assumption most economists reject. FE is consistent regardless, but it cannot estimate time-invariant covariates (race, gender, etc.). CRE has both advantages: it includes time-invariant variables and remains consistent. Mundlak and Wooldridge show that the CRE coefficients on $X_{it}$ are identical to the FE estimates. Testing whether the coefficients on $\bar X_i$ are jointly zero gives the Mundlak test for choosing between RE and CRE.


## Example

Stata 18 added a `cre` option to `xtreg`. The following uses `nlswork`.

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#| include: false

library(Statamarkdown)
stataexe <- find_stata()
#stataexe <- "/usr/local/bin/stata"
knitr::opts_chunk$set(engine.path=list(stata=stataexe))
```

```{stata}
*| label: stata1
*| echo: true
*| collectcode: true

webuse nlswork

xtreg ln_wage tenure age i.race, cre vce(cluster idcode)
```

To compare with a fixed effect model:

```{stata}
*| label: stata2
*| echo: true
*| collectcode: true

webuse nlswork

xtreg ln_wage tenure age i.race, fe vce(cluster idcode)
```

The coefficients on `tenure` and `age` are identical. CRE additionally estimates the effect of `race`.

We can replicate this manually by including the group means in a RE model:

```{stata}
*| label: stata3
*| echo: true
*| collectcode: true

webuse nlswork
* Restrict the mean calculation to the estimation sample (listwise-deleted
* on the model's variables), matching what xtreg's native cre option does
* internally. Both tenure and age have missing values in nlswork, so
* computing age_mean/tenure_mean over all rows for an idcode (rather than
* only rows with complete data on ln_wage, tenure, age, and race) would
* make the manual means diverge from cre's.
egen age_mean = mean(age) if !missing(ln_wage, tenure, age, race), by(idcode)
egen tenure_mean = mean(tenure) if !missing(ln_wage, tenure, age, race), by(idcode)
xtreg ln_wage tenure tenure_mean age age_mean i.race, vce(cluster idcode)
```

This is what Stata's `cre` option does internally.




