---
title: "Pre-registration, TOST, and why Bayes doesn't let you skip the hard part"
date: "2026-07-24"
---

## Why this came up

A colleague asked me to look over the analysis plan for a paper being submitted as a Registered Report (RR). Three of the hypotheses were the interesting kind: the authors expected an effect, but part of what they wanted to claim was that a *moderator* of that effect was absent, or negligible, in some subgroup. The draft's plan was to run the usual test, and if it came back non-significant, report that as evidence the moderator doesn't matter.

That doesn't work, and the journal's own guidelines say so explicitly. This post is about why it doesn't work, what the fix actually is (frequentist equivalence testing, or a Bayes factor), and a point I kept coming back to while thinking this through: the two "fixes" don't actually free you from the hard judgment call you were trying to avoid. They just relocate it.

## A non-significant result is not "no effect"

Ordinary null hypothesis significance testing (NHST) tests $H_0: \theta = 0$ against $H_1: \theta \neq 0$. A non-significant result ($p > 0.05$) tells you only that you couldn't reject zero. There are two very different reasons that could happen:

1. the true effect really is about zero, or
2. the effect is real, but the data are too noisy or the sample too small to detect it.

Both give you the same large $p$-value, and nothing in the test itself tells you which one you're looking at. This is the reductio I like to use to make the point concrete: you could "prove" almost any effect is zero just by collecting a tiny sample. A small $n$ guarantees a wide confidence interval (CI) and non-significance, regardless of what's actually true. A non-significant result is silence, not evidence of absence.

## What the journal actually requires

Registered Reports grant *in-principle acceptance* before the results exist — the journal commits to publishing the paper whatever the outcome. That's the mechanism that kills publication bias and p-hacking. But it creates an obligation: if the result is a null, the journal has to publish it, and an uninterpretable null would defeat the whole point. So Nature Human Behaviour's submission guidelines say, in essentially so many words:

> You cannot interpret lack of evidence (e.g. a $p > 0.05$ in a $t$-test) as evidence for the absence of an effect. Where authors intend to interpret a negative result as evidence that an effect is absent, they must propose an inferential method capable of drawing such a conclusion — Bayesian hypothesis testing or frequentist equivalence testing.

This isn't a Nature-specific quirk. The same requirement, close to word for word, appears in the Registered Report guidelines at PNAS Nexus, Royal Society Open Science, and most other journals that run the format — it traces back to a shared template from the Center for Open Science. It also comes with a power requirement most people don't expect: proposals should be powered to detect *the smallest effect that is plausible and of theoretical value*, with a priori power of at least 0.9 for frequentist plans, and pilot data alone is explicitly called out as "unlikely to form an acceptable basis" for choosing that effect size on its own. I'll come back to why that specific phrase matters.

## Equivalence testing (TOST), plainly

Equivalence testing flips the question from "is the effect exactly zero" to "can I rule out an effect large enough to matter." You pre-specify a bound $\Delta$ — the smallest effect size of interest (SESOI), the largest effect you'd still call negligible — and run two one-sided tests (TOST):

$$
H_{01}: \theta \le -\Delta \qquad \text{vs.} \qquad H_{02}: \theta \ge +\Delta.
$$

Reject both, and the effect lies inside $[-\Delta, +\Delta]$. Operationally this collapses to one check: build the $(1-2\alpha)$ CI and see whether it sits entirely inside the bounds. There's no separate machine computing something new here — TOST is a decision rule wrapped around a CI you already had.

Here's a worked example using `mtcars`, purely to show the mechanics — nobody should take conclusions about 1974 cars seriously. Is the quarter-mile time (`qsec`) meaningfully different between automatic and manual transmissions?

```{r}
#| message: false
library(TOSTER)
data(mtcars)
mtcars$am <- factor(mtcars$am, labels = c("automatic", "manual"))

t.test(qsec ~ am, data = mtcars)
```

Not significant ($p = 0.21$). Under plain NHST, that's the whole story: "no evidence of a difference." Now suppose I decide, for the sake of illustration, that anything under one second doesn't matter ($\Delta = 1$):

```{r}
t_TOST(formula = qsec ~ am, data = mtcars, eqb = 1, var.equal = FALSE)
```

TOST also fails to declare equivalence — the 90% CI is $[-0.27, 1.91]$, and it spills past $+1$. So both the plain test and TOST agree here: the data just aren't informative, in either direction. That's expected, and it's important to see: TOST doesn't rescue an underpowered comparison. Now widen the bound to $\Delta = 2$, using the exact same data:

```{r}
t_TOST(formula = qsec ~ am, data = mtcars, eqb = 2, var.equal = FALSE)
```

Now the CI sits entirely inside $[-2, 2]$, and TOST declares equivalence. The plain $t$-test's $p$-value hasn't moved at all — it's still 0.21 either way, because it never looked at $\Delta$ in the first place. That's the whole value of TOST in one pair of examples: NHST can't distinguish "underpowered nonsense" from "precisely estimated and negligible," because it never asked the second question. TOST can, but only once you've told it what "negligible" means.

## The Bayesian alternative

A Bayes factor (BF) compares the marginal likelihood of the data under $H_1$ versus $H_0$:

$$
\text{BF}_{10} = \frac{p(\text{data} \mid H_1)}{p(\text{data} \mid H_0)}.
$$

Conventionally, $\text{BF}_{10} < 1/3$ counts as evidence for the null, $\text{BF}_{10} > 3$ for the alternative, and anything between is inconclusive. The default JZS prior puts a Cauchy distribution on the standardized effect under $H_1$.

```{r}
#| message: false
library(BayesFactor)
ttestBF(formula = qsec ~ am, data = mtcars)
```

$\text{BF}_{10} \approx 0.64$: mildly favors the null, but not by much — inconclusive by the usual thresholds. Notice this default BF is answering a slightly different question than TOST did: it compares "no effect at all" to "some effect, of unknown size, drawn from a Cauchy prior," not "effect inside $\pm 2$" to "effect outside $\pm 2$." If you want the actual Bayesian analogue of an equivalence claim, you need an interval null:

```{r}
ttestBF(formula = qsec ~ am, data = mtcars, nullInterval = c(-2, 2))
```

The first line is the Bayes factor for "the effect is inside $\pm 2$" against the point null; the second is for "the effect is outside $\pm 2$." Their ratio is overwhelmingly in favor of *inside* — 0.82 against $6.2\times 10^{-6}$, tens of thousands to one. That's the Bayesian version of what TOST just told us at $\Delta = 2$: same data, same bound, same conclusion, different formalism.

## The hard decision doesn't disappear under Bayes — it only hides

This is the part I think gets missed. People often reach for a Bayes factor specifically because they don't want to defend a $\Delta$. But look at what just happened: to get a Bayesian claim that actually resembles "no meaningful effect," I had to hand `nullInterval` the exact same $\pm 2$ I used for TOST. The judgment didn't go away.

Even the *default*, no-interval Bayes factor doesn't escape it — it just moves the judgment somewhere less visible. The Cauchy prior's scale parameter ($r = \sqrt{2}/2$ by default) encodes an assumption about what effect sizes are plausible under $H_1$, and that assumption drives whether $\text{BF}_{10}$ ends up favoring the null. Change the prior scale and you change the verdict, on the same data. Nobody escapes stating (or silently assuming) how big an effect has to be before it counts. TOST just makes you write the number down and defend it, in public, before you've seen the data. A Bayes factor lets you not write it down — which feels like an advantage, right up until a reviewer asks about your prior's sensitivity.

There isn't even a neutral option here. Plain NHST also makes this choice — it just makes it silently, by testing against exactly zero, which is a default nobody questions rather than a threshold anyone defended.

## So how do you actually justify $\Delta$?

This is genuinely the hard part, and I don't think there's a way around doing real work here. The general playbook (Lakens, Scheel & Isager, 2018) gives three routes:

1. **Just-noticeable difference** — set $\Delta$ to the smallest change that would actually be perceptible or consequential in the world. Works well when there's a natural perceptual or operational unit (a price difference a customer would notice; a response-time difference a user would feel).
2. **Cost-benefit analysis** — decide what magnitude would actually be worth acting on, given the stakes. This is a substantive argument, not a formula, and it's field- and context-specific by nature — a "negligible" effect in one domain can be highly consequential in another, and that's appropriate, not a flaw.
3. **Small telescopes** (Simonsohn, 2015) — anchor $\Delta$ to some fraction (Simonsohn used 33%) of the effect size a comparable prior study's sample size could have reliably detected. This one actually has a formula behind it, but it only works if a genuinely comparable prior study exists.

What doesn't work, and I've caught myself reaching for it: deriving $\Delta$ by converting some other bound through an ad hoc formula (say, dividing a level-based bound by a predictor's range to get a slope-based one) without a citation behind the conversion. That's just a self-defined bound wearing a formula as a disguise, and it has the same problem an invented $\Delta$ always has — nothing stops you from picking a convenient one after you've seen how the data come out, if you didn't pre-register it and can't point to where it came from.

## "Why not just report the CI and let the reader decide?"

I pushed on this myself before landing on TOST as the thing to actually recommend, so it's worth setting out honestly rather than waving away. The estimation-focused view — report the point estimate and CI, skip the dichotomous decision — is a real, well-respected position (Cumming's "New Statistics" is the standard reference). And it's true that TOST computes nothing beyond what the CI already contains; the whole procedure collapses to a single inclusion check.

But "let the reader decide" quietly reconstructs the same problem: deciding a CI is "narrow enough" to call an effect negligible requires exactly the same judgment as choosing $\Delta$ — it's just delegated to each reader's private, unstated threshold instead of the author's public, defended one. And a Registered Report specifically can't defer that judgment to the reader, because the whole mechanism depends on the *authors* committing, before they've seen the data, to what counts as support, null, and inconclusive. "We'll show a CI and let people judge" isn't a commitment; it's a non-answer to the thing the journal is actually asking for.

The resolution I landed on: do both, and they stop being in tension. Report $\Delta$ as the pre-registered, defended criterion — that's the box the journal needs checked. Report the raw coefficient and CI right alongside it — that's what lets a reader who disagrees with your $\Delta$ apply their own, using numbers you've already published, with no re-analysis required. That last part is also, I'd argue, a real point in TOST's favor over a bare Bayes factor: a skeptical reader can route around a disputed $\Delta$ using the reported CI directly; disputing a Bayes factor means recomputing the whole thing under a different prior.

## What I'd actually do

Lead with equivalence testing, report the Bayes factor alongside as a convergent check — cheap, and pure upside when they agree. Pre-register $\Delta$ using whichever of the three justification routes above actually fits the field, and say so explicitly rather than asserting it's obviously correct; every $\Delta$ is contestable, and the point of pre-registering it is that it can be contested on the record, not that it's beyond dispute. And always report the plain coefficient and CI next to the equivalence verdict — not because the formal test is wrong, but because the CI is the part that survives even if a reader rejects your bound.
