---
title: "Equivalence Testing and Two One-Sided Tests (TOST)"
date: "2026-07-24"
---

## The Problem: Non-Significance Is Not Evidence of Absence

In standard Null Hypothesis Significance Testing (NHST), we test:

$$H_0: \theta = 0 \quad \text{vs.} \quad H_1: \theta \neq 0$$

Failing to reject $H_0$ ($p > 0.05$) does not demonstrate that $\theta = 0$. A large $p$-value occurs in two very different situations:
1. The true effect is negligible ($\theta \approx 0$).
2. The effect is meaningful, but the test is underpowered due to small sample size or noisy measurements.

Standard NHST cannot distinguish between these two states. To establish that an effect is absent or negligible, we must invert the testing burden using **equivalence testing**.

## Two One-Sided Tests (TOST)

Equivalence testing defines a Smallest Effect Size of Interest (SESOI), denoted by $[-\Delta, +\Delta]$. Any effect within this bound is considered practically negligible.

The Two One-Sided Tests (TOST) procedure (Schuirmann, 1987; Lakens, 2017) tests whether the parameter falls outside this interval:

$$
H_{01}: \theta \le -\Delta \quad \text{and} \quad H_{02}: \theta \ge +\Delta
$$

If both null hypotheses are rejected at significance level $\alpha$, we conclude that $-\Delta < \theta < +\Delta$ with $(1 - 2\alpha)$ confidence. 

Operationally, TOST corresponds to checking whether the $(1 - 2\alpha)$ confidence interval falls entirely within $[-\Delta, +\Delta]$. For a standard $\alpha = 0.05$ test, this corresponds to the 90% confidence interval.

## Implementation in R: `TOSTER`

Consider 1974 motor trend data (`mtcars`), comparing quarter-mile acceleration time (`qsec`) between automatic ($n = 19$) and manual ($n = 13$) transmissions.

```{r}
#| message: false
library(TOSTER)
data(mtcars)
mtcars$am <- factor(mtcars$am, labels = c("automatic", "manual"))

# Standard two-sample t-test
t.test(qsec ~ am, data = mtcars)
```

The standard two-sample $t$-test yields $p = 0.21$, which is inconclusive.

### TOST with $\Delta = 1$ second

Suppose we pre-specify that a difference of less than 1 second ($\Delta = 1$) is negligible:

```{r}
t_TOST(formula = qsec ~ am, data = mtcars, eqb = 1, var.equal = FALSE)
```

The 90% confidence interval is $[-0.27, 1.91]$, which extends beyond $+1.0$. We fail to reject $H_{02}$; the data are insufficiently powered to claim equivalence within $\pm 1$ second.

### TOST with $\Delta = 2$ seconds

If the substantive tolerance is instead $\Delta = 2$ seconds:

```{r}
t_TOST(formula = qsec ~ am, data = mtcars, eqb = 2, var.equal = FALSE)
```

The 90% confidence interval $[-0.27, 1.91]$ lies completely within $[-2, 2]$. Both one-sided tests reject ($p_1 < 0.001$, $p_2 = 0.039$), confirming statistical equivalence within $\pm 2$ seconds at $\alpha = 0.05$.

## Bayesian Equivalence Testing: The Unit Trap

An alternative approach evaluates the Bayes Factor (BF) for an interval null:

$$\text{BF}_{\text{in/out}} = \frac{p(\text{data} \mid \theta \in [-\Delta, +\Delta])}{p(\text{data} \mid \theta \notin [-\Delta, +\Delta])}$$

A critical implementation trap arises when moving between frequentist and Bayesian software in R:
- `TOSTER::t_TOST(eqb = ...)` takes bounds in **raw outcome units** (seconds).
- `BayesFactor::ttestBF(nullInterval = ...)` expects bounds in **standardized effect size units** (Cohen's $d$).

Passing raw bounds into `ttestBF` will distort the test. The bound must be explicitly converted using the pooled standard deviation:

```{r}
#| message: false
library(BayesFactor)

# Compute pooled SD to standardize the 2-second bound
grp <- split(mtcars$qsec, mtcars$am)
n1 <- length(grp[[1]])
n2 <- length(grp[[2]])
sd_pooled <- sqrt(((n1 - 1) * var(grp[[1]]) + (n2 - 1) * var(grp[[2]])) / (n1 + n2 - 2))

delta_d <- 2 / sd_pooled  # convert +/- 2 seconds to Cohen's d
c(sd_pooled = sd_pooled, delta_d = delta_d)
```

With $s_{\text{pooled}} \approx 1.77$, $\Delta = 2$ corresponds to $d \approx 1.13$. We can now test the interval null:

```{r}
ttestBF(formula = qsec ~ am, data = mtcars, nullInterval = c(-delta_d, delta_d))
```

The Bayes factor for the effect falling within $[-\Delta, +\Delta]$ relative to the point null is 0.98, while the Bayes factor for the effect falling outside is 0.022. The ratio favors the equivalence interval by approximately $45:1$ ($0.98 / 0.022 \approx 44.5$), matching the frequentist TOST conclusion.

## Specifying the Equivalence Bound ($\Delta$)

The equivalence bound $\Delta$ cannot be determined statistically from the sample at hand. Valid strategies include:
1. **Substantive / Practical Significance:** The minimum threshold that translates to real-world impact or economic relevance.
2. **Cost-Benefit Anchoring:** The minimum effect size required to justify the operational costs of an intervention.
3. **Power-Based Benchmark ("Small Telescopes", Simonsohn 2015):** The effect size a prior study had only 33% power to detect ($d_{33\%}$) --- an effect the original design would have missed two times in three. Note that 33% is a power level, not a fraction of the original estimate: $d_{33\%}$ is a little over half of $d_{80\%}$.

## Summary

1. Do not interpret $p > 0.05$ as evidence for the absence of an effect.
2. Formally test equivalence via TOST by evaluating whether the $(1 - 2\alpha)$ confidence interval falls within pre-specified bounds $[-\Delta, +\Delta]$.
3. When using Bayesian interval testing, ensure equivalence bounds match the scale expected by the software (raw vs. Cohen's $d$).
4. Always report the point estimate and $(1 - 2\alpha)$ confidence interval alongside the test decision so readers can assess sensitivity to alternative bounds.
