---
title: "Fixed or Random Effect, or Both?"
date: "2019-05-23"
---



## Panel data

With panel data — repeated observations over time or observations clustered at a higher level — the standard choice is between fixed and random effects. Consider

$$ y_{it} = \beta_0 + \beta_1 x_{it} + c_i + \epsilon_{it} $$ {#eq-fixed-random-effect-1}

for individuals $i = 1, \ldots, n$ measured at $t = 1, \ldots, T$, where $c_i$ is an unobserved time-invariant individual effect. The two approaches differ in how they handle $c_i$.

Fixed effects eliminate $c_i$ by demeaning or by including individual dummies (the two are equivalent in a linear model). In nonlinear models only Poisson avoids the incidental-parameter problem; the standard solution for other models is conditional likelihood (conditional logit, for example), which absorbs the fixed effects in the likelihood. When no conditional likelihood exists, the incidental-parameter bias shrinks with panel depth but does not vanish.

Random effects treat $c_i$ as part of the error term. This is more efficient but requires $x_{it}$ to be uncorrelated with $c_i$ — an assumption economists are generally skeptical of.

## Time-invariant variables

Sometimes we want the effect of a time-invariant variable, so the model becomes

$$ y_{it} = \beta_0  + \beta_1 x_{it} + c_i  + \gamma z_i+ \epsilon_{it} $$ {#eq-fixed-random-effect-2}

Fixed effects cannot identify $\gamma$ because $z_i$ is perfectly collinear with $c_i$. Random effects can still estimate both, treating $z_i$ as another covariate.

## Between-within model

The textbook advice is a Hausman test: random effects are more efficient when the uncorrelatedness assumption holds; if not, fixed effects are still consistent. The between-within model (BW) sidesteps the choice. [Neuhaus and Kalbfleisch (1998)](https://www.ncbi.nlm.nih.gov/pubmed/9629647) introduced the BW estimator,

$$ y_{it} = \beta_0  + \beta_1 (x_{it} - \bar x_i) + \beta_2 \bar x_i + c_i  + \gamma z_i+ \epsilon_{it} $$ {#eq-fixed-random-effect-3}

Here $\beta_1$ is the within effect — identical to the fixed-effect coefficient. $\beta_2$ is the between effect (mean of $x$ on mean of $y$). $\gamma$ is the effect of the time-invariant variable.

The other specification of BW estimator is

$$ y_{it} = \beta_0  + \beta_1 x_{it}  + \beta_2 \bar x_i + c_i  + \gamma z_i+ \epsilon_{it} $$ {#eq-fixed-random-effect-4}

This is the same model reparametrized. $\beta_1$ is unchanged, and $\beta_2$ is now the between-minus-within difference (substitute $\beta_1(x_{it}-\bar x_i) + \beta_2 \bar x_i = \beta_1 x_{it} + (\beta_2-\beta_1)\bar x_i$ to see this). This is the "contextual model"; $\beta_2$ is the contextual effect and acts as an embedded Hausman test.

The BW model has two advantages. It recovers the fixed-effect coefficient while estimating time-invariant covariates. And it extends naturally to cross-level interactions, random slopes, and other multilevel structures. Implementation is straightforward: run a random-effects model on either equation above.

## BW model in R

R's [`panelr`](https://panelr.jacob-long.com/articles/wbm.html) package implements BW models directly.


```{r}
#| echo: true
#| message: false

library(panelr)
data("WageData")
wages <- panel_data(WageData, id = id, wave = t)
model1 <- wbm(lwage ~ wks + union + ms + occ | blk + fem, data = wages)
summary(model1)
```

Compare with `lfe`.

```{r}
#| echo: true
#| message: false

library(lfe)
model2 <- felm(lwage ~ wks + union + ms + occ | id, data = wages)
summary(model2)
```
The within coefficients are identical to machine precision (`WageData` is a balanced panel: 7 waves, 595 individuals). `panelr` additionally estimates the time-invariant covariates `blk` and `fem`. `lfe` supports clustered standard errors, which `panelr` does not.

```{r}
#| echo: true
#| message: false

model3 <- felm(lwage ~ wks + union + ms + occ | id | 0 | id, data = wages)
summary(model3)
```

## BW model in Stata

Stata has no dedicated BW command, but `xtreg` suffices.

```{r}
#| echo: false
#| message: false

library(Statamarkdown)
```

```{stata}
*| cache: true

webuse nlswork
xtset idcode
xtreg ln_w age, fe cluster(idcode)

```

Generate the group mean and run the BW estimation.
```{stata}
*| cache: true

webuse nlswork
xtset idcode
bysort idcode: center age, prefix(d) mean(m)
xtreg ln_w dage mage i.race, re cluster(idcode)
```
In this BW model, we use the centered (within-group-demeaned) `dage` together with the group mean `mage`, which gives the standard within-between (Mundlak) decomposition: the coefficient on `dage`, .0181, is exactly the fixed-effect ("within") coefficient on age, and the coefficient on `mage`, .0226, is directly the between effect. And we have the effect of time-invariant covariate race estimated.  The advantage of using xtreg is that we have clustered standard errors implemented.

Note: if we instead ran `xtreg ln_w age mage i.race, re cluster(idcode)` — using the raw (uncentered) `age` together with `mage` — we would get the *contextual model* instead: the coefficient on `age` stays .0181, but the coefficient on `mage` becomes .0044, the "contextual effect" (the *additional* between-effect on top of the within effect). The two models are algebraically related: contextual-model mage = Mundlak-model mage − Mundlak-model dage (here, .0226 − .0181 ≈ .0044).


## BW model in non-linear models

[Allison](https://statisticalhorizons.com/between-within-contextual-effects) applies the BW model to binary outcomes. The bias relative to conditional logit is an open question, but if the linear probability model is a reasonable approximation the BW approach should perform similarly for binary panel data.



