---
title: "Frengression and Engression"
date: "2026-03-21"
---

In the [causal simulation chapter](causal-simulation.qmd), we saw how Evans & Didelez (2024) decompose the joint distribution $p(x,a,y)$ into three pieces:

$$ p(x,a,y) = p(x,a) \cdot p_a(y) \cdot c(x,y|a) $$

where $p(x,a)$ is the "past" (marginal of covariates and treatment), $p_a(y)$ is the causal margin (marginal structural model), and $c(x,y|a)$ is the copula (dependence between covariates and outcome given treatment).  The `causl` package implements this with parametric families: you specify Gaussian, binomial, etc. for each piece.

But what if the real data has complicated structure that's hard to specify parametrically?  What if we don't know the right copula family, or the marginal distributions are non-standard?

**Frengression** (Yang, Evans and Shen, 2025) solves this by learning all three pieces from data using neural networks.  Same decomposition, but nonparametric.

It has two major functions:  first, causal simulation using nonparametric generative models; second, interventional prediction using learned confounding structure.  The latter is called **Engression**.  First let's talk about Engression.

## Engression

Engression (Shen & Meinshausen, 2024, JRSSB) stands for "energy regression."  Standard regression predicts $E[Y|X]$ — a single number.  Engression learns the entire conditional distribution $P(Y|X)$ using neural networks trained with the **energy score**.

### The Energy Score

The energy score is a proper scoring rule for distributions.  Given observations $Y$ and samples $\hat Y, \hat Y'$ from the model:

$$ S(Y, \hat Y, \hat Y') = E\|Y - \hat Y\| - \frac{1}{2}E\|\hat Y - \hat Y'\| $$

The first term measures how close the model's samples are to the truth.  The second term penalizes lack of spread — without it, the model could collapse to a point prediction.  Minimizing this score trains the network to match the full distribution, not just the mean.

### The Stochastic Network (StoNet)

The building block is the `StoNet` — a neural network that concatenates random noise $\epsilon \sim N(0,I)$ with the input at each layer:

$$\text{StoNet}(x) = f(x, \epsilon), \quad \epsilon \sim N(0,I)$$

Different noise draws produce different outputs — each one is a sample from the learned conditional distribution.  This is what makes engression a generative model.  You can:

- **Predict the mean**: average over many noise draws → $E[Y|X=x]$
- **Predict quantiles**: take quantiles of the noise draws → conditional quantile regression
- **Sample**: each forward pass with fresh noise is a sample from $P(Y|X=x)$

### Why Engression?

What makes engression special compared to other distributional regression methods (quantile regression, mixture density networks)?

1. **Extrapolation**: the energy score encourages the model to extrapolate sensibly outside the training range.  Standard neural networks often produce wild predictions outside training data; engression stays more calibrated.
2. **Multivariate**: it handles multivariate responses naturally — no need to specify marginals separately.
3. **No distributional assumptions**: unlike mixture density networks (which assume Gaussian mixtures) or quantile regression (which estimates quantiles one at a time), engression learns the full joint distribution of $Y|X$ without parametric assumptions.

### Quick Example

```{r}
#| label: engression_example
#| warning: false
#| cache: false
#| message: false
#| eval: false

library(engression)

# Nonlinear DGP with heteroskedastic noise
n <- 1000
X <- matrix(rnorm(n * 2), ncol = 2)
Y <- (X[,1])^2 + X[,2] + rnorm(n) * (0.5 + abs(X[,1]))

# Fit engression — learns the full conditional distribution
engr <- engression(X, Y)

# Predict conditional mean (like lm, but nonlinear)
Yhat_mean <- predict(engr, X, type = "mean")

# Predict conditional quantiles (like quantile regression, but joint)
Yhat_quant <- predict(engr, X, type = "quantiles")

# Sample from the learned distribution (generative)
Ysample <- predict(engr, X, type = "sample", nsample = 1)
```

The key point: engression gives you a *distributional* model of $Y|X$, not just a point prediction. 

## From Engression to Frengression

Frengression = **fr**ugal simulation + **engression**.  It uses three engression models (StoNets) to decompose the joint distribution $P(X,A,Y)$ into the causal simulation framework.  Each sub-network is an engression model trained with the energy score.

## The Architecture

Frengression has three sub-networks, matching the three pieces of the decomposition:

| Decomposition | causl (parametric) | Frengression (neural net) |
|---|---|---|
| $P(X,A)$ — the past | Specified by family + params | `model_xz`: noise-only generator |
| $P_a(Y)$ — causal margin | Specified by family + params | `model_y`: takes $(x, \text{eta})$, outputs $y$ |
| $c(X,Y \mid A)$ — copula | Gaussian/Frank/etc. copula | `model_eta`: takes $(x,z)$, outputs eta |

The key insight is how `model_eta` works.  It takes the covariates $(x,z)$ and outputs a latent variable `eta` that encodes the dependence between confounders and outcome.  During training, a Z-permutation trick forces `eta` to be marginally $N(0,I)$.  This means:

- When `eta` comes from `model_eta(x,z)` (observational): it carries the confounding information, reproducing the observed data
- When `eta` is sampled from $N(0,1)$ independently (interventional): no confounding, giving you $P(Y|do(X))$

Same network, different source of `eta`.  That's the whole trick.

## Using Rfrengression

The `Rfrengression` package is a native R implementation using the `torch` package.

```{r}
#| label: setup
#| warning: false
#| cache: false
#| message: false
library(Rfrengression)
```

### Example: Binary Treatment DGP

Let's use a DGP with binary treatment, binary outcome, and four confounders:

```{r}
#| label: dgp
#| warning: false
#| cache: false
#| message: false
set.seed(42)
n <- 2000

w1 <- rbinom(n, 1, 0.5)
w2 <- rbinom(n, 1, 0.5)
w3 <- round(runif(n, 0, 4), 3)
w4 <- round(runif(n, 0, 5), 3)

A <- rbinom(n, 1, plogis(-0.4 + 0.2*w2 + 0.15*w3 + 0.2*w4 + 0.15*w2*w4))
Y.1 <- rbinom(n, 1, plogis(-1 + 1 - 0.1*w1 + 0.3*w2 + 0.25*w3 + 0.2*w4 + 0.15*w2*w4))
Y.0 <- rbinom(n, 1, plogis(-1 + 0 - 0.1*w1 + 0.3*w2 + 0.25*w3 + 0.2*w4 + 0.15*w2*w4))
Y <- Y.1 * A + Y.0 * (1 - A)

true_ate <- mean(Y.1 - Y.0)
cat("True ATE:", round(true_ate, 4), "\n")
```

### Train Frengression

Two-stage training: first the outcome model (`model_y` + `model_eta`), then the marginal (`model_xz`).

```{r}
#| label: train
#| warning: false
#| cache: false
#| message: false
x_mat <- matrix(A, ncol = 1)
z_mat <- as.matrix(data.frame(w1, w2, w3, w4))
y_mat <- matrix(Y, ncol = 1)

model <- frengression(x_dim = 1, y_dim = 1, z_dim = 4,
                      x_binary = TRUE, y_binary = TRUE,
                      z_binary_dims = 2, noise_dim = 10)

model <- train_y(model, x_mat, z_mat, y_mat,
                 num_iters = 1000, lr = 1e-3, print_every = 500)
model <- train_xz(model, x_mat, z_mat,
                   num_iters = 1000, lr = 1e-4, print_every = 500)
```

### Estimate ATE

The interventional prediction samples `eta` from $N(0,1)$ — no confounding:

```{r}
#| label: ate
#| warning: false
#| cache: false
#| message: false
y1 <- predict(model, matrix(1, ncol = 1), type = "mean", nsample = 2000)
y0 <- predict(model, matrix(0, ncol = 1), type = "mean", nsample = 2000)

cat("Frengression ATE:", round(y1 - y0, 4), "\n")
cat("True ATE:", round(true_ate, 4), "\n")
```

## Frengression as DGP

Frengression can estimate ATE, but that's not its main strength — dedicated semiparametric estimators (npcausal, TMLE, DoubleML) are designed for that.  The real value is using frengression as a **data-generating process** for benchmarking.

The idea:

1. Train frengression on observational data — it learns the realistic confounding structure
2. Replace the causal margin with a known function via `specify_causal()`
3. Generate synthetic datasets — realistic confounding, known ground truth

```{r}
#| label: dgp_bench
#| warning: false
#| cache: false
#| message: false
# Specify a known causal effect: Y = sigmoid(0.2*X + eta)
model_bench <- specify_causal(model, function(x, eta) {
  torch::torch_sigmoid(0.2 * x + eta)
})

# True ATE computed empirically from the model
y1_true <- predict(model_bench, matrix(1, ncol = 1), type = "mean", nsample = 5000)
y0_true <- predict(model_bench, matrix(0, ncol = 1), type = "mean", nsample = 5000)
cat("True ATE (specified):", round(as.numeric(y1_true - y0_true), 4), "\n")
```

Now we can generate synthetic datasets and benchmark any method:

```{r}
#| label: syn
#| warning: false
#| cache: false
#| message: false
syn <- sample_joint(model_bench, sample_size = 1000)
cat("Synthetic data: n =", nrow(syn$x), "\n")
cat("Treatment prevalence:", mean(syn$x), "\n")
cat("Outcome prevalence:", mean(syn$y), "\n")
```

## Real Data: LaLonde

Where this really shines is with real data.  The LaLonde dataset (job training program, 614 observations) has complex confounding that no one could specify parametrically — many zeros in prior earnings, non-standard covariate distributions, poor overlap.

```{r}
#| label: lalonde
#| warning: false
#| cache: false
#| message: false
library(MatchIt)
data("lalonde", package = "MatchIt")
lalonde$black <- as.integer(lalonde$race == "black")
lalonde$hispan <- as.integer(lalonde$race == "hispan")

x_lal <- matrix(lalonde$treat, ncol = 1)
y_lal <- matrix(lalonde$re78, ncol = 1)
z_lal <- as.matrix(lalonde[, c("age", "educ", "black", "hispan",
                                "married", "nodegree", "re74", "re75")])

# Standardize for training
z_sc <- scale(z_lal)
y_sc <- (y_lal - mean(y_lal)) / sd(y_lal)

model_lal <- frengression(x_dim = 1, y_dim = 1, z_dim = 8,
                           x_binary = TRUE, y_binary = FALSE,
                           z_binary_dims = 4, noise_dim = 10)

model_lal <- train_y(model_lal, x_lal, z_sc, y_sc,
                      num_iters = 1000, lr = 1e-3, print_every = 500)
model_lal <- train_xz(model_lal, x_lal, z_sc,
                        num_iters = 1000, lr = 1e-4, print_every = 500)
```

Now specify a known effect and generate:

```{r}
#| label: lalonde_gen
#| warning: false
#| cache: false
#| message: false
model_lal_bench <- specify_causal(model_lal, function(x, eta) {
  0.3 * x + eta
})

# Generate synthetic LaLonde-like data with known ATE = 0.3 (standardized)
syn_lal <- sample_joint(model_lal_bench, sample_size = 2000)
cat("Synthetic LaLonde: n =", nrow(syn_lal$x), "\n")
cat("Treatment prevalence:", round(mean(syn_lal$x), 3), "\n")
cat("True ATE:", 0.3, "(standardized) =", round(0.3 * sd(y_lal), 0), "dollars\n")
```

The synthetic data inherits LaLonde's messy confounding structure — the complex propensity score, the zero-inflated prior earnings, the demographic imbalances — but we know the true causal effect.  No researcher could hand-specify this DGP.

## Comparison: causl vs frengression

Both packages implement the same decomposition:

| Feature | causl | Frengression |
|---|---|---|
| Marginal $P(X,A)$ | Parametric families | Learned by neural net |
| Causal margin $P_a(Y)$ | Parametric families | Learned (or specified) |
| Copula | Gaussian, Frank, etc. | Learned by `model_eta` |
| Requires specification | Yes — families + params | No — learns from data |
| Can learn from real data | Limited (plasmode) | Yes — full joint |
| Distributional flexibility | Limited by family choice | Arbitrary |

The `causl` approach is more transparent — you know exactly what you specified.  Frengression is more flexible — it can capture structure you didn't anticipate.  They're complementary.

## Benchmarking: Frengression vs Semiparametric Methods

Now let's compare frengression against standard semiparametric causal inference tools on the same data.  The three comparison methods are:

- **npcausal**: AIPW estimator with SuperLearner nuisance models
- **TMLE**: Targeted minimum loss-based estimation with SuperLearner
- **DoubleML**: Partially linear regression with ranger (random forest)

These methods are purpose-built for ATE estimation using influence function theory — they should be more precise for this specific estimand.  The question is how frengression compares, and what it can do that they cannot.

```{r}
#| label: bench_libs
#| warning: false
#| cache: true
#| message: false
library(npcausal)
library(tmle)
library(SuperLearner)
library(DoubleML)
library(mlr3)
library(mlr3learners)
library(tidyverse)
library(knitr)
```

### Single-Dataset Comparison

First, run all four methods on the binary DGP data created above.

```{r}
#| label: npcausal_ate
#| warning: false
#| cache: true
#| message: false
W <- data.frame(w1, w2, w3, w4)
SL.library <- c("SL.earth", "SL.glm.interaction", "SL.mean",
                "SL.ranger", "SL.glmnet")

set.seed(123)
aipw_fit <- ate(y = Y, a = A, x = W, nsplits = 2, sl.lib = SL.library)

ate_npcausal <- aipw_fit$res$est[3]
se_npcausal  <- aipw_fit$res$se[3]
ci_npcausal  <- c(aipw_fit$res$ci.ll[3], aipw_fit$res$ci.ul[3])

cat("npcausal ATE:", round(ate_npcausal, 4), "\n")
cat("SE:", round(se_npcausal, 4), "\n")
```

```{r}
#| label: tmle_ate
#| warning: false
#| cache: true
#| message: false
set.seed(123)
tmle_fit <- tmle(Y = Y, A = A, W = W, family = "binomial",
                 Q.SL.library = SL.library, g.SL.library = SL.library)

ate_tmle <- tmle_fit$estimates$ATE$psi
se_tmle  <- sqrt(tmle_fit$estimates$ATE$var.psi)
ci_tmle  <- tmle_fit$estimates$ATE$CI

cat("TMLE ATE:", round(ate_tmle, 4), "\n")
cat("SE:", round(se_tmle, 4), "\n")
```

```{r}
#| label: doubleml_ate
#| warning: false
#| cache: true
#| message: false
lgr::get_logger("mlr3")$set_threshold("warn")
data_obs <- data.frame(w1, w2, w3, w4, A, Y)

dml_data <- DoubleMLData$new(data_obs,
                              y_col = "Y", d_cols = "A",
                              x_cols = c("w1", "w2", "w3", "w4"))

learner_l <- lrn("regr.ranger", num.trees = 500, max.depth = 5, min.node.size = 2)
learner_m <- learner_l$clone()

set.seed(123)
dml_plr <- DoubleMLPLR$new(dml_data, ml_l = learner_l, ml_m = learner_m)
dml_plr$fit()

ate_dml <- dml_plr$coef
se_dml  <- dml_plr$se
ci_dml  <- c(dml_plr$confint()[1], dml_plr$confint()[2])

cat("DoubleML ATE:", round(ate_dml, 4), "\n")
cat("SE:", round(se_dml, 4), "\n")
```

```{r}
#| label: frengression_se
#| warning: false
#| cache: false
#| message: false
# Frengression ATE already computed above (y1 - y0)
ate_freng <- as.numeric(y1 - y0)

# Bootstrap SE via repeated prediction (Monte Carlo variability)
set.seed(123)
ate_boot <- replicate(200, {
  predict(model, matrix(1, ncol = 1), type = "mean", nsample = 500) -
    predict(model, matrix(0, ncol = 1), type = "mean", nsample = 500)
})
se_freng <- sd(ate_boot)
```

```{r}
#| label: comparison_table
#| warning: false
#| cache: true
#| message: false
results <- data.frame(
  Method = c("True ATE", "Frengression", "npcausal (AIPW)", "TMLE", "DoubleML (PLR)"),
  ATE = c(true_ate, ate_freng, ate_npcausal, ate_tmle, ate_dml),
  SE = c(NA, se_freng, se_npcausal, se_tmle, se_dml),
  CI_lower = c(NA,
               ate_freng - 1.96 * se_freng,
               ci_npcausal[1], ci_tmle[1], ci_dml[1]),
  CI_upper = c(NA,
               ate_freng + 1.96 * se_freng,
               ci_npcausal[2], ci_tmle[2], ci_dml[2])
)

results$Bias <- results$ATE - true_ate
results$Covers <- ifelse(is.na(results$CI_lower), NA,
                         results$CI_lower <= true_ate & true_ate <= results$CI_upper)

kable(results, digits = 4,
      caption = "ATE Estimation: Frengression vs Semiparametric Methods")
```

The semiparametric estimators are purpose-built for ATE and leverage influence function theory to achieve $\sqrt{n}$-consistent, doubly-robust estimates.  Frengression learns the *entire joint distribution* $P(X,Y,Z)$, from which ATE is just one functional — so some loss of precision for this specific estimand is expected.

### Monte Carlo Benchmarking (Simulated DGP)

This is frengression's "killer app."  We use the model trained above, plug in a known causal effect via `specify_causal()`, generate B=10 synthetic datasets, and benchmark all four estimators.

```{r}
#| label: benchmark_sim
#| warning: false
#| cache: false
#| message: false
B <- 10
n_syn <- 2000

methods <- c("Frengression", "npcausal", "TMLE", "DoubleML")
ate_mat <- matrix(NA, nrow = B, ncol = length(methods),
                  dimnames = list(NULL, methods))
se_mat  <- ate_mat
cover_mat <- ate_mat

SL.lib_fast <- c("SL.glm.interaction", "SL.ranger", "SL.mean")

# True ATE from the specified model (already computed above as y1_true - y0_true)
true_ate_bench <- as.numeric(y1_true - y0_true)

set.seed(2024)
for (b in seq_len(B)) {
  syn <- sample_joint(model_bench, sample_size = n_syn)

  x_syn <- as.numeric(syn$x)
  y_syn <- as.numeric(syn$y)
  z_syn <- as.data.frame(syn$z)
  colnames(z_syn) <- paste0("z", 1:ncol(z_syn))
  df_syn <- cbind(data.frame(A = x_syn, Y = y_syn), z_syn)

  # --- Frengression ---
  tryCatch({
    mod_f <- frengression(x_dim = 1, y_dim = 1, z_dim = ncol(z_syn),
                          x_binary = TRUE, y_binary = TRUE,
                          noise_dim = 10, hidden_dim = 100, num_layer = 3)
    mod_f <- train_y(mod_f, matrix(x_syn, ncol=1), as.matrix(z_syn),
                     matrix(y_syn, ncol=1),
                     num_iters = 500, lr = 1e-3, silent = TRUE)
    mod_f <- train_xz(mod_f, matrix(x_syn, ncol=1), as.matrix(z_syn),
                       num_iters = 500, lr = 1e-4, silent = TRUE)

    y1_f <- predict(mod_f, matrix(1, ncol=1), type="mean", nsample=1000)
    y0_f <- predict(mod_f, matrix(0, ncol=1), type="mean", nsample=1000)
    ate_mat[b, "Frengression"] <- y1_f - y0_f

    ate_mc <- replicate(100, {
      predict(mod_f, matrix(1,ncol=1), type="mean", nsample=200) -
        predict(mod_f, matrix(0,ncol=1), type="mean", nsample=200)
    })
    se_mat[b, "Frengression"] <- sd(ate_mc)
    ci_f <- ate_mat[b, "Frengression"] + c(-1,1) * 1.96 * se_mat[b, "Frengression"]
    cover_mat[b, "Frengression"] <- (ci_f[1] <= true_ate_bench & true_ate_bench <= ci_f[2])
  }, error = function(e) message("Frengression failed on rep ", b, ": ", e$message))

  # --- npcausal ---
  tryCatch({
    fit_np <- ate(y = y_syn, a = x_syn, x = z_syn,
                  nsplits = 2, sl.lib = SL.lib_fast)
    ate_mat[b, "npcausal"] <- fit_np$res$est[3]
    se_mat[b, "npcausal"]  <- fit_np$res$se[3]
    ci <- c(fit_np$res$ci.ll[3], fit_np$res$ci.ul[3])
    cover_mat[b, "npcausal"] <- (ci[1] <= true_ate_bench & true_ate_bench <= ci[2])
  }, error = function(e) message("npcausal failed on rep ", b, ": ", e$message))

  # --- TMLE ---
  tryCatch({
    fit_tmle <- tmle(Y = y_syn, A = x_syn, W = z_syn, family = "binomial",
                     Q.SL.library = SL.lib_fast, g.SL.library = SL.lib_fast)
    ate_mat[b, "TMLE"] <- fit_tmle$estimates$ATE$psi
    se_mat[b, "TMLE"]  <- sqrt(fit_tmle$estimates$ATE$var.psi)
    ci_t <- fit_tmle$estimates$ATE$CI
    cover_mat[b, "TMLE"] <- (ci_t[1] <= true_ate_bench & true_ate_bench <= ci_t[2])
  }, error = function(e) message("TMLE failed on rep ", b, ": ", e$message))

  # --- DoubleML ---
  tryCatch({
    dml_d <- DoubleMLData$new(df_syn, y_col = "Y", d_cols = "A",
                               x_cols = colnames(z_syn))
    lr_l <- lrn("regr.ranger", num.trees = 300, max.depth = 5, min.node.size = 2)
    lr_m <- lr_l$clone()
    dml_obj <- DoubleMLPLR$new(dml_d, ml_l = lr_l, ml_m = lr_m)
    dml_obj$fit()
    ate_mat[b, "DoubleML"] <- dml_obj$coef
    se_mat[b, "DoubleML"]  <- dml_obj$se
    ci_d <- c(dml_obj$confint()[1], dml_obj$confint()[2])
    cover_mat[b, "DoubleML"] <- (ci_d[1] <= true_ate_bench & true_ate_bench <= ci_d[2])
  }, error = function(e) message("DoubleML failed on rep ", b, ": ", e$message))

  if (b %% 5 == 0) cat("Completed replication", b, "of", B, "\n")
}
```

```{r}
#| label: benchmark_results
#| warning: false
#| cache: true
#| message: false
bench_summary <- data.frame(
  Method = colnames(ate_mat),
  Mean_ATE = colMeans(ate_mat, na.rm = TRUE),
  Bias = colMeans(ate_mat, na.rm = TRUE) - true_ate_bench,
  SD = apply(ate_mat, 2, sd, na.rm = TRUE),
  RMSE = sqrt(colMeans((ate_mat - true_ate_bench)^2, na.rm = TRUE)),
  Mean_SE = colMeans(se_mat, na.rm = TRUE),
  Coverage_95 = colMeans(cover_mat, na.rm = TRUE),
  N_valid = colSums(!is.na(ate_mat))
)

kable(bench_summary, digits = 4,
      caption = paste0("Monte Carlo Benchmarking (B=", B,
                       ", n=", n_syn, ", True ATE=",
                       round(true_ate_bench, 4), ")"))
```

```{r}
#| label: benchmark_plot
#| warning: false
#| cache: true
#| message: false
#| fig-cap: "ATE Estimates Across Synthetic Datasets"
ate_long <- as.data.frame(ate_mat) |>
  pivot_longer(everything(), names_to = "Method", values_to = "ATE") |>
  filter(!is.na(ATE))

# Trim extreme outliers for readable plot
q_bounds <- quantile(ate_long$ATE, c(0.02, 0.98), na.rm = TRUE)
ate_long_trim <- ate_long |> filter(ATE >= q_bounds[1] & ATE <= q_bounds[2])

ggplot(ate_long_trim, aes(x = Method, y = ATE, fill = Method)) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = true_ate_bench, linetype = "dashed", color = "red",
             linewidth = 1) +
  annotate("text", x = 0.5, y = true_ate_bench, label = "True ATE",
           hjust = 0, vjust = -0.5, color = "red") +
  labs(title = "ATE Estimates Across Frengression-Generated Datasets",
       subtitle = paste0("B=", B, " replications, n=", n_syn, " per dataset"),
       y = "Estimated ATE") +
  theme_minimal() +
  theme(legend.position = "none")
```

### Monte Carlo Benchmarking (LaLonde)

Now the same exercise, but trained on real LaLonde data — where the confounding structure is unknown and complex.  We already trained `model_lal` above and specified a known effect with `model_lal_bench`.

```{r}
#| label: lalonde_bench
#| warning: false
#| cache: false
#| message: false
# Compute true ATE empirically from the specified model
y1_lal_true <- predict(model_lal_bench, matrix(1, ncol=1), type="mean", nsample=5000)
y0_lal_true <- predict(model_lal_bench, matrix(0, ncol=1), type="mean", nsample=5000)
true_ate_lal <- as.numeric(y1_lal_true - y0_lal_true)
cat("True ATE (standardized):", round(true_ate_lal, 4), "\n")
cat("True ATE (dollars):", round(true_ate_lal * sd(y_lal), 0), "\n")

B_lal <- 10
n_syn_lal <- 2000

methods <- c("Frengression", "npcausal", "TMLE", "DoubleML")
ate_mat_lal <- matrix(NA, nrow = B_lal, ncol = length(methods),
                       dimnames = list(NULL, methods))
se_mat_lal  <- ate_mat_lal
cover_mat_lal <- ate_mat_lal

SL.lib_fast <- c("SL.glm.interaction", "SL.ranger", "SL.mean")

set.seed(2024)
for (b in seq_len(B_lal)) {
  syn <- sample_joint(model_lal_bench, sample_size = n_syn_lal)

  x_syn <- as.numeric(syn$x)
  y_syn <- as.numeric(syn$y)
  z_syn <- as.data.frame(syn$z)
  colnames(z_syn) <- paste0("z", 1:ncol(z_syn))
  df_syn <- cbind(data.frame(A = x_syn, Y = y_syn), z_syn)

  # --- Frengression ---
  tryCatch({
    mod_f <- frengression(x_dim = 1, y_dim = 1, z_dim = ncol(z_syn),
                          x_binary = TRUE, y_binary = FALSE,
                          noise_dim = 10, hidden_dim = 100, num_layer = 3)
    mod_f <- train_y(mod_f, matrix(x_syn, ncol=1), as.matrix(z_syn),
                     matrix(y_syn, ncol=1),
                     num_iters = 500, lr = 1e-3, silent = TRUE)
    mod_f <- train_xz(mod_f, matrix(x_syn, ncol=1), as.matrix(z_syn),
                       num_iters = 500, lr = 1e-4, silent = TRUE)

    y1_f <- predict(mod_f, matrix(1, ncol=1), type="mean", nsample=1000)
    y0_f <- predict(mod_f, matrix(0, ncol=1), type="mean", nsample=1000)
    ate_mat_lal[b, "Frengression"] <- y1_f - y0_f

    ate_mc <- replicate(100, {
      predict(mod_f, matrix(1,ncol=1), type="mean", nsample=200) -
        predict(mod_f, matrix(0,ncol=1), type="mean", nsample=200)
    })
    se_mat_lal[b, "Frengression"] <- sd(ate_mc)
    ci_f <- ate_mat_lal[b, "Frengression"] + c(-1,1) * 1.96 * se_mat_lal[b, "Frengression"]
    cover_mat_lal[b, "Frengression"] <- (ci_f[1] <= true_ate_lal & true_ate_lal <= ci_f[2])
  }, error = function(e) message("Frengression failed on rep ", b, ": ", e$message))

  # --- npcausal ---
  tryCatch({
    fit_np <- ate(y = y_syn, a = x_syn, x = z_syn,
                  nsplits = 2, sl.lib = SL.lib_fast)
    ate_mat_lal[b, "npcausal"] <- fit_np$res$est[3]
    se_mat_lal[b, "npcausal"]  <- fit_np$res$se[3]
    ci <- c(fit_np$res$ci.ll[3], fit_np$res$ci.ul[3])
    cover_mat_lal[b, "npcausal"] <- (ci[1] <= true_ate_lal & true_ate_lal <= ci[2])
  }, error = function(e) message("npcausal failed on rep ", b, ": ", e$message))

  # --- TMLE ---
  tryCatch({
    fit_tmle <- tmle(Y = y_syn, A = x_syn, W = z_syn, family = "gaussian",
                     Q.SL.library = SL.lib_fast, g.SL.library = SL.lib_fast)
    ate_mat_lal[b, "TMLE"] <- fit_tmle$estimates$ATE$psi
    se_mat_lal[b, "TMLE"]  <- sqrt(fit_tmle$estimates$ATE$var.psi)
    ci_t <- fit_tmle$estimates$ATE$CI
    cover_mat_lal[b, "TMLE"] <- (ci_t[1] <= true_ate_lal & true_ate_lal <= ci_t[2])
  }, error = function(e) message("TMLE failed on rep ", b, ": ", e$message))

  # --- DoubleML ---
  tryCatch({
    dml_d <- DoubleMLData$new(df_syn, y_col = "Y", d_cols = "A",
                               x_cols = colnames(z_syn))
    lr_l <- lrn("regr.ranger", num.trees = 300, max.depth = 5, min.node.size = 2)
    lr_m <- lr_l$clone()
    dml_obj <- DoubleMLPLR$new(dml_d, ml_l = lr_l, ml_m = lr_m)
    dml_obj$fit()
    ate_mat_lal[b, "DoubleML"] <- dml_obj$coef
    se_mat_lal[b, "DoubleML"]  <- dml_obj$se
    ci_d <- c(dml_obj$confint()[1], dml_obj$confint()[2])
    cover_mat_lal[b, "DoubleML"] <- (ci_d[1] <= true_ate_lal & true_ate_lal <= ci_d[2])
  }, error = function(e) message("DoubleML failed on rep ", b, ": ", e$message))

  if (b %% 5 == 0) cat("Completed replication", b, "of", B_lal, "\n")
}
```

```{r}
#| label: lalonde_results
#| warning: false
#| cache: true
#| message: false
bench_lal <- data.frame(
  Method = colnames(ate_mat_lal),
  Mean_ATE = colMeans(ate_mat_lal, na.rm = TRUE),
  Bias = colMeans(ate_mat_lal, na.rm = TRUE) - true_ate_lal,
  SD = apply(ate_mat_lal, 2, sd, na.rm = TRUE),
  RMSE = sqrt(colMeans((ate_mat_lal - true_ate_lal)^2, na.rm = TRUE)),
  Mean_SE = colMeans(se_mat_lal, na.rm = TRUE),
  Coverage_95 = colMeans(cover_mat_lal, na.rm = TRUE),
  N_valid = colSums(!is.na(ate_mat_lal))
)

kable(bench_lal, digits = 4,
      caption = paste0("LaLonde-Based Benchmarking (B=", B_lal,
                       ", n=", n_syn_lal,
                       ", True ATE=", round(true_ate_lal, 4),
                       " [", round(true_ate_lal * sd(y_lal), 0), " dollars])"))
```

```{r}
#| label: lalonde_plot
#| warning: false
#| cache: true
#| message: false
#| fig-cap: "ATE Estimates from LaLonde-Based Synthetic Data"
ate_long_lal <- as.data.frame(ate_mat_lal) |>
  pivot_longer(everything(), names_to = "Method", values_to = "ATE") |>
  filter(!is.na(ATE))

q_bounds_lal <- quantile(ate_long_lal$ATE, c(0.02, 0.98), na.rm = TRUE)
ate_long_lal_trim <- ate_long_lal |>
  filter(ATE >= q_bounds_lal[1] & ATE <= q_bounds_lal[2])

ggplot(ate_long_lal_trim, aes(x = Method, y = ATE, fill = Method)) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = true_ate_lal, linetype = "dashed", color = "red",
             linewidth = 1) +
  annotate("text", x = 0.5, y = true_ate_lal, label = "True ATE",
           hjust = 0, vjust = -0.5, color = "red") +
  labs(title = "ATE Estimates: LaLonde-Based Frengression DGP",
       subtitle = paste0("B=", B_lal, " replications, n=", n_syn_lal,
                         " per dataset (confounding learned from real data)"),
       y = "Estimated ATE (standardized)") +
  theme_minimal() +
  theme(legend.position = "none")
```

The semiparametric methods show meaningful bias on the LaLonde-generated data, suggesting that neural-network-generated confounding is harder to adjust for than hand-specified simulations.  The synthetic data inherits LaLonde's messy structure — complex propensity score surfaces, zero-inflated prior earnings, demographic imbalances — which violates the smoothness assumptions that SuperLearner and ranger rely on.

## Summary

- Frengression implements the same causal simulation framework as `causl` (Evans & Didelez, 2024), but learns the decomposition nonparametrically
- Three networks: `model_xz` (past), `model_y` (causal margin), `model_eta` (copula/dependence)
- The `specify_causal()` function lets you plug in known effects while keeping learned confounding
- Main use case: realistic Monte Carlo benchmarking with ground truth
- R package: `Rfrengression` (native R torch, no Python dependency)
- Paper: Yang, Evans, and Shen (2025), "Frugal, Flexible, Faithful: Causal Data Simulation via Frengression" (arXiv:2508.01018)
