---
title: "Gender Wage Gap Analysis: Oaxaca-Blinder and Causal Estimation"
author: "Xiang Ao"
date: "2025-11-17"
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(broom)
library(marginaleffects)
library(npcausal)
```

## Introduction: Oaxaca-Blinder as Regression Adjustment

The classical Oaxaca-Blinder (OB) decomposition separates the raw difference in mean outcomes between groups into an "explained" component (differences in observable covariates) and an "unexplained" component (differences in regression coefficients, often interpreted as discrimination):

$$E[Y \mid A = 1] - E[Y \mid A = 0] = \underbrace{(E[X \mid A = 1] - E[X \mid A = 0]) \beta_0}_{\text{Explained (Covariate Differences)}} + \underbrace{E[X \mid A = 1] (\beta_1 - \beta_0)}_{\text{Unexplained (Coefficient Differences)}}$$

where $A = 1$ denotes female and $A = 0$ denotes male. A well-known limitation of the traditional OB decomposition is its sensitivity to the choice of reference group ($\beta_0$ vs. $\beta_1$).

Słoczyński (2013, 2015) connects the OB decomposition directly to the potential outcomes framework:
- Demeaning covariates at the **overall sample mean** identifies the Population Average Gender Effect ($\tau_{\text{PAGE}}$, equivalent to the Average Treatment Effect, ATE).
- Demeaning covariates at the **female subpopulation mean** identifies the Average Treatment Effect on the Treated ($\tau_{\text{PATT}}$, or ATT).

In this framing, the unexplained gap is simply the treatment effect evaluated at the chosen covariate distribution, while the explained gap represents selection bias. This formulation allows us to estimate the gap via standard regression adjustment or modern doubly robust estimators.

## Empirical Application: 2015 CPS Data

We analyze a subsample from the 2015 March Supplement of the U.S. Current Population Survey (`wage2015_subsample_inference.csv`), restricted to white, non-Hispanic full-time workers aged 25–64 ($N = 5,150$). The outcome $Y$ is the logarithm of hourly wage (`lwage`), and the indicator is `female` ($A \in \{0, 1\}$).

```{r}
#| label: ra1
#| cache: true
#| message: false
#| warning: false

data <- read_csv("wage2015_subsample_inference.csv") |>
  rename(socl = scl, sohs = shs, sout = so) |>
  mutate(
    exp_dm   = exp1 - mean(exp1, na.rm = TRUE),
    female   = sex,
    occ2     = factor(occ2),
    ind2     = factor(ind2),
    sohs_dm  = sohs - mean(sohs, na.rm = TRUE),
    hsg_dm   = hsg - mean(hsg, na.rm = TRUE),
    socl_dm  = socl - mean(socl, na.rm = TRUE),
    clg_dm   = clg - mean(clg, na.rm = TRUE),
    mw_dm    = mw - mean(mw, na.rm = TRUE),
    sout_dm  = sout - mean(sout, na.rm = TRUE),
    we_dm    = we - mean(we, na.rm = TRUE)
  )

# 1. Unadjusted (raw) difference
reg1 <- lm(lwage ~ female, data = data)
tidy(reg1) |> filter(term == "female")

# 2. Regression adjustment with demeaned human capital & region controls
reg2 <- lm(lwage ~ female * (exp_dm + sohs_dm + hsg_dm + socl_dm + clg_dm + mw_dm + sout_dm + we_dm), 
           data = data)
tidy(reg2) |> filter(term == "female")
```

### The Composition Effect

The raw wage gap is $-3.8\%$ ($p = 0.012$). After controlling for potential experience, education, and region, the estimated gap widens to $-11.3\%$ ($p < 0.001$). 

Adding controls usually attenuates an observed gap; here it nearly triples it. This occurs because of sample composition:

```{r}
#| label: composition
#| cache: true

data |>
  group_by(female) |>
  summarise(
    n           = n(),
    mean_lwage  = mean(lwage),
    exp1        = mean(exp1),
    `<HS`       = mean(sohs),
    HS          = mean(hsg),
    `Some Coll` = mean(socl),
    College     = mean(clg),
    Advanced    = mean(ad)
  ) |>
  mutate(across(where(is.numeric), ~round(.x, 3))) |>
  knitr::kable(caption = "Covariate means by gender.")
```

Women in this sample have higher average educational attainment than men (17.5% hold advanced degrees vs. 10.7%; only 1.3% have less than high school vs. 3.2%), with identical experience ($\approx 13.8$ years). Because education raises wages, women's educational advantage partially masks the earnings gap in the raw comparison. Controlling for education holds this advantage constant, revealing an adjusted gap of $-11.3\%$ within equivalent education-experience cells.

### Regression Adjustment via `marginaleffects`

Instead of manual demeaning, the same estimands can be obtained directly using `marginaleffects::avg_comparisons()`:

```{r}
#| label: ra2
#| cache: true

reg3 <- lm(lwage ~ female * (exp1 + sohs + hsg + socl + clg + mw + sout + we), data = data)

# Population Average Gender Effect (ATE / PAGE)
avg_comparisons(reg3, variables = "female")

# Female Subpopulation Gender Effect (ATT / PATT)
avg_comparisons(reg3, variables = "female", newdata = subset(data, female == 1))
```

Both ATE and ATT land at $-11.3\%$, confirming that the fully interacted OLS specification evaluated at the respective empirical covariate distributions yields the OB decomposition directly.

## Adding Occupation and Industry: Mediators vs. Bad Controls

A central question in pay gap analyses is whether to adjust for occupation and industry:

```{r}
#| label: ra4
#| cache: true

reg4 <- lm(lwage ~ female * (exp1 + sohs + hsg + socl + clg + mw + sout + we + occ2 + ind2), 
           data = data)

avg_comparisons(reg4, variables = "female", newdata = subset(data, female == 1))
```

Adjusting for 22 occupation and industry categories narrows the gap from $-11.3\%$ to $-7.5\%$. 

### Interpretation Trade-Off

1. **Descriptive Mediation:** If the goal is accounting for occupational sorting, the move from $-11.3\%$ to $-7.5\%$ shows that about one-third of the adjusted wage gap reflects differences in occupation and industry.
2. **Causal Interpretation (Bad Control):** Occupation and industry are post-treatment choices ($A \to \text{Occupation} \to Y$). Conditioning on post-treatment variables can introduce collider bias if unobserved factors (e.g., career preferences, flexibility constraints) simultaneously affect occupational selection and wages ($A \to \text{Occupation} \leftarrow U \to Y$).
3. **Controlled Direct Effect:** The $-7.5\%$ estimate represents a controlled direct effect—the within-occupation wage disparity between men and women with similar observable characteristics.

## Nonparametric Doubly Robust Estimation (AIPW)

Linear regression adjustment assumes that the relationship between covariates and wages is additive and correctly specified. To relax this functional form assumption, we use Augmented Inverse Probability Weighting (AIPW) with SuperLearner via `npcausal` (Kennedy, 2022), incorporating generalized additive models (`earth`), regularized regression (`glmnet`), random forests (`ranger`), and gradient boosting (`xgboost`).

```{r}
#| label: npcausal1
#| warning: false
#| message: false

library(npcausal)

# To avoid prolonged cross-fitting during interactive compilation, we load the
# pre-computed SuperLearner results (SL.earth, SL.glmnet, SL.mean, SL.ranger, SL.xgboost)
# from the book cache if present; otherwise, compute from scratch.
cache_base <- "gwg_cache/html/npcausal1_5538101f2b9ff4f8b21593068dc74ef6"
if (file.exists(paste0(cache_base, ".rdx"))) {
  cache_env <- new.env()
  lazyLoad(cache_base, envir = cache_env)
  aipw_ate <- cache_env$aipw_ate
  aipw_att <- cache_env$aipw_att
} else {
  SL.library <- c("SL.earth", "SL.glmnet", "SL.mean", "SL.ranger", "SL.xgboost")
  reg3 <- lm(lwage ~ exp_dm + sohs + hsg + socl + clg + mw + sout + we + occ2 + ind2, data = data)
  W <- as.data.frame(model.matrix(reg3)[, -1])
  Y <- data$lwage
  A <- data$female
  set.seed(123)
  aipw_ate <- ate(y = Y, a = A, x = W, nsplits = 5, sl.lib = SL.library)
  aipw_att <- att(y = Y, a = A, x = W, nsplits = 5, sl.lib = SL.library)
}
```

```{r}
#| label: npcausal-results
#| warning: false
#| message: false

aipw_ate$res
aipw_att$res
```

### Takeaways from Nonparametric Estimation

1. **Robustness to Functional Form:** The nonparametric AIPW ATT is $-6.7\%$ ($95\%\text{ CI}: [-9.5\%, -3.9\%]$), very close to the parametric interacted OLS estimate of $-7.5\%$. This indicates that linear interaction models provide an adequate approximation.
2. **Substantive Consistency:** Both parametric and machine-learning estimators confirm a statistically significant within-occupation wage gap of approximately $7\%$.

## Summary

1. **Oaxaca-Blinder Connection:** The classical OB decomposition is equivalent to regression adjustment where treatment is interacted with centered covariates. Demeaning at the overall mean yields the ATE, while demeaning at the female subpopulation mean yields the ATT.
2. **Composition Matters:** Conditioning on education widens the wage gap from $-3.8\%$ to $-11.3\%$ because women in the sample possess higher average educational credentials than men.
3. **Occupation as a Mediator:** Conditioning on occupation and industry narrows the gap to $-7.5\%$, capturing the within-job pay disparity (controlled direct effect).
4. **Machine Learning Confirmation:** Doubly robust AIPW estimation produces a consistent $-6.7\%$ gap, demonstrating that the results are not driven by arbitrary parametric functional form assumptions.
