---
title: "Gender Wage Gap Analysis Using US Current Population Survey Data"
author: "Xiang Ao"
date: "2025-11-17"
---

Traditionally Oaxaca-Blinder decomposition is used to decompose the wage gap into explained and unexplained parts. The explained part is due to differences in observable characteristics, such as education, experience, occupation, etc. The unexplained part is due to discrimination or other unobserved factors.

It has a problem that when decomposing, using female as reference group or male as reference group will give different results.

Słoczyński (2015) and Słoczyński (2013) have shown that a regression adjustment type of analysis is equivalent to Oaxaca-Blinder decomposition in a nonstandard way.

## Oaxaca-Blinder decomposition

Assume the regression model is different for male and female: $$ y_i = X_i \beta_a + \epsilon_a $$

where $a$ is gender (female is 1), $X$'s are person characteristics.

$$ E[y_i | a_i = 1] - E[y_i | a_i = 0] = E[X_i | a_i=1]  (\beta_1 - \beta_0) + (E[X_i | a_i =1 ] - E[X_i | a_i = 0]) \beta_0 $$

or

$$ E[y_i | a_i = 1] - E[y_i | a_i = 0] = E[X_i | a_i=0]  (\beta_1 - \beta_0) + (E[X_i | a_i =1 ] - E[X_i | a_i = 0]) \beta_1 $$ These two equations are different forms of the Oaxaca-Blinder decomposition. The first element is the "unexplained component", which is the difference in the coefficients, and the second element is the "explained component", which is the difference in the means of the covariates.

These two equations can give different results for the two components, one has a baseline of male, and other one female.

We can further shows that

$$
\begin{align}
E[y_i | a_i =1] - E[y_i | a_i=0] &= E[X_i | a_i=1]  (\beta_1 - \beta_0) + (E[X_i | a_i =1 ] - E[X_i | a_i = 0]) \beta_0  \\
&= E[y_i(1) - y_i(0) | a_i=1] + (E[y_i(0) | a_i=1] - E[y_i(0) | a_i = 0]) \\
&= \tau_{PATT} + (E[y_i(0) | a_i=1] - E[y_i(0) | a_i=0])
\end{align}
$$

That is, the observed difference is the sum of $\tau_{PATT}$ and selection bias. It is also known in the OB decomposition literature as the unexplained component and the explained component. The former as "discrimination".

We can simply use Regression Adjustment for these quantities.  The easiest is an OLS of $y_i$ on $a_i$, the demeaned $X_i$, and their interaction; the coefficient on $a_i$ is then the effect evaluated at the mean of the covariates used for the demeaning.  Demean at the **overall** mean and that coefficient is $\tau_{PAGE}$ (the ATE), which is what the example below does; demean at the mean among **women** instead and it is $\tau_{PATT}$.  The choice of centring is what selects the estimand, so it is worth being explicit about which one you have.

He also advocates the "population average gender effect" (PAGE):

$$ \tau_{PAGE} = E[\tau(X_i)] $$

which is basically ATE, or conditional average effect (CATE), then average over the $X$'s.

In this version, the two forms of OB decomposition can be $\tau_{PAGM}$ and $\tau_{PAGW}$, for men and women.

The nice thing about this estimator $\tau_{PAGE}$ is that first, it is consistent with the treatment effect (causal inference) literature, and it is invariant to the choice of reference group.

The second advantage is that we also do not need to assume homogeneity of treatment effect, which is a strong assumption in the Oaxaca-Blinder decomposition.

The third advantage is that we can use nonparametric methods to estimate the treatment effect. Traditional OB decomposition is based on linear regression, which assumes a linear relationship between the covariates and the outcome.

## Example

I am using the data from https://github.com/d2cml-ai/CausalAI-Course/tree/main/data.

"The data set we consider is from the 2015 March Supplement of the U.S. Current Population Survey. We select white non-hispanic individuals, aged 25 to 64 years, and working more than 35 hours per week for at least 50 weeks of the year. We exclude self-employed workers; individuals living in group quarters; individuals in the military, agricultural or private household sectors; individuals with inconsistent reports on earnings and employment status; individuals with allocated or missing information in any of the variables used in the analysis; and individuals with hourly wage below 3. "

I am using this data to study gender wage gap (GWP).

We can model GWP using the potential outcome framework, with some care about what plays the role of "treatment." Gender itself is an immutable characteristic, not something that can be manipulated or randomized — the standard "no causation without manipulation" principle (Holland, 1986) says we should not treat gender itself as a causal treatment. What we really have in mind, and what the "treatment" indicator below is best understood as a proxy for, is something like *being perceived/treated as female by employers* (e.g., employer discrimination), which is at least conceptually manipulable even though we cannot experimentally assign it here. With that caveat in mind, we let $D=1$ if female, $D=0$ if male, and outcome $Y$ is logged wage.

The ATE would be

$$ \delta_{ATE} = E[Y(1) - Y(0)] $$

where $Y(1)$ is the potential outcome if treated, and $Y(0)$ is the potential outcome if not treated.

### Regression adjustment

Since this is to estimate ATE, we can use regression adjustment to estimate the ATE. First let's use de-meaned $X$'s. Then the coefficient on gender is the ATE.

We use These variables as $X$'s:

"shs","hsg","scl","clg","ad","ne","mw","so","we","exp1", "occ", "occ2", "ind", and "ind2".

They are:

"Less then High School","High School Graduate","Some College","College Graduate","Advanced Degree", "Northeast","Midwest","South","West","Experience", "occupation", "occupation 2", "industry", and "industry 2".

I decide to use "occ2" and "ind2" as they are broader categories of occupation and industry. We don't have enough data for finer categories "occ" and "ind".

"exp1" seems to be the only variable that is continuous, let's demean it first. Let's also input "occ2" and "ind2" as factors.

Let's include demeaned "exp1", "sohs", "hsg", "socl", "clg", "mw", "sout", and "we" first.  I leave out "occ2" and "ind2" since there are too many levels to demean.

```{r}
#| label: ra1
#| cache: true
#| warning: false
#| message: false
library(tidyverse)
library(broom)


data <- read_csv("wage2015_subsample_inference.csv") |>
  rename(socl = scl, sohs = shs, sout = so) |>
  mutate(exp_dm = exp1 - mean(exp1, na.rm = TRUE), female=sex, occ2 = factor(occ2), ind2 = factor(ind2)) |>
  mutate(sohs_dm = sohs - mean(sohs, na.rm = TRUE),
         hsg_dm = hsg - mean(hsg, na.rm = TRUE),
         socl_dm = socl - mean(socl, na.rm = TRUE),
         clg_dm = clg - mean(clg, na.rm = TRUE),
         mw_dm = mw - mean(mw, na.rm = TRUE),
         sout_dm = sout - mean(sout, na.rm = TRUE),
         we_dm = we - mean(we, na.rm = TRUE))

glimpse(data)


# construct matrices for estimation from the data

# 1. basic model
reg1 <- lm(lwage ~ female, data=data)
tidy(reg1)

reg2 <- lm(lwage ~ female*(exp_dm + sohs_dm + hsg_dm + socl_dm + clg_dm + mw_dm + sout_dm + we_dm), data=data)
tidy(reg2)
```

We see the raw gap is 3.8%.  Adjusting for some covariates we have 11% gap.

Or, we can use "marginaleffects" instead of demeaning.

```{r}
#| label: ra2
#| cache: true
#| warning: false
#| message: false

reg3 <- lm(lwage ~ female*(exp1 + sohs + hsg+ socl + clg + mw + sout + we ), data=data)

library(marginaleffects)

avg_comparisons(reg3,
                variables = "female")
```

We get the same result as the demeaned regression.

If we want $\tau_{ATT}$, then we can use the "newdata" argument to specify the treatment group,
```{r}
#| label: ra3
#| cache: true
#| warning: false
#| message: false
avg_comparisons(reg3,
                variables = "female",
                newdata = subset(data, female == 1))
```

### Occupation and industry: a bad control problem

Before adding occupation and industry, it is worth being clear about what kind of variables they are. Everything used so far --- experience, education, region --- is plausibly settled before the wage is determined and is not itself much of a consequence of gender. Occupation and industry are different. They are **post-treatment**: whatever "being perceived and treated as female" does, one of the things it does is shift which occupations people end up in.

That makes them *bad controls* in the sense of Cinelli, Forney and Pearl (2022), and for two separate reasons at the same time.

```{r}
#| label: gwg-dag
#| fig-cap: "Occupation is both a mediator of the gender--wage relationship and a collider on a path through unobserved wage determinants $U$."
#| fig-width: 7
#| fig-height: 3.6
#| warning: false
#| message: false

library(ggdag)
library(ggplot2)

g <- dagify(
  Wage ~ Female + Occupation + U,
  Occupation ~ Female + U,
  exposure = "Female",
  outcome  = "Wage",
  coords = list(x = c(Female = 0, Occupation = 1.1, U = 1.1, Wage = 2.2),
                y = c(Female = 0, Occupation = -1,  U = 1,   Wage = 0))
)

# node_size has to stay small: ggdag stops each edge at the node boundary, so
# large nodes swallow the arrowheads -- and the arrowheads are the whole point.
ggdag(g, text = FALSE, node_size = 9) +
  geom_dag_text_repel(aes(label = name), colour = "black", size = 3.2,
                      box.padding = 0.9, seed = 42) +
  theme_dag() +
  expand_limits(x = c(-0.45, 2.65), y = c(-1.5, 1.5))
```

Here $U$ stands for unobserved wage determinants --- drive, negotiation propensity, family constraints, expected career continuity --- and `Occupation` covers occupation and industry together.

**Occupation is a mediator.** Part of the total effect of gender on wages runs `Female -> Occupation -> Wage`. Conditioning on occupation blocks that path, so the coefficient stops estimating a total effect. On its own this would be tolerable, as long as we said so: we would be reporting a within-occupation gap, which is a perfectly coherent descriptive quantity.

**Occupation is also a collider.** It is a common effect of gender and of $U$. Conditioning on a common effect *induces* an association between its causes, so conditioning on occupation opens the non-causal path `Female -> Occupation <- U -> Wage`, which was closed before we touched anything (Elwert and Winship, 2014).

We do not have to take that on faith. `dagitty` will enumerate the paths and say which are open, with and without conditioning on occupation:

```{r}
#| label: gwg-paths
#| warning: false
#| message: false

paths_table <- function(Z = NULL) {
  p <- dagitty::paths(g, from = "Female", to = "Wage", Z = Z)
  data.frame(path = p$paths, open = p$open)
}

paths_table()                   # nothing conditioned on
paths_table(Z = "Occupation")   # conditioning on occupation
```

The two tables are the whole argument. Conditioning on occupation flips `Female -> Occupation -> Wage` from open to closed --- that is the mediation we knowingly gave up --- and simultaneously flips `Female -> Occupation <- U -> Wage` from closed to open. That second flip is the damage. Blocking a mediator removes something we wanted; conditioning on a collider manufactures something we did not. After adjustment, gender and $U$ are correlated *within* occupation cells even if they are independent in the population, and because $U$ also drives wages, that induced correlation leaks into the gender coefficient.

Stated formally: conditioning on a mediator identifies a *controlled direct effect* only under **no unmeasured mediator--outcome confounding** (VanderWeele and Robinson, 2014). $U$ is exactly such a confounder, and in a wage setting it is not a remote worry --- it is more or less why this literature exists.

Two consequences worth flagging before we run the model. First, this is not a problem that a better estimator solves: the AIPW and SuperLearner fits later in this chapter relax functional form, and do nothing at all about a structurally uninterpretable estimand. Second, the same graph applies to the sample itself. This extract keeps only full-year, full-time workers, and employment is likewise a common effect of gender and $U$ --- so the sample restriction is a second conditioning-on-a-collider step, built in before any regression is run.

With all that on the table, here is the model.

```{r}
#| label: ra4
#| cache: true
#| warning: false
#| message: false
reg4 <- lm(lwage ~ female*(exp1 + sohs + hsg+ socl + clg + mw + sout + we + occ2 + ind2), data=data)
avg_comparisons(reg4,
                variables = "female",
                newdata = subset(data, female == 1))
```

So in the model with full interaction, we have a gap of 7.5%, against 11.3% without occupation and industry.

The tempting reading of that pair --- "occupational and industrial sorting accounts for roughly four points of the gap" --- is exactly the reading the graph above forbids. Three things are mixed into that four-point move, and nothing in the output separates them:

1. genuine mediation, the part of the gap that really does operate through sorting;
2. collider bias, from the gender--$U$ association induced by conditioning;
3. ordinary confounding of the occupation--wage relationship.

The likely *direction* of (2) is worth spelling out. If women who enter male-dominated, high-paying occupations are positively selected on $U$ --- they had more to overcome to get there --- then within those occupations they are better on unobservables than the men they are being compared with, which pushes the within-occupation gap **toward zero**. Some of the attenuation from 11.3% to 7.5% may therefore be bias rather than mediation.

The defensible way to report this: 11.3% is the total-effect estimate; 7.5% is a within-occupation descriptive contrast whose causal reading rests on an assumption we cannot verify. What we *can* do is ask how strong $U$ would have to be to account for the whole difference.

#### How strong would $U$ have to be?

The omitted-variable-bias machinery in `sensemakr` (Cinelli and Hazlett, 2020) needs a single treatment coefficient, so for this exercise we drop the interactions and use additive specifications. That costs us the heterogeneity, but the attenuation we are interrogating survives the simplification almost exactly.

```{r}
#| label: cde-sens
#| cache: true
#| warning: false
#| message: false

library(sensemakr)

Xb <- "exp1 + sohs + hsg + socl + clg + mw + sout + we"

fit_total <- lm(as.formula(paste("lwage ~ female +", Xb)), data = data)
fit_adj   <- lm(as.formula(paste("lwage ~ female +", Xb, "+ occ2 + ind2")), data = data)

c(total    = coef(fit_total)["female"],
  adjusted = coef(fit_adj)["female"],
  attenuation = coef(fit_adj)["female"] - coef(fit_total)["female"])

sens <- sensemakr(model = fit_adj, treatment = "female",
                  benchmark_covariates = c("clg", "exp1"),
                  kd = c(1, 2, 3))
summary(sens)
```

The additive models put the total effect at $-11.4\%$ and the occupation-adjusted effect at $-7.3\%$, an attenuation of $4.1$ log points --- essentially the same gap as in the interacted specifications.

Now invert the bias formula. Instead of asking how strong a confounder must be to drive the adjusted estimate to zero (the usual question), ask how strong it must be to move the adjusted estimate *back* to the total effect. That is the strength at which the entire apparent contribution of occupation is bias rather than mediation.

```{r}
#| label: cde-invert
#| cache: true
#| warning: false
#| message: false

target <- coef(fit_total)["female"]

grid <- expand.grid(r2dz = c(0.0125, 0.02, 0.033, 0.06),
                    r2yz = seq(0.001, 0.60, length.out = 4000))
grid$adj <- adjusted_estimate(fit_adj, treatment = "female",
                              r2dz.x = grid$r2dz, r2yz.dx = grid$r2yz,
                              reduce = FALSE)

grid |>
  group_by(r2dz) |>
  slice_min(abs(adj - target), n = 1) |>
  ungroup() |>
  transmute(`partial R2 with gender` = r2dz,
            `partial R2 with wage`   = round(r2yz, 3),
            `restored estimate`      = round(adj, 4))
```

```{r}
#| label: cde-contour
#| fig-cap: "Adjusted estimate of the gender gap as a function of a hypothetical confounder's partial $R^2$ with gender (horizontal) and with log wage (vertical), benchmarked against college and experience."
#| fig-width: 7
#| fig-height: 5
#| warning: false
#| message: false

# sensemakr is attached again here on purpose: the chunk above is cached, so on a
# cache hit its library() call never runs and plot() would dispatch to
# plot.default instead of the sensemakr method.
library(sensemakr)

plot(sens)
```

The answer is uncomfortable. A confounder explaining about **3% of the residual variance of gender and about 4% of the residual variance of wages** is enough to restore the full total effect --- that is, enough for occupation to be contributing no genuine mediation at all. For scale, `sensemakr` reports a robustness value of $6.6\%$: a confounder explaining more than that share of both residual variances would drive the adjusted estimate to zero. So a $U$ roughly *half* as strong as the one needed to erase the within-occupation gap entirely is already strong enough to explain the whole $11.3\% \to 7.5\%$ attenuation.

The benchmark rows give another angle. Experience explains $3.4\%$ of the residual variance of wages but only $0.16\%$ of the residual variance of gender, so along the outcome dimension the required $U$ is only a little stronger than experience. Along the gender dimension the required multiple looks large, but that is mostly because gender is close to orthogonal to experience in this sample; in absolute terms we are still only asking for a few percent of residual variance.

None of this shows that occupational sorting explains nothing. It shows that this design cannot distinguish "sorting explains four points" from "conditioning on a collider costs four points," and that the confounder needed to tip the balance is unremarkable in size. Which is why the total effect, not the occupation-adjusted number, is the estimate to lead with.

Now let's try nonparametric estimation.


### Nonparametric estimation

If we'd like to go for nonparametric, relaxing linearity assumption, then we can use "npcausal". Note this may take a while, depending on which machine learning packages you decide to include. We can also use other nonparametric estimators, such as TMLE or doubleML.

```{r}
#| label: npcausal1
#| cache: true
#| warning: false
#| message: false


library(npcausal)

#SL.library <- c("SL.earth","SL.glmnet","SL.glm.interaction", "SL.mean","SL.ranger", "SL.xgboost")


SL.library <- c("SL.earth","SL.glmnet","SL.mean","SL.ranger", "SL.xgboost")

# construct W from model matrix of reg3
reg3 <- lm(lwage ~ exp_dm + sohs + hsg+ socl + clg + mw + sout + we + occ2 + ind2, data=data)
W <- as.data.frame(model.matrix(reg3)[, -1]) # remove intercept

Y <- data$lwage
A <- data$female
set.seed(123)
aipw_ate <- ate(y=Y, a=A, x=W, nsplits=5, sl.lib=SL.library)

aipw_att <- att(y=Y, a=A, x=W, nsplits=5, sl.lib=SL.library)
```

The two fits above print only SuperLearner's progress bars, so the estimates
themselves have to be asked for:

```{r}
#| label: npcausal-results
#| warning: false
#| message: false

aipw_ate$res
aipw_att$res
```

Two estimates come out of this: an AIPW **ATE** of $-5.8\%$ (95% CI $-8.9\%$ to
$-2.7\%$) and an AIPW **ATT** of $-6.7\%$ (95% CI $-9.5\%$ to $-3.9\%$).

The comparison worth making is like with like. The $-7.5\%$ from the fully
interacted linear model above was an ATT (it was computed on
`newdata = subset(data, female == 1)`), so it should be set against the
nonparametric **ATT** of $-6.7\%$, not against the ATE. Those two agree closely,
which is reassuring: dropping the linearity assumption altogether moves the
estimate by less than a percentage point, and both use the same covariate set
including occupation and industry.

That last clause is also the caveat. Because both specifications condition on
occupation and industry, both inherit the bad-control problem set out above: the
agreement between $-7.5\%$ and $-6.7\%$ tells us the two estimators recover the
same quantity, not that the quantity is a causal effect. Relaxing linearity
addresses functional form; it leaves the collider untouched.

The ATE and ATT point estimates do differ a little, in the direction of a larger
gap among women than in the sample as a whole — which would indicate effect
heterogeneity across the covariates. Their confidence intervals overlap heavily,
though, so this data cannot really distinguish them.
