---
title: "Interaction term in a non-linear model"
date: "2017-12-07"
---



## Interaction term in a non-linear model

In a non-linear model (for example, logit or poisson model), the interpretation of the coefficient on the interaction term is tricky.  [Ai and Norton (2003)](https://pdfs.semanticscholar.org/6285/8e64d9a337504d72cb862c4cc1e7fd27a7a0.pdf) points out that the interaction term coefficient is not the same as people can interpret as in a linear model; that is, how much effect of $x1$ changes with the value of $x2$.  They interpret this as a cross partial derivative of the nonlinear response function $E(y)$ with respect to $x_1$ and $x_2$ — a quantity that involves the model's other coefficients and the values of all covariates, not just the estimated interaction coefficient $\beta_{12}$. As a result, this cross partial derivative can differ in magnitude, sign, and even statistical significance from $\beta_{12}$ itself.

If we have a linear model with interaction:

$$ E(y) = \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1*x_2 $$ {#eq-interaction-nonlinear-1}

Then, the marginal effect

$$ \frac{\partial^2 E(y)}{\partial x_1 \partial x_2} = \beta_{12} $$ {#eq-interaction-nonlinear-2}

That is, $\beta_{12}$ is the second derivative of $E(y)$ on $x_1$ and $x_2$.  The marginal effect of $x_1$ is $\frac{\partial E(y)}{\partial x_1} = \beta_1 + \beta_{12} x_2$, which is linear in $x_2$ but does not itself depend on any nonlinear transformation, since the model is linear.

In a non-linear model,

$$ F(E(y)) = \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1*x_2 $$ {#eq-interaction-nonlinear-3}

$$ \frac{\partial^2 F(E(y))}{\partial x_1 \partial x_2} = \beta_{12} $$ {#eq-interaction-nonlinear-4}

Here, the partial derivative of $F(E(y))$ on $x_1$ and $x_2$ is still $\beta_{12}$.  However, most people are interested in $\frac{\partial^2 E(y)}{\partial x_1 \partial x_2}$.  

$$ \frac{\partial^2 E(y)}{\partial x_1 \partial x_2} = \beta_{12} G'(\eta) + (\beta_{1} + \beta_{12} x_2)(\beta_2 + \beta_{12} x_1) G''(\eta)$$ {#eq-interaction-nonlinear-5}

where $G()$ is the inverse function of $F()$ and $\eta$ is the linear predictor.

It is true that in a non-linear model with interaction, the marginal effect of $x_1$ differs with different values of $x_2$.  However, even if we have a non-linear model without interaction, the marginal effect of $x_1$ is still different with different values of $x_2$. To see this, consider a nonlinear model with no interaction term at all:

$$ F(E(y)) = \beta_1 x_1 + \beta_2 x_2 $$ {#eq-interaction-nonlinear-6}

$$ \frac{\partial^2 E(y)}{\partial x_1 \partial x_2} =  (\beta_{1} \beta_2 ) G''(\eta)$$ {#eq-interaction-nonlinear-7}

Therefore, when we set up our model, 

$$ F(E(y)) = \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1*x_2 $$ {#eq-interaction-nonlinear-8}

we have in mind that we allow interaction of $x_1$ and $x_2$ to interact for the effect on $F(E(y))$; not on $E(y)$.

We agree with Bill Greene (2013), "Interaction Effects in Models for Nonlinear
Outcomes" (Lugano lecture notes; the copy formerly at
`people.stern.nyu.edu/wgreene/Lugano2013/` is no longer online, but the argument
appears in Greene's *Econometric Analysis*, 8th ed., sec. 17.3). In
a nonlinear model, the partial effects (as Greene calls it) is
nonlinear, regardless of the model.  For example, in a logit model,
even if you don't have an interaction term in your model, the effect
of $x_1$ will still be different for every value of $x_2$, simply
because it's a nonlinear model.

 As Greene put it at the summary section,
"Build the model based on appropriate statistical procedures and principles. Statistical testing
about the model specification is done at this step Hypothesis tests are about model coefficients
and about the structural aspects of the model specifications. Partial effects are neither
coefficients nor elements of the specification of the model. They are implications of the specified
and estimated model."

We also agree with [Maarten Buis 2010](http://www.stata-journal.com/sjpdf.html?articlenum=st0194), that we should use multiplicative effect in a non-linear model.  That is, in a non-linear model,

$$ F(E(y)) = \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1*x_2 $$ {#eq-interaction-nonlinear-9}

We should pay more attention to 

$$ \frac{\partial^2 F(E(y))}{\partial x_1 \partial x_2} = \beta_{12} $$ {#eq-interaction-nonlinear-10}

For example, in a logit model,

$$ log(P(y=1)/(1-P(y=1))) = \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1*x_2 $$ {#eq-interaction-nonlinear-11}

That is, the log of odds is a linear function of $x_1$ and $x_2$ and interaction.  The interaction effect has the same interpretation as the linear model, in terms of log of odds.  

Or, it becomes multiplicative effect when we talk about odds ratios.  Stata's "margins" command is a great tool to calculate marginal effects in various situations, as shown in [Maarten Buis 2010](http://www.stata-journal.com/sjpdf.html?articlenum=st0194).

## A worked example

Let's illustrate the Ai-Norton point directly: in a logit model with an interaction term, the cross partial derivative $\frac{\partial^2 E(y)}{\partial x_1 \partial x_2}$ (what we actually care about) is not the same object as the estimated coefficient $\beta_{12}$ on $x_1 x_2$ in the linear index. We simulate data with a true interaction effect on the log-odds scale, fit a logit model, and then use the `marginaleffects` package for two distinct things: first the marginal effect of $x_1$ on the probability scale at several values of $x_2$, and then the cross partial derivative itself.

```{r}
#| label: ai-norton-example
#| cache: true
#| warning: false
#| message: false
library(dplyr)
library(marginaleffects)

set.seed(42)
n <- 2000
x1 <- rnorm(n)
x2 <- rnorm(n)
eta <- -0.3 + 0.8 * x1 + 0.5 * x2 + 0.6 * x1 * x2
p <- plogis(eta)
y <- rbinom(n, 1, p)
dat <- data.frame(y, x1, x2)

m <- glm(y ~ x1 * x2, data = dat, family = binomial)
summary(m)$coefficients
```

The coefficient on `x1:x2` above is $\hat\beta_{12}$, the interaction effect on the log-odds (linear index) scale — constant by construction. On the *probability* scale nothing is constant. Start with the marginal effect of $x_1$ itself, evaluated at three values of $x_2$:

```{r}
#| label: ai-norton-slopes
#| cache: true
#| warning: false
#| message: false
slopes(m, variables = "x1", by = "x2", newdata = datagrid(x2 = c(-1, 0, 1)))
```

The marginal effect of $x_1$ on $P(y=1)$ changes across the three values of
$x_2$ — and notice that at $x_2=-1$ it is not significant ($p \approx 0.11$)
even though $\hat\beta_{12}$ is significant at better than $0.001$. Already the
probability-scale answer and the coefficient tell different stories.

The cross partial derivative is a *further* step: not the effect of $x_1$ at a
given $x_2$, but how much that effect changes as $x_2$ moves. That is a
difference of slopes, so it needs a hypothesis test on top of the `by` argument
rather than a second call to `slopes()`:

```{r}
#| label: ai-norton-cross-partial
#| cache: true
#| warning: false
#| message: false
slopes(m, variables = "x1", by = "x2",
       newdata = datagrid(x2 = c(-1, 1)),
       hypothesis = "b2 - b1 = 0")
```

Here the cross partial comes out large and highly significant, which is worth
being honest about: in *this* example it agrees with $\hat\beta_{12}$ in
significance, differing only in scale and magnitude. Ai and Norton's warning is
that the two quantities *need not* agree — not that they never do. The
dependable lesson is the one visible in the table above: on the probability
scale the answer depends on where you evaluate it, so a single coefficient
cannot summarise it.

