---
title: "Interaction term in a non-linear model"
date: "2017-12-07"
---



## The issue

In a linear model with interaction,

$$ E(y) = \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1 x_2, $$ {#eq-interaction-nonlinear-1}

the cross partial derivative is simply the coefficient:

$$ \frac{\partial^2 E(y)}{\partial x_1 \partial x_2} = \beta_{12}. $$ {#eq-interaction-nonlinear-2}

In a nonlinear model — logit, Poisson, or any link function $F$ — the same coefficient $\beta_{12}$ sits on the *link* scale:

$$ F(E(y)) = \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1 x_2. $$ {#eq-interaction-nonlinear-3}

The cross partial of $F(E(y))$ is still $\beta_{12}$. But the cross partial of $E(y)$ itself — the quantity most researchers want — is not:

$$ \frac{\partial^2 E(y)}{\partial x_1 \partial x_2} = \beta_{12} G'(\eta) + (\beta_{1} + \beta_{12} x_2)(\beta_2 + \beta_{12} x_1) G''(\eta), $$ {#eq-interaction-nonlinear-5}

where $G = F^{-1}$ and $\eta$ is the linear predictor. This is the point of [Ai and Norton (2003)](https://pdfs.semanticscholar.org/6285/8e64d9a337504d72cb862c4cc1e7fd27a7a0.pdf): the probability-scale interaction effect depends on every coefficient, varies across the covariate space, and can differ in sign and significance from $\beta_{12}$.

The deeper issue is that nonlinearity alone generates a nonzero cross partial — even without an interaction term. Drop $\beta_{12}$ and the cross partial on $E(y)$ is still $\beta_1 \beta_2 G''(\eta) \ne 0$ whenever the link is curved. The marginal effect of $x_1$ depends on $x_2$ in any nonlinear model, interaction or not.

Greene (*Econometric Analysis*, 8th ed., sec. 17.3) and [Buis (2010)](http://www.stata-journal.com/sjpdf.html?articlenum=st0194) draw the same conclusion: the model is specified on the link scale, and that is where coefficients and their tests are interpretable. In a logit, $\beta_{12}$ is the interaction on log-odds; exponentiating gives an odds-ratio interaction that is multiplicative. Partial effects on the probability scale are *implications* of the model, not parameters of it — they should be computed from the fitted model (via `margins` or `marginaleffects`), not read off the coefficient table.

## A worked example

Let's illustrate the Ai-Norton point directly: in a logit model with an interaction term, the cross partial derivative $\frac{\partial^2 E(y)}{\partial x_1 \partial x_2}$ (what we actually care about) is not the same object as the estimated coefficient $\beta_{12}$ on $x_1 x_2$ in the linear index. We simulate data with a true interaction effect on the log-odds scale, fit a logit model, and then use the `marginaleffects` package for two distinct things: first the marginal effect of $x_1$ on the probability scale at several values of $x_2$, and then the cross partial derivative itself.

```{r}
#| label: ai-norton-example
#| cache: true
#| warning: false
#| message: false
library(dplyr)
library(marginaleffects)

set.seed(42)
n <- 2000
x1 <- rnorm(n)
x2 <- rnorm(n)
eta <- -0.3 + 0.8 * x1 + 0.5 * x2 + 0.6 * x1 * x2
p <- plogis(eta)
y <- rbinom(n, 1, p)
dat <- data.frame(y, x1, x2)

m <- glm(y ~ x1 * x2, data = dat, family = binomial)
summary(m)$coefficients
```

The coefficient on `x1:x2` above is $\hat\beta_{12}$, the interaction effect on the log-odds (linear index) scale — constant by construction. On the *probability* scale nothing is constant. Start with the marginal effect of $x_1$ itself, evaluated at three values of $x_2$:

```{r}
#| label: ai-norton-slopes
#| cache: true
#| warning: false
#| message: false
slopes(m, variables = "x1", by = "x2", newdata = datagrid(x2 = c(-1, 0, 1)))
```

The marginal effect of $x_1$ on $P(y=1)$ changes across the three values of
$x_2$ — and notice that at $x_2=-1$ it is not significant ($p \approx 0.11$)
even though $\hat\beta_{12}$ is significant at better than $0.001$. Already the
probability-scale answer and the coefficient tell different stories.

The cross partial derivative is a *further* step: not the effect of $x_1$ at a
given $x_2$, but how much that effect changes as $x_2$ moves. That is a
difference of slopes, so it needs a hypothesis test on top of the `by` argument
rather than a second call to `slopes()`:

```{r}
#| label: ai-norton-cross-partial
#| cache: true
#| warning: false
#| message: false
slopes(m, variables = "x1", by = "x2",
       newdata = datagrid(x2 = c(-1, 1)),
       hypothesis = "b2 - b1 = 0")
```

Here the cross partial comes out large and highly significant, which is worth
being honest about: in *this* example it agrees with $\hat\beta_{12}$ in
significance, differing only in scale and magnitude. Ai and Norton's warning is
that the two quantities *need not* agree — not that they never do. The
dependable lesson is the one visible in the table above: on the probability
scale the answer depends on where you evaluate it, so a single coefficient
cannot summarise it.

