---
title: "Interpreting interaction in a regression model"
date: "2017-12-07"
---



## Interaction with two binary variables

Start with the simplest case: two binary variables $x_1$ and $x_2$ coded 0/1.

$$ E(y) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1x_2 $$ {#eq-interaction-regression-1}

The model is saturated — three coefficients for three contrasts from the baseline $(x_1, x_2) = (0, 0)$, one for each of the other three cells. Often only the interaction term gets attention. When the two variables are treatment and group, the interaction is the difference in treatment effects between groups, and that may be all that matters. But in other designs all four cells are substantively interesting, and then $\beta_1$ and $\beta_2$ matter too.

Consider the model:

$$ E(y) = \beta_0 + \beta_1 female + \beta_2 treatment + \beta_{12} female*treatment $$ {#eq-interaction-regression-2}

The four cell means are:


|               |  male                   |  female                                        |
| ------------- |:-----------------------:|:----------------------------------------------:|
| control       |  $$ \beta_0 $$ {#eq-interaction-regression-3}          |  $$ \beta_0 + \beta_1 $$ {#eq-interaction-regression-4}                        |
| treatment     | $$ \beta_0 + \beta_2 $$ {#eq-interaction-regression-5} |  $$ \beta_0 + \beta_1 + \beta_2 + \beta_{12}$$ {#eq-interaction-regression-6} |

Each $\beta$ reads off the table: $\beta_0$ is the baseline cell mean, $\beta_1$ the female premium among controls, $\beta_2$ the treatment effect among males. The interaction

$$ \beta_{12} = (E(Y|(1,1))-E(Y|(0,1)))-(E(Y|(1,0))-E(Y|(0,0))) $$ {#eq-interaction-regression-9}

is the difference in differences — the treatment effect for females minus the treatment effect for males. When the research question is "does the treatment effect differ by group?", the interaction term is the answer. Replace female/male with before/after and the same expression is the standard DiD estimator.

But sometimes the interest is in all four cells. Union membership crossed with race, for example — we may want all six pairwise comparisons, not just the interaction. Stata's `margins` command computes these directly:

```{r}
#| label: stata-chunk
#| engine: 'stata'
#| engine.path: '/usr/local/bin/stata'
#| cache: true

webuse union3
reg ln_wage i.union##i.black, r
margins union#black
margins union#black, pwcompare

```

The `margins` call reports the four cell means; `pwcompare` gives all six pairwise comparisons. In this example, all six pairwise contrasts are significant at 95% — yet the interaction term is insignificant ($p = 0.455$). The two facts are compatible. Each pairwise comparison asks whether two cells differ from each other. The interaction asks whether the *difference between two of those differences* is non-zero — a second-order quantity with its own, larger standard error. Here the union premium is clearly positive for both groups, but the gap between the two premiums ($0.039$) is not distinguishable from zero.


## Interaction with continuous variables

With continuous $x_1$ and $x_2$,

$$ E(y) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1 x_2 $$ {#eq-interaction-regression-10}

centering both variables — subtracting each mean before forming the product — is a linear transformation that leaves fitted values, $R^2$, and everything about $\beta_{12}$ unchanged. What it changes is $\beta_1$ and $\beta_2$.

First, it reduces collinearity. When $x_1$ and $x_2$ are positive, $x_1 x_2$ is mechanically correlated with both, inflating the standard errors of $\beta_1$ and $\beta_2$. Centering breaks most of that correlation (all of it under normality), stabilizing the main-effect estimates without altering the interaction.

Second, it fixes the interpretation. In the uncentered model, $\beta_1$ is the effect of $x_1$ when $x_2 = 0$, and zero may be outside the support of $x_2$. After centering, $\beta_1$ is the effect of $x_1$ evaluated at the mean of $x_2$ — a quantity that is always interpretable. When a dummy interacts with a continuous variable, only the continuous variable should be centered.

Stata's `margins` command computes the conditional predictions directly:


```{r}
#| label: stata-chunk2
#| engine: 'stata'
#| engine.path: '/usr/local/bin/stata'
#| cache: true

sysuse auto
sum mpg
gen mpg_centered=mpg-r(mean)
sum mpg_centered
reg price i.foreign##c.mpg_centered
margins foreign, at(mpg_centered=(-3 (1) 3))
marginsplot

```

The graph shows predicted price for foreign and domestic cars across the range of mpg.
