---
title: "Modeling Variation, Not Just the Mean, in Likert-Scale Data"
date: "2026-07-16"
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(brms)
```

## The Problem: The Mechanical Ceiling on Dispersion

When analyzing Likert-scale ratings (e.g., 1–7 satisfaction or agreement scales), researchers often want to know whether certain groups exhibit greater disagreement or dispersion than others. A common approach is to calculate the standard deviation for each unit and regress it on covariates.

On a bounded scale $[a, b]$, however, variance is mechanically constrained by the mean. By the Bhatia–Davis (2000) inequality, for any random variable bounded in $[a, b]$ with mean $\mu$:

$$\text{Var}(X) \leq (\mu - a)(b - \mu)$$ {#eq-bhatia-davis}

On a 1–7 Likert item, the maximum possible standard deviation is:

$$\text{SD}_{\max}(\mu) = \sqrt{(\mu - 1)(7 - \mu)}$$

```{r}
#| label: bound-curve
#| echo: true
#| fig-cap: "The mechanical ceiling on SD as a function of the mean on a 1–7 scale."

mu_grid <- seq(1, 7, length.out = 200)
sd_max  <- sqrt(pmax((mu_grid - 1) * (7 - mu_grid), 0))

ggplot(data.frame(mu = mu_grid, sd_max = sd_max), aes(mu, sd_max)) +
  geom_line(linewidth = 1, color = "#2d5fa3") +
  labs(x = "Mean Rating (μ)", y = expression(SD[max](mu)),
       title = "Maximum Possible SD as a Function of the Mean") +
  theme_minimal(base_size = 12)
```

The maximum standard deviation peaks at mid-scale ($\mu = 4, \text{SD}_{\max} = 3$) and drops to zero at the boundaries ($\mu = 1$ and $\mu = 7$). When ratings are skewed toward the ceiling—typical of satisfaction or performance evaluations—groups with higher means have mechanically smaller standard deviations, even if their true underlying dispersion is identical.

### Why Standard Ad-Hoc Adjustments Fail

1. **Coefficient of Variation ($\text{SD}/\mu$):** Dividing by the mean assumes variance scales proportionally with the mean, as in count data. On a bounded scale, variance pinches at *both* ends; dividing by $\mu$ simply shifts the peak from $\mu = 4$ to $\mu = 1.75$ without eliminating mean dependence.
2. **Gaussian Location-Scale Models:** Linear mixed-effects models that specify `sigma ~ group` assume an unbounded normal distribution. Because the model is unaware of the scale limits, boundary compression is misinterpreted as smaller residual variance.

## The Solution: Ordinal Cumulative-Probit Location-Scale Model

The appropriate approach treats Likert responses as a discretized continuous latent variable $Y^*$:

$$Y_i^* = \eta_i + \sigma_i \epsilon_i, \qquad \epsilon_i \sim N(0, 1)$$

where $\eta_i$ is the location predictor and $\sigma_i$ is the latent residual standard deviation. The observed rating $Y_i \in \{1, \dots, K\}$ is determined by ordered cutpoints $\theta_1 < \dots < \theta_{K-1}$:

$$Y_i = k \iff \theta_{k-1} < Y_i^* \leq \theta_k$$

Under a probit link, the cumulative probability is:

$$P(Y_i \leq k) = \Phi\left(\frac{\theta_k - \eta_i}{\sigma_i}\right) = \Phi\bigl(\text{disc}_i (\theta_k - \eta_i)\bigr)$$

where discrimination is defined as the inverse of latent dispersion: $\text{disc}_i \equiv 1/\sigma_i$.

In `brms`, we model $\log(\text{disc}_i)$ directly. Higher discrimination corresponds to smaller latent dispersion (more consensus), while lower discrimination indicates greater spread. Because the cutpoints $\theta$ are estimated on an unbounded latent scale, boundary compression no longer biases the dispersion parameter.

## Simulation: Identical Dispersion with Different Means

We simulate 100 subjects per group, each evaluated by 6 raters. Both groups share the identical latent standard deviation ($\sigma = 1$), but Group A's latent mean is near the ceiling ($\mu_A = 2.4$), while Group B is centered mid-range ($\mu_B = 0.6$).

```{r}
#| label: simulate
#| echo: true
#| cache: true

set.seed(42)

n_per_group <- 100
n_raters    <- 6
sigma_true  <- 1.0   # Identical latent dispersion across groups
mu_A <- 2.4          # Group A: high mean (near ceiling)
mu_B <- 0.6          # Group B: mid-scale mean

subjects <- data.frame(
  subject = 1:(2 * n_per_group),
  group   = rep(c("A", "B"), each = n_per_group),
  mu      = rep(c(mu_A, mu_B), each = n_per_group)
)

df <- subjects[rep(seq_len(nrow(subjects)), each = n_raters), ]
df$latent <- rnorm(nrow(df), mean = df$mu, sd = sigma_true)

# Discretize using 6 fixed cutpoints into categories 1 through 7
cuts <- c(-2, -1.2, -0.5, 0.2, 1.0, 1.8)
df$rating     <- as.integer(cut(df$latent, breaks = c(-Inf, cuts, Inf), labels = 1:7))
df$rating_ord <- factor(df$rating, ordered = TRUE, levels = 1:7)
```

### Observed Ratings and the Apparent Gap

```{r}
#| label: raw-summary
#| echo: true

subj_summary <- df |>
  group_by(subject, group) |>
  summarise(mean_rating = mean(rating), sd_rating = sd(rating), .groups = "drop")

subj_summary |>
  group_by(group) |>
  summarise(mean_rating = round(mean(mean_rating), 2),
            mean_sd     = round(mean(sd_rating), 2)) |>
  knitr::kable(caption = "Observed summary statistics by group.")
```

```{r}
#| label: scatter
#| echo: true
#| fig-cap: "Subject-level mean vs. SD against the theoretical Bhatia-Davis bound."

ggplot(subj_summary, aes(mean_rating, sd_rating, color = group)) +
  geom_jitter(width = 0.05, height = 0, alpha = 0.5) +
  geom_line(data = data.frame(mean_rating = mu_grid, sd_rating = sd_max),
            aes(mean_rating, sd_rating), color = "grey40", inherit.aes = FALSE) +
  labs(x = "Subject Mean Rating", y = "Subject SD",
       title = "Boundary Effect: Group A's SD is Mechanically Compressed") +
  theme_minimal(base_size = 12)
```

Group A averages an observed SD of 0.58 compared to 1.22 for Group B. Looking only at raw statistics suggests that raters agree much more about Group A.

## Model Comparison: Gaussian vs. Ordinal Location-Scale

We fit two models in `brms`:
1. **Gaussian MELSM:** Models $\log(\sigma)$ as a function of group.
2. **Ordinal Cumulative-Probit MELSM:** Models $\log(\text{disc})$ with Group A fixed at reference ($\text{disc}_A = 1$).

```{r}
#| label: fit-models
#| echo: true
#| cache: true
#| message: false
#| warning: false

# 1. Gaussian location-scale model
fit_gaussian <- brm(
  bf(rating ~ group + (1 | subject), sigma ~ group),
  data = df, family = gaussian(),
  chains = 2, iter = 2000, warmup = 1000, cores = 2, seed = 1,
  refresh = 0, silent = 2
)

# 2. Ordinal cumulative-probit location-scale model
fit_ordinal <- brm(
  bf(rating_ord ~ group + (1 | subject), disc ~ 0 + group),
  data = df, family = cumulative("probit"), init = 0,
  prior = c(
    prior(constant(0), class = "b", coef = "groupA", dpar = "disc"),
    prior(normal(0, 1), class = "b", coef = "groupB", dpar = "disc")
  ),
  chains = 2, iter = 2000, warmup = 1000, cores = 2, seed = 1,
  refresh = 0, silent = 2
)
```

```{r}
#| label: compare-results
#| echo: true

knitr::kable(round(fixef(fit_gaussian), 2), caption = "Gaussian MELSM: Incorrectly finds group differences in residual SD.")
knitr::kable(round(fixef(fit_ordinal), 2), caption = "Ordinal MELSM: Correctly recovers equal latent dispersion.")
```

### Interpretation

- **Gaussian Model:** The coefficient on `sigma_groupB` is $0.68$, indicating that Group B's residual standard deviation is estimated to be $\exp(0.68) \approx 1.97$ times larger than Group A's ($95\%\text{ CI}: [1.82, 2.14]$). This large effect is an artifact of the scale ceiling.
- **Ordinal Model:** The coefficient on `disc_groupB` is $-0.01$ ($95\%\text{ CI}: [-0.17, 0.15]$), corresponding to a latent standard deviation ratio of $\exp(0.01) \approx 1.01$. The ordinal model correctly recovers that the underlying dispersion is identical.

## Benchmarking Relative Dispersion: The $R$ Metric

An intuitive distribution-free alternative is the ratio of observed variance to maximum possible variance:

$$R = \frac{\text{Var}(Y)}{\text{Var}_{\max}(\mu)} \in [0, 1]$$

where $R = 0$ indicates perfect agreement and $R = 1$ indicates maximum polarization.

When computed on small individual clusters ($n = 6$), $R$ can be noisy or undefined if all raters choose 7 ($0/0$). However, when evaluated at the group level pooling all ratings, $R$ accurately captures dispersion:

```{r}
#| label: r-pooled
#| echo: true

df |>
  group_by(group) |>
  summarise(
    mean_rating = mean(rating),
    var_obs     = mean((rating - mean(rating))^2),
    var_max     = (mean(rating) - 1) * (7 - mean(rating)),
    R_pooled    = round(var_obs / var_max, 3)
  ) |>
  knitr::kable(caption = "Group-level pooled R metric.")
```

Both groups yield $R \approx 0.21–0.22$, correctly reflecting identical relative dispersion.

## Summary

1. **Mechanical Confound:** On bounded scales, variance is bounded by $(\mu - a)(b - \mu)$. Skewed ratings create spurious group differences in raw standard deviations.
2. **Gaussian Failure:** Linear mixed-effects location-scale models (`sigma ~ x`) misinterpret ceiling compression as true consensus.
3. **Ordinal Solution:** Cumulative probit location-scale models (`disc ~ x`) model dispersion on the latent scale, separating true variance differences from boundary artifacts.
4. **Descriptive Metric:** When a simple summary statistic is required, group-level pooled $R = \text{Var} / \text{Var}_{\max}$ removes scale-boundary effects without fitting complex models.
