---
title: "Modeling variation, not just the mean, in Likert-scale data"
date: "2026-07-16"
---

```{r}
#| label: setup
#| include: false

# Loaded here (uncached) rather than inside a cache: true chunk below, so
# these packages are attached to the search path even when downstream
# chunks are skipped and lazy-loaded from a valid cache.
library(ggplot2)
library(dplyr)
library(brms)
```

## When you model the variance of a Likert item

Most of the time, when we regress a Likert-scale outcome (a 1--7 satisfaction
item, a 1--5 agreement item, a peer-evaluation rating) on some covariates, we
only model the mean. But plenty of real questions are about the *spread*, not
the average: do evaluators agree more about some employees than others? Does
one group of raters show more disagreement than another? Is a training
program tightening or loosening the distribution of responses, not just
shifting its center?

The natural instinct is to compute an SD (or variance) for each unit --- each
subject's ratings across raters, each respondent's answers across items --- and
regress *that* on your covariates, the same way you'd regress a mean. This
chapter is about why that instinct runs into a specific, avoidable problem on
a bounded scale, and what to do about it instead.

## The problem: on a bounded scale, variance is mechanically tied to the mean

Here is the part that is easy to miss: on a scale with a hard floor and
ceiling, the *maximum possible* variance a set of responses can depend
on where their mean sits. This isn't a tendency in typical data -- it's a
mathematical fact, true for any distribution on a bounded interval.

For any random variable $X$ confined to $[a,b]$, both $(X-a)$ and $(b-X)$ are
non-negative for every draw, so their product is too:

$$(X-a)(b-X) \geq 0 \quad \text{for every } X \in [a,b]$$

Taking expectations preserves the inequality, and expanding the left side
gives $E[X^2] \leq (a+b)\mu - ab$ where $\mu = E[X]$. Substituting into
$\text{Var}(X) = E[X^2] - \mu^2$:

$$\text{Var}(X) \leq (a+b)\mu - ab - \mu^2 = (\mu - a)(b - \mu)$$

This is the **Bhatia--Davis inequality** (Bhatia and Davis 2000). Equality is
achieved, not just approached: put all the mass at the two endpoints, with
$P(X=a) = (b-\mu)/(b-a)$ and $P(X=b) = (\mu-a)/(b-a)$, and you get variance
exactly $(\mu-a)(b-\mu)$ -- the most polarized a set of responses can possibly
be, given that mean.

On a 1--7 Likert item, that's

$$\text{SD}_{\max}(\mu) = \sqrt{(\mu-1)(7-\mu)}$$

a downward parabola, widest at the scale's midpoint ($\mu=4$, where
$\text{SD}_{\max}=3$) and pinched to exactly zero at both ends. A respondent
whose raters all pick 7 has, by definition, zero disagreement -- there's
nowhere else for a rating to go.

```{r}
#| label: bound-curve
#| echo: true
#| message: false
#| warning: false
#| fig-cap: "The mechanical ceiling on SD, given the mean, for a 1--7 scale."

mu_grid <- seq(1, 7, length.out = 200)
sd_max  <- sqrt(pmax((mu_grid - 1) * (7 - mu_grid), 0))

ggplot(data.frame(mu = mu_grid, sd_max = sd_max), aes(mu, sd_max)) +
  geom_line(linewidth = 1, color = "#2d5fa3") +
  labs(x = "mean rating (μ)", y = expression(SD[max](mu)),
       title = "Maximum possible SD, as a function of the mean") +
  theme_minimal(base_size = 12)
```

The consequence: any two groups whose *means* differ will have different
*ceilings* on their SD, whether or not anything about their true dispersion
differs. If most of your respondents already sit near one end of the scale
(a very common situation -- satisfaction items, performance ratings, and
health scales are all typically skewed toward one end), this isn't a
corner case. It's the region where the curve above is steepest, and where
raw SD comparisons are most likely to mislead.

::: {.callout-note}
## A finite-sample refinement

The curve above is the *population* bound. For a genuinely finite sample of
$n$ real numbers with mean $\mu$, the true maximum is strictly tighter still:
Ellis (2025) gives it exactly as $\text{Var}_{\max}(n,\mu) = (\mu-1)(7-\mu) -
\tfrac{36}{n}a(1-a)$, where $a$ is the fractional part of $n(\mu-1)/6$, which
simplifies to $(\mu-1)(7-\mu) - 9/n \leq \text{Var}_{\max}(n,\mu) \leq
(\mu-1)(7-\mu)$. At $n=5$ raters, that's up to 1.8 points$^2$ of extra
slack -- worth knowing if your groups are rated by only a handful of people
each, since the population curve above then understates how tight the true
ceiling is.
:::

A natural first fix is to divide the SD (or variance) by the mean, the way
you would for a coefficient of variation. This doesn't work here, because
that correction is designed for a *different* mechanism -- variance scaling
proportionally with the mean, as in count data -- not a ceiling that pinches
toward zero at *both* ends. $\text{SD}_{\max}(\mu)/\mu$ is still a curve
(zero at $\mu=1$, a peak of $0.75$ at $\mu=4$, back to zero at $\mu=7$), just
a differently-shaped one. Dividing by the mean doesn't remove the
mean-dependence; it just relabels it.

## Three ways to actually model it

### Gaussian location-scale model (MELSM)

The simplest fix that people reach for is a mixed-effects location-scale
model: model the mean as usual, and *also* model the residual SD as a
function of covariates.

```r
library(brms)

fit_gaussian <- brm(
  bf(rating ~ group + (1 | subject),
     sigma ~ group),                 # model the SD, not just the mean
  data = df, family = gaussian()
)
```

`brms` puts a log link on `sigma` by default, so the `group` coefficient on
the `sigma` part is a difference in *log*-SD; exponentiate it to get an
SD ratio between groups.

**The catch:** a Gaussian model has no boundary in it. It doesn't know the
scale stops at 1 and 7, so it has no way to distinguish "this group's raters
genuinely agree more" from "this group's mean happens to sit closer to the
ceiling." Both produce the identical signature in the fitted `sigma`.

### What belongs in the dispersion submodel?

A natural worry at this point: why does `sigma ~ group` (or, further below,
`disc ~ group` and $\varphi \sim \text{group}$) only include `group`, when
the *mean* model might include other covariates too -- shouldn't the
dispersion submodel be "whatever is in the model for the mean"? Not
necessarily. There is no requirement that the two submodels share a
right-hand side, and in general they *shouldn't* automatically, because they
answer different questions. The mean model asks what predicts the average
rating; the dispersion model asks what predicts how spread out the ratings
are. In this chapter's simulation `group` is the only variable in the
data-generating process at all -- there is nothing else *to* include, on
either side.

In a real dataset, though, this is a substantive modeling choice, not a
mechanical one, and getting it wrong has a real cost -- symmetric to the
familiar one for the mean. If some covariate $Z$ (rater experience, item
wording, subject-level traits) genuinely affects dispersion and happens to be
correlated with `group`, but $Z$ is left out of the dispersion submodel, its
effect is absorbed into the `group` coefficient on `sigma` (or `disc`, or
$\varphi$) -- an omitted-variable problem for the *variance*, exactly
analogous to the ordinary one for the mean. The fix is the same one you'd
already apply on the mean side: include, in the dispersion submodel,
whichever covariates you have a substantive reason to think shift dispersion,
and give that specification the same scrutiny you'd give the mean model --
neither reflexively copying its formula nor stripping the dispersion
submodel down to only the one comparison you happen to be interested in.

### Beta regression: model precision instead of variance directly

If your outcome can be rescaled to $(0,1)$ (e.g. $(y-1)/6$ for a 1--7 item),
Beta regression gives you a mean/precision parameterization built exactly for
bounded outcomes (Ferrari and Cribari-Neto 2004; R implementation in
Cribari-Neto and Zeileis 2010's `betareg` package). With mean $\mu$ and
precision $\varphi$:

$$\text{Var}(Y) = \frac{(\mu-1)(7-\mu)}{1+\varphi} \qquad \Longrightarrow
\qquad \varphi = \frac{(\mu-1)(7-\mu)}{\text{Var}(Y)} - 1$$

Set $\varphi = 0$ and this *is* the Bhatia--Davis bound -- the maximum
possible variance given $\mu$. So $\varphi$ is dispersion measured *after*
the boundary effect is divided out, and $\varphi \sim \text{group}$ is a
comparison that doesn't inherit the ceiling confound the way raw SD does.

The catch here is smaller but real: it's a continuous distributional
assumption grafted onto what is usually genuinely discrete, ordinal data
(a 1--7 Likert response isn't a continuous proportion), and dividing by a
small-sample variance estimate can behave badly when a handful of raters
happen to agree by chance.

$\varphi$ itself is awkward to report -- it lives on $(0,\infty)$ with no
natural reference point. Its bounded, more interpretable twin is

$$R := \frac{\text{Var}(Y)}{\text{Var}_{\max}(\mu)} = \frac{1}{1+\varphi}$$

By Bhatia--Davis, $\text{Var}(Y) \leq \text{Var}_{\max}(\mu)$ for *any*
distribution on $[1,7]$ -- not just Beta ones -- so $R \in [0,1]$ always,
with no distributional assumption required. $R=1$ means a subject's raters
are as polarized as the ceiling allows, given their mean; $R=0$ means
perfect consensus. It's just $(\mu, \text{Var}(Y))$ re-expressed in
mean-neutral units. We'll come back to $R$ in the worked example below,
because it turns out to illustrate the small-sample catch *very* concretely.

### Ordinal cumulative-probit MELSM: put the boundary inside the model

The cleanest fix treats the 1--7 response as what it is: a discretized slice
of a latent continuous trait, cut by six ordered thresholds.

**The latent-variable model.** Posit an unobserved continuous $Y_i^*$ behind
each rating,

$$Y_i^* = \eta_i + \sigma_i \epsilon_i, \qquad \epsilon_i \sim N(0,1),$$

where $\eta_i$ is the usual linear predictor for *location* (`group` plus a
subject random intercept, say) and $\sigma_i$ is the latent residual SD,
which we will also let depend on covariates. The observed 1--7 rating is a
coarsened version of $Y_i^*$, cut at six fixed thresholds $\theta_1 < \cdots
< \theta_6$ (with $\theta_0=-\infty$, $\theta_7=\infty$):

$$Y_i = k \iff \theta_{k-1} < Y_i^* \leq \theta_k.$$

The probability of falling at or below category $k$ is then

$$P(Y_i \leq k) = P(Y_i^* \leq \theta_k) = P\!\left(\epsilon_i \leq
\frac{\theta_k - \eta_i}{\sigma_i}\right) = \Phi\!\left(\frac{\theta_k -
\eta_i}{\sigma_i}\right),$$

using $\Phi$, the standard normal CDF, for the probit link. Writing
$\text{disc}_i := 1/\sigma_i$ turns this into `brms`'s parameterization
exactly:

$$P(Y_i \leq k) = \Phi\bigl(\text{disc}_i\,(\theta_k - \eta_i)\bigr).$$

So "discrimination" is *literally* the reciprocal of the latent residual SD
-- not an analogy, an algebraic identity -- reached by rescaling the argument
of $\Phi$ rather than dividing by $\sigma_i$ directly. Letting $\text{disc}$
depend on covariates is the ordinal-probit analogue of `sigma ~ group` in the
Gaussian model above, just written in reciprocal form: a positivity-friendly
way to let dispersion vary ($\text{disc} > 0$ everywhere, with no constraint
needed on $\sigma_i$ itself). `brms` puts a log link on `disc`, exactly as it
does on Gaussian `sigma`, so a `disc ~ 0 + group` submodel really means
$\log(\text{disc}_i) = \gamma_{\text{group}[i]}$: a group difference in
*log*-discrimination, parallel to the Gaussian model's log-SD coefficients
(Bürkner and Vuorre 2019).

```r
fit_ordinal <- brm(
  bf(rating ~ group + (1 | subject),
     disc ~ 0 + group),           # log(disc) = 0 + group; disc = 1 / latent SD
  data = df, family = cumulative("probit"), init = 0,
  prior = c(
    prior(constant(0), class = "b", coef = "groupA", dpar = "disc"),
    prior(normal(0, 1), class = "b", coef = "groupB", dpar = "disc")
  )
)
```

The `constant(0)` prior pins the reference group's **log**-discrimination --
the linear predictor $\gamma_A$, not $\text{disc}$ itself -- at zero, so
$\text{disc}_A = \exp(0) = 1$ and the reference group's latent SD is
$1/\text{disc}_A = 1$. (Fixing $\gamma_A=0$ is what makes `disc = 1` and
`SD = 1` coincide for the reference group; it is not a claim that
discrimination itself is zero, which would instead mean an infinite latent
SD.) Only $\gamma_B$, the comparison group's log-discrimination, is estimated
freely: $\text{disc}_B = \exp(\gamma_B)$, so $\gamma_B=0$ means identical
dispersion, $\gamma_B>0$ means Group B is *more* discriminating (a
*smaller* latent SD), and $\gamma_B<0$ the reverse. Because the boundary is
now an explicit part of the model -- six estimated cutpoints on an
*unbounded* latent scale -- a group whose mean sits near 7 doesn't
mechanically show a tighter latent SD the way its raw SD is forced to.

## Worked example: two groups, identical true dispersion, different means

To see the confound in action, simulate two groups of 100 subjects each,
6 raters per subject. Both groups' raters draw from a **latent** continuous
opinion with the *same* true SD ($\sigma = 1$) -- there is no real
disagreement difference to find. Group A's latent mean sits high (near the
ceiling once discretized); Group B's sits in the middle of the range.

```{r}
#| label: simulate
#| echo: true
#| message: false
#| warning: false
#| cache: true

set.seed(42)

n_per_group <- 100
n_raters    <- 6
sigma_true  <- 1.0   # SAME latent SD in both groups -- no real difference
mu_A <- 2.4          # group A's latent mean sits near the ceiling
mu_B <- 0.6          # group B's sits mid-range

subjects <- data.frame(
  subject = 1:(2 * n_per_group),
  group   = rep(c("A", "B"), each = n_per_group),
  mu      = rep(c(mu_A, mu_B), each = n_per_group)
)

df <- subjects[rep(seq_len(nrow(subjects)), each = n_raters), ]
df$latent <- rnorm(nrow(df), mean = df$mu, sd = sigma_true)

# six fixed cutpoints -> seven observed categories, 1 through 7
cuts <- c(-2, -1.2, -0.5, 0.2, 1.0, 1.8)
df$rating     <- as.integer(cut(df$latent, breaks = c(-Inf, cuts, Inf), labels = 1:7))
df$rating_ord <- factor(df$rating, ordered = TRUE, levels = 1:7)
```

Before looking at the raw summary, it is worth seeing exactly where the
group difference enters the code, because it is easy to miss: `rnorm(...,
sd = sigma_true)` uses one `sigma_true` for every row, in both groups --
`group` never touches the standard deviation, only the mean, via `df$mu`.
The two groups differ *only* in where their latent mean sits relative to
the six fixed cutpoints, which are the same `cuts` object for both groups.
That is enough, on its own, to produce very different observed spreads:

```{r}
#| label: category-probs
#| echo: true
#| message: false
#| warning: false

edges <- c(-Inf, cuts, Inf)
tibble(
  category = 1:7,
  prob_A = round(diff(pnorm(edges, mean = mu_A, sd = sigma_true)), 3),
  prob_B = round(diff(pnorm(edges, mean = mu_B, sd = sigma_true)), 3)
) |>
  knitr::kable(caption = "Category probabilities implied by each group's latent mean, with the same cutpoints and the same sigma.")
```

Group A's latent mean (2.4) sits 0.6 SDs above the highest cutpoint (1.8),
so `pnorm(0.6)` $\approx 0.73$: nearly three-quarters of Group A's
individual ratings are collapsed into the single top category, and fewer
than 2% fall below category 5. Group B's latent mean (0.6) sits close to
the middle of the cutpoint range, so its ratings spread across all seven
categories, with no single category exceeding about 30%.

That imbalance -- not a difference in $\sigma$ -- is what shrinks Group A's
raw SD. When roughly three-quarters of a subject's six ratings are forced
into the same top bin by `cut()`, the *discrete* variance across those six
ratings is mechanically small, even though the *continuous* latent draws
behind them are exactly as dispersed as Group B's ($\sigma=1$ for both,
by construction). It is the same Bhatia--Davis mechanism from the
introduction, now visible subject-by-subject through `cut()`'s fixed
boundaries rather than through the population-level $\text{SD}_{\max}(\mu)$
curve directly.

The raw, per-subject summary already shows exactly the pattern that tempts
people into thinking there's a real disagreement difference:

```{r}
#| label: raw-summary
#| echo: true
#| message: false
#| warning: false

subj_summary <- df |>
  group_by(subject, group) |>
  summarise(mean_rating = mean(rating), sd_rating = sd(rating), .groups = "drop")

knitr::kable(
  subj_summary |> group_by(group) |>
    summarise(mean_of_means = round(mean(mean_rating), 2),
              mean_of_sds   = round(mean(sd_rating), 2)),
  caption = "Raw per-subject summary by group -- true latent SD is IDENTICAL in both."
)
```

```{r}
#| label: scatter
#| echo: true
#| message: false
#| warning: false
#| fig-cap: "Subject-level mean vs. SD, with the theoretical ceiling overlaid."

ggplot(subj_summary, aes(mean_rating, sd_rating, color = group)) +
  geom_jitter(width = 0.05, height = 0, alpha = 0.5) +
  geom_line(data = data.frame(mean_rating = mu_grid, sd_rating = sd_max),
            aes(mean_rating, sd_rating), color = "grey40", inherit.aes = FALSE) +
  labs(x = "subject mean rating", y = "subject SD",
       title = "Group A's raw SD looks smaller -- purely because its mean sits\nnear the grey ceiling curve, not because raters agree more") +
  theme_minimal(base_size = 12)
```

Now fit both a Gaussian MELSM and the ordinal cumulative-probit MELSM to the
same data.

```{r}
#| label: fit-gaussian
#| echo: true
#| message: false
#| warning: false
#| cache: true

fit_gaussian <- brm(
  bf(rating ~ group + (1 | subject), sigma ~ group),
  data = df, family = gaussian(),
  chains = 2, iter = 2000, warmup = 1000, cores = 2, seed = 1,
  refresh = 0, silent = 2
)
```

```{r}
#| label: fit-ordinal
#| echo: true
#| message: false
#| warning: false
#| cache: true

fit_ordinal <- brm(
  bf(rating_ord ~ group + (1 | subject), disc ~ 0 + group),
  data = df, family = cumulative("probit"), init = 0,
  prior = c(
    prior(constant(0), class = "b", coef = "groupA", dpar = "disc"),
    prior(normal(0, 1), class = "b", coef = "groupB", dpar = "disc")
  ),
  chains = 2, iter = 2000, warmup = 1000, cores = 2, seed = 1,
  refresh = 0, silent = 2
)
```

```{r}
#| label: compare-results
#| echo: true
#| message: false
#| warning: false

knitr::kable(round(fixef(fit_gaussian), 2), caption = "Gaussian MELSM fixed effects (sigma part is on the log-SD scale)")
knitr::kable(round(fixef(fit_ordinal), 2), caption = "Ordinal MELSM fixed effects (disc part is on the log scale: 0 = same latent SD as the reference group)")
```

On the raw scale, Group A averages 6.64 with an SD of 0.58; Group B averages
4.89 with an SD of 1.22 -- the pooled correlation between subject-level mean
and SD is $-0.74$. That pattern looks like a real, sizable disagreement gap,
and the Gaussian MELSM agrees with it: the fitted `sigma_groupB` coefficient
of 0.68 (on the log-SD scale) implies Group B's SD is **1.97 times** Group
A's (95% CI: 1.82--2.14) -- a difference whose credible interval is nowhere
near zero.

But there is no true difference. Both groups' raters were drawn from a
latent distribution with the *same* SD ($\sigma=1$) by construction --
the only thing that differs between the groups is where their mean sits
relative to the ceiling. The ordinal cumulative-probit model's
`disc_groupB` estimate -- $\gamma_B = -0.015$ on the log scale (95% CI:
$-0.171$ to $0.146$) -- is indistinguishable from zero, i.e.
$\text{disc}_B = e^{-0.015} \approx 0.985$, a latent SD of $\approx 1.015$
against the reference group's $1$: correctly recovering the fact that the
two groups' underlying dispersion is identical. The Gaussian model's "Group B
disagrees twice as much" is not a finding about the raters; it's the
mechanical ceiling effect from the previous section, expressed as a
regression coefficient.

### What about R, on this same data?

$R$ was introduced above as mean-independent by construction -- so it should
land on the same answer as the ordinal model: no real group difference.
Compute it per subject, using the exact finite-sample bound from the earlier
callout ($n=6$ raters):

```{r}
#| label: r-measure
#| echo: true
#| message: false
#| warning: false

var_max_finite <- function(n, mu) {
  c <- (mu - 1) / 6
  a <- (n * c) %% 1
  36 * (c * (1 - c) - (1 / n) * a * (1 - a))
}

subj_R <- df |>
  group_by(subject, group) |>
  summarise(mean_rating = mean(rating),
            var_pop = mean((rating - mean(rating))^2),   # population (1/n) convention
            .groups = "drop") |>
  mutate(var_max = var_max_finite(n_raters, mean_rating),
         R = ifelse(var_max > 1e-8, var_pop / var_max, NA_real_))

knitr::kable(
  subj_R |> group_by(group) |>
    summarise(n_undefined = sum(is.na(R)), mean_R = round(mean(R, na.rm = TRUE), 3)),
  caption = "Per-subject R, using each subject's own 6 ratings for both mean and variance."
)
```

This is worse than the raw SD comparison, not better: Group A's per-subject
$R$ averages 0.63, Group B's 0.20 -- an even larger apparent gap, and in the
opposite direction from what the ceiling artifact would predict. Fourteen
Group A subjects even have an *undefined* $R$ (all six raters happened to
pick 7, so both the observed variance and the mechanical ceiling are
simultaneously zero). Has $R$ failed?

No -- the theoretical guarantee ($R\in[0,1]$ at every $\mu$, no forced
relationship to the mean) is about the *population* quantity. Plugging in a
mean estimated from only 6 raters breaks it, because the finite-sample bound
in the denominator is itself estimated from that same noisy mean, and it is
far more sensitive near the ceiling than in the middle of the scale:

```{r}
#| label: r-diagnostic
#| echo: true
#| message: false
#| warning: false

knitr::kable(
  subj_R |> group_by(group) |>
    summarise(mean_var_pop = round(mean(var_pop), 3),
              mean_var_max = round(mean(var_max, na.rm = TRUE), 3)),
  caption = "The denominator (mean_var_max) is nearly 7x smaller for Group A -- not because Group A's true ceiling is that much tighter, but because a 6-rater sample mean near 7 makes the *estimated* ceiling collapse."
)
```

A Group A subject's 6 raters average close to 7 purely because that's where
their true mean sits -- but exactly *how* close is noisy, and the
finite-sample bound shrinks fast as the sample mean approaches the boundary.
A modest, unremarkable observed variance divided by an unluckily tiny
denominator produces a large $R$, for no reason connected to real
disagreement. Swapping in the simpler continuous bound (no finite-sample
correction) shrinks the gap a great deal -- 0.181 vs. 0.173 -- but each
value is still an average of 100 individually noisy single-subject
estimates. Only pooling each group's full 600 ratings into one mean and one
variance before computing $R$ removes the small-sample noise entirely:

```{r}
#| label: r-pooled
#| echo: true
#| message: false
#| warning: false

pooled <- df |> group_by(group) |>
  summarise(mean_rating = mean(rating), var_pop = mean((rating - mean(rating))^2)) |>
  mutate(var_max_cont = (mean_rating - 1) * (7 - mean_rating),
         R_pooled = round(var_pop / var_max_cont, 3))

knitr::kable(pooled |> select(group, mean_rating, R_pooled),
             caption = "R computed once per group, from all 600 ratings pooled.")
```

Pooled, $R$ is 0.22 for Group A and 0.21 for Group B -- essentially
identical, agreeing with the ordinal model's null finding. The lesson isn't
"don't use $R$" -- it's that $R$, like $\varphi$, is only as trustworthy as
the sample size behind the mean and variance plugged into it. A hand-computed
per-subject $R$ with 6 raters can manufacture a *larger* spurious gap than
raw SD, in the opposite direction. The ordinal model never has this problem
in the first place, because it never computes a per-subject summary
statistic at all -- `disc_groupB` is estimated directly from the full,
pooled 600-observation likelihood per group, with the same effect as the
"pooled" calculation above but done coherently, and with a proper posterior
interval instead of an average of point estimates.

## Takeaways

::: {.callout-note}
## Summary

- On any bounded scale, the *maximum possible* variance is a function of the
  mean -- the Bhatia--Davis inequality, $\text{Var}(X) \leq (\mu-a)(b-\mu)$ --
  not a tendency to check for, a fact to design around.
- Dividing SD (or variance) by the mean does not remove this confound; the
  mechanism isn't proportional scaling, it's a ceiling that pinches at both
  ends.
- A Gaussian location-scale model can estimate `sigma ~ x`, but inherits the
  confound, because it has no boundary in it.
- Beta regression's precision parameter $\varphi$ (and its bounded twin $R$)
  remove the confound algebraically, at the cost of a continuous
  approximation to discrete data -- and $R$ is only as trustworthy as the
  sample size behind the numbers plugged into it. With few raters per
  subject, a hand-computed per-subject $R$ can manufacture a *larger*
  spurious group gap than raw SD, in the opposite direction.
- An ordinal cumulative-link mixed model puts the boundary where it belongs
  -- as estimated thresholds on a latent scale -- and is the cleanest way to
  ask "is this dispersion difference real?" on Likert data, because it never
  needs a per-unit summary statistic at all.
:::

## References

- Bhatia, R., and Davis, C. (2000). "A Better Bound on the Variance."
  *American Mathematical Monthly* 107(4): 353--357.

- Bürkner, P.-C., and Vuorre, M. (2019). "Ordinal Regression Models in
  Psychology: A Tutorial." *Advances in Methods and Practices in
  Psychological Science* 2(1): 77--101.

- Cribari-Neto, F., and Zeileis, A. (2010). "Beta Regression in R." *Journal
  of Statistical Software* 34(2): 1--24.

- Ellis, J. L. (2025). "The maximum variance of a finite dataset, given its
  mean, minimum, and maximum."
  [arXiv:2508.17525](https://arxiv.org/abs/2508.17525).

- Ferrari, S. L. P., and Cribari-Neto, F. (2004). "Beta Regression for
  Modelling Rates and Proportions." *Journal of Applied Statistics* 31(7):
  799--815.
