---
title: "longitudinal modified treatment policy (LMTP)"
date: "2025-08-23"
---


The longitudinal modified treatment policy (LMTP) framework, developed in epidemiology and biostatistics, handles settings where the standard ATE is poorly defined: continuous or multivalued treatments, time-varying exposures, or hypothetical interventions that would violate positivity. This chapter follows [Nicholas Williams](https://beyondtheate.com/).

Suppose we have such a DAG:


```{r}
#| label: dag1
#| include: true
#| warning: false
#| cache: false
#| message: false
#| echo: true

library(ggplot2)
library(ggdag)

g <- dagify(
  A1 ~ L1,
  L2 ~ L1,
  L2 ~ A1,
  A2 ~ A1,
  A2 ~ L2,
  Y ~ A2,
  Y ~ L2,
  Y ~ A1,
  exposure = "A2",
  outcome = "Y",
  coords = list(x = c(L1 = 1,  L2 = 2, Y = 3, A1 = 1.5, A2 = 2.5),
                y = c(L1 = 1,  L2 = 1, Y = 1, A1 = 1.5, A2 = 1.5))
)

ggdag(g) +
  theme_dag()
```

We have multiple time points, and the treatment $A$ is time-varying.  We are interested in not only the ATE of $A_2$ on $Y$, but also some other hypothetical interventions.

## Definitions and assumptions

### Assumptions

1. Positivity.  Basically if there is a unit with $a_t$ and $h_t$, then there is a unit with $d(a_t, h_t)$ and $h_t$.

2. Sequential unconfoundedness.  There is no unmeasured confounders.

### dynamic treatment regime

Some notations commonly used in this literature:  We observe $Z = (L_0, A_0, L_1, A_1, Y)$, where $L$ is the observed confounder, $A$ is the treatment, and $Y$ is the outcome.  The treatment $A$ can be time-varying.  We can also have multiple treatments at different time points.  For example, we can have $A_2$, $A_3$, etc.  In this graph, baseline $L_0$ affects $A_0$, which in turn affects $L_1$, which then affects $A_1$, and finally $A_1$ affects the outcome $Y$.

History $H_t$ is the history of data up to time $t$, right before $A_t$. for example, $H_1 = (L_0, A_0, L_1)$, and $H_2 = (L_0, A_0, L_1, A_1)$.  $d$ is the hypothetical intervention function, or shift function, which is a function of the history $H_t$ and the treatment $A_t$.  For example, $d_0(a_0,h_0,\epsilon_0)$ is a user-given function to map $a_0$, $h_0$, and $\epsilon_0$ to a potential treatment value.  The function $d$ can be deterministic, or it can be stochastic.  Then we can replace $A_0$ with $A_0^d = d_0(A_0, H_0, \epsilon_0)$.  Then after that $A_1(A_0^d)$ is called the natural value of treatment.

This is very general, comparing to the static treatment regime.


Suppose treatment $A$ is a function of the history of treatment and confounders, for example, $A = d(A_1, L_1, A_2, L_2)$.  $A$ can be set to a fixed value, say 1 or 0, or some value $A^d$.  This function $d$ can be anything, it can be taking a deterministic value, or it can be a function that takes the natural treatment value $A$ as input.  In the package "lmtp", this is called a shift function, or hypothetical intervention.  For example, $d$ can be set to 1 if $age < 30$, or $d$ can be set to double the natural value of $A$.  Many possibilities.

In comparison, for ATE, we only need to set $A$ to 0 or 1.

Under this LMTP, the causal parameter is

$$ \theta = E[Y^{\bar A^d}] $$ {#eq-lmtp-1}

$Y^{\bar A^d}$ is the potential outcome under the hypothetical intervention $\bar A^d$. At time 1, $A^d_1 = d(A_1,H_1)$.

### modified treatment policy

A simulated data set makes the estimation concrete.

This simulation is from Susmann et al. (2024) "Longitudinal Generalizations of the Average Treatment Effect on the Treated for Multi-valued and Continuous Treatments".  I modified slightly to fit the DAG above.


```{r}
#| label: sim1
#| include: true
#| warning: false
#| cache: true
#| message: false
#| echo: true

library(tidyverse)
library(tidyr)

mtp <- function(data, trt) {
  a <- data[[trt]]
  a * 0 + 1
}
simulate_data <- function(seed, N, tau, sigma = 0.5) {
  set.seed(seed)

  data <- tibble(id = 1:N)

  for(t in 1:tau) {
    Lt <- paste0("L_", t)
    Ltd <- paste0("L_", t, "d")

    At <- paste0("A_", t)
    Atd <- paste0("A_", t, "d")

    Lt1 <- paste0("L_", t - 1)
    Lt1d <- paste0("L_", t - 1, "d")

    At1 <- paste0("A_", t - 1)
    At1d <- paste0("A_", t - 1, "d")

    if(t == 1) {
      data[[Lt]] <- runif(N, 0, 1)
      data[[Ltd]] <- data[[Lt]]

      data[[At]] <- rbinom(N, size = 1, prob = 0.5)
    }
    else {
      # L_t depends on the previous treatment A_{t-1} (the DAG edge A1 -> L2),
      # which is the time-varying confounding this chapter is about.
      data[[Lt]] <- rnorm(N, mean = 0.25 * data[[Lt1]] + 0.4 * data[[At1]], 0.5)
      data[[Ltd]] <- rnorm(N, mean = 0.25 * data[[Lt1d]] + 0.4 * data[[At1d]], 0.5)

      data[[At]] <- rbinom(N, size = 1, prob = plogis(0.5 - 0.2 * data[[At1]] + 0.1 * data[[Lt1]]))
    }

    data[[Atd]] <- mtp(data, At)
  }
  # Y depends on the final treatment A_2 and confounder L_2, and also directly
  # on the first-period treatment A_1 (the DAG edge A1 -> Y).
  data$Y  <- rnorm(N, data[[At]] + data[[Lt]] + 0.5 * data[["A_1"]], sigma)
  data$Yd <- rnorm(N, data[[Atd]] + data[[Ltd]] + 0.5 * data[["A_1d"]], sigma)
  data
}

simulated_data1 <- simulate_data(seed = 123, N = 10000, tau = 2)

mean(simulated_data1$Yd)

mtp <- function(data, trt) {
  a <- data[[trt]]
  a * 0 + 0
}

simulated_data2 <- simulate_data(seed = 123, N = 10000, tau = 2)

mean(simulated_data2$Yd)

mean(simulated_data1$Yd) - mean(simulated_data2$Yd)
```

Note in this simulation the variables ending with "d" are the variables under hypothetical intervention, or modified treatment policy.  $L_1$ is from $\text{Uniform}(0,1)$, and $L_2$ is from $N(0.25 L_1 + 0.4 A_1,\ 0.5)$ -- so the time-varying confounder $L_2$ depends on the earlier treatment $A_1$, which is the time-varying confounding this chapter is about.  The treatment $A_1$ is from a Bernoulli distribution with probability 0.5, and the treatment $A_2$ is from a Bernoulli distribution with probability $\text{plogis}(0.5 - 0.2 A_1 + 0.1 L_1)$.  The outcome $Y$ is from a normal distribution with mean $A_2 + L_2 + 0.5 A_1$, so $Y$ depends on both the current treatment $A_2$ and the lagged treatment $A_1$.  In the code, the first modified treatment policy sets treatment to 1 in every period (`simulated_data1`) and the second sets it to 0 in every period (`simulated_data2`); the reported difference $E[Y^{d}_{\text{always }1}] - E[Y^{d}_{\text{always }0}]$ is the effect of the always-treat versus never-treat static regimes.


The always-treat regime gives $E[Y^d] = 2.029$, always-never gives 0.129, and
their difference is 1.900 — and that 1.9 decomposes exactly into the three
routes the DGP contains: 1 from $A_2$ direct, 0.5 from $A_1$ direct, and 0.4
through the $A_1 \to L_2 \to Y$ path, since $L_2$ shifts by 0.4 and enters $Y$
with coefficient 1.

In this case, we can just do a linear regression to get the effect of A2 on Y, knowing the exact DAG.

```{r}
#| label: sim1a
#| include: true
#| warning: false
#| cache: true
#| message: false
#| echo: true

m1 <- glm(Y ~ L_2 + A_1   + A_2, data = simulated_data1)
summary(m1)
```

Knowing the DAG, the regression recovers every structural coefficient: $L_2$ at
0.9998, $A_1$ at 0.4940 and $A_2$ at 1.0000, against true values of 1, 0.5 and 1.
Note this identifies the *direct* effect of $A_1$ only; its path through $L_2$ is
blocked by conditioning on $L_2$, which is why the total of 1.9 above cannot be
read off this table.

Now consider a different MTP: set half of the time to 0 and leave the other half unchanged.
$$
d(a_t, \epsilon_t) = \begin{cases}
    0, & \text{if } \epsilon_t < .5 \ and \ a_t =1 \\
    a_t, & \text{otherwise}
\end{cases}
$$ {#eq-lmtp-2}

This is, say, to set half of smokers to non-smokers, and the other half remain smokers.


```{r}
#| label: sim2
#| include: true
#| warning: false
#| cache: true
#| message: false
#| echo: true

# mtp <- function(data, trt) {
#   a <- data[[trt]]
#   epsilon <- rbinom(nrow(data), size = 1, prob = 0.5)
#   ifelse(epsilon <.5 & a == 1, 0, a)
# }
d <- function(a) {
  epsilon <- runif(length(a))
  ifelse(epsilon < 0.5 & a == 1, 0, a)
}
simulated_data1$m3_d <- simulated_data1$Y

m2 <- glm(m3_d ~ L_1 + A_1 + L_2 + A_2, data = simulated_data1)
summary(m2)
simulated_data1$m2_d <- predict(m2, mutate(simulated_data1, A_2 = d(A_2)))

m1 <- glm(m2_d ~ L_1 + A_1, data = simulated_data1)
summary(m1)

m1_d <- predict(m1, mutate(simulated_data1, A_1 = d(A_1)))
mean(m1_d)
```

The recursion makes the blocked path visible. The first regression, on the full
history, again gives $A_1 = 0.4945$ — the direct effect. The second, after
substituting the shifted $A_2$ and regressing on $L_1$ and $A_1$ only, gives
$A_1 = 0.8760$, which is close to the total $0.5 + 0.4 = 0.9$: dropping $L_2$
from the conditioning set lets $A_1$'s indirect path back in. The final average
is 0.652, the expected outcome under the stochastic half-off regime, sitting
between the always-treat 2.029 and never-treat 0.129 as it should.

Note this recursive process is based on Diaz, et al. (2023) "Nonparametric Causal Effects Based on Longitudinal Modified Treatment Policies".

Here is how exactly we estimate it:  we start with the last time point.  Regress it on previous treatment and confounders, and then get the predicted value with $A$ changed based on the MTP.  Regress that predicted value on the previous treatment and confounders, get predicted values with $A$ changed based on MTP.  Repeat until the first time point.   The average of the predicted value at time 1 is the expected value under this MTP.

The shift function `d()` above is a *stochastic* intervention: at each
call it draws a fresh $\epsilon \sim \text{Uniform}(0,1)$ and, with
probability 0.5, switches a treated unit ($a=1$) to $0$, otherwise
leaves $a$ unchanged. The regime is "at each time point, draw
independently conditional on history and apply this rule"; because the
draw is independent across time points and units, calling `d()`
separately for $A_2$ and $A_1$ is what implements that regime. (By
contrast, the `d1`/`d2` functions used with `lmtp_tmle` below are
*deterministic* static regimes -- always 1 and always 0.)

This is basically g-formula extended to longitudinal data.

In a general case, this is the generalized g-formula to estimate $\theta$:

Set $m_{\tau + 1} = Y$, let $A_t^d = d(A_t, H_t)$.   For $t = \tau, \ldots, 1$, recursively define:

$$ m_t(a_t, h_t)  = E[m_{t + 1} (A_{t+1}^d, H_{t+1}) | A_t = a_t, H_t =h_t] $$ {#eq-lmtp-3}

Then $\theta = E[m_1(A_1^d, L_1)]$.

We start from the last period.  Regress $m_{\tau + 1}$, which is $Y$ on $A_{\tau}$ and $H_{\tau}$.  Then get the predicted value with $A_{\tau}$ changed based on the MTP.  Then regress that predicted value on $A_{\tau - 1}$ and $H_{\tau - 1}$.  Repeat until the first time point.  The average of the predicted value at time 1 is the expected value under this MTP.

### Estimators

The authors advocate two estimators, TMLE and SDR (sequentially doubly robust estimator).  The procedures are the same, starting from the last time point, then apply TMLE or SDR, iterate to the first time point.


## Using lmtp package

```{r}
#| label: sim3
#| include: true
#| warning: false
#| cache: true
#| message: false
#| echo: true

library(tidyverse)
library(lmtp)
d1 <- function(data, trt) {
  rep(1, nrow(data))
}
A <- "A_2"
Y <- "Y"
W <- c("L_1", "L_2", "A_1")

set.seed(34465)

treat <- lmtp_tmle(
  data = simulated_data1,
  trt = "A_2",
  outcome = "Y",
  baseline = W,
  outcome_type = "continuous",
  shift = d1,
  folds = 1,
  learners_trt = "SL.glm",
  learners_outcome = "SL.glm"
)

print(treat)
```

```{r}
#| label: sim4
#| include: true
#| warning: false
#| cache: true
#| message: false
#| echo: true

d2 <- function(data, trt) {
  rep(0, nrow(data))
}
A <- "A_2"
Y <- "Y"
W <- c("L_1", "L_2", "A_1")

set.seed(34465)

control <- lmtp_tmle(
  data = simulated_data1,
  trt = "A_2",
  outcome = "Y",
  baseline = W,
  outcome_type = "continuous",
  shift = d2,
  folds = 1,
  learners_trt = "SL.glm",
  learners_outcome = "SL.glm"
)

print(control)

lmtp_contrast(treat, ref = control)
```

Note that `folds = 1` turns off cross-fitting (sample splitting). That is
fine for a small, fast demonstration with `SL.glm`, but in applied work
with flexible learners you should use several folds so the nuisance fits
are cross-fitted; otherwise the standard errors can be anti-conservative.

**Note on the estimand:** this `lmtp_tmle` example only shifts $A_2$ (`trt =
"A_2"`), treating $A_1$ as an observed baseline covariate (via `baseline =
W`) rather than also intervening on it. That is a different, single-period
policy from the always-1-vs-always-0 *both-period* regime computed manually
via `simulated_data1$Yd`/`simulated_data2$Yd` above (lines 139-152) — the two
numbers are not expected to match, and should not be used as a sanity check
against each other. To replicate the both-period always-treat/never-treat
contrast with `lmtp`, shift both time points by passing `trt = c("A_1",
"A_2")` and supplying `time_vary` (the time-varying covariates measured
before each treatment time, e.g. `list(character(0), "L_2")`), with `shift`
functions that map each treatment in the vector to 1 (or 0).

## another example

The data set bmi, from the DynTxRegime package, are simulated to reflect a two-stage RCT {A1, A2} that studied the effect of meal replacement (MR) shakes versus a calorie deficit (CD) diet on adolescent

### shift function 1

Consider a shift function that assigns meal replacement to all observations at time 1, but only meal replacement at time 2 to those observations whose 4-month BMI is greater than 30.


```{r}
#| label: sim5
#| include: true
#| warning: false
#| cache: true
#| message: false
#| echo: true

library(DynTxRegime)
data(bmiData)
bmi <- bmiData
glimpse(bmi)

d_dtr <- function(data, trt) {
  if (trt == "A1") return(rep("MR", nrow(data)))

  ifelse(data$month4BMI > 30, "MR", "CD")
}

fit_dtr <- lmtp_sdr(
  data = bmi,
  trt = c("A1", "A2"),
  outcome = "month12BMI",
  baseline = c("gender", "race", "parentBMI"),
  time_vary = list("baselineBMI", "month4BMI"),
  shift = d_dtr,
  outcome_type = "continuous",
  folds = 1,
  learners_trt = "SL.glm",
  learners_outcome = c("SL.mean", "SL.glm", "SL.gam")
)

fit_dtr
```

### shift function 2

Suppose we are interested in comparing the dynamic treatment regime to a static treatment regime where all patients receive meal replacement at both time points. Using the SDR estimator, estimate the effect of this static intervention.



```{r}
#| label: sim6
#| include: true
#| warning: false
#| cache: true
#| message: false
#| echo: true

fit_MR <- lmtp_sdr(
  data = bmi,
  trt = c("A1", "A2"),
  outcome = "month12BMI",
  baseline = c("gender", "race", "parentBMI"),
  time_vary = list("baselineBMI", "month4BMI"),
  shift = \(data, trt) rep("MR", nrow(data)),
  outcome_type = "continuous",
  folds = 1,
  learners_trt = "SL.glm",
  learners_outcome = c("SL.mean", "SL.glm", "SL.gam")
)

fit_MR
```
Let’s also estimate the effect of an intervention where all patients receive a calorie deficit diet at both time points.

```{r}
#| label: sim7
#| include: true
#| warning: false
#| cache: true
#| message: false
#| echo: true

fit_CD <- lmtp_sdr(
  data = bmi,
  trt = c("A1", "A2"),
  outcome = "month12BMI",
  baseline = c("gender", "race", "parentBMI"),
  time_vary = list("baselineBMI", "month4BMI"),
  shift = \(data, trt) rep("CD", nrow(data)),
  outcome_type = "continuous",
  folds = 1,
  learners_trt = "SL.glm",
  learners_outcome = c("SL.mean", "SL.glm", "SL.gam")
)

fit_CD
```

Finally, we can compare the three treatment regimes using the `lmtp_contrast` function.

```{r}
#| label: sim8
#| include: true
#| warning: false
#| cache: true
#| message: false
#| echo: true

lmtp_contrast(fit_dtr, fit_MR, ref = fit_CD)

```

---

<!-- see-also-footer -->
*Systematic treatment: [R](https://xiangao.github.io/causal_econometrics_guide/g-methods.html) · [Julia](https://xiangao.github.io/causal_econometrics_julia/g-methods.html).*
