---
title: "Marginal effects in models with fixed effects"
date: "2019-01-25"
---


## Marginal effects in a linear model

Stata's `margins` command computes predicted means and marginal effects, but care is needed when fixed effects are present. In a linear model the issue is benign: demeaning and dummy-variable approaches return the same coefficients and the same marginal effects.

We check that on the `auto` data, treating the 5-category repair record `rep78`
as the panel identifier and regressing `price` on `mpg` interacted with `trunk`
— once with `xtreg, fe` (within-demeaning) and once with `reg` plus `i.rep78`
dummies.


```{r}
#| label: stata-chunk
#| engine: 'stata'
#| engine.path: '/usr/local/bin/stata'
#| cache: true

clear
sysuse auto
xtset rep78
xtreg price c.mpg##c.trunk, fe
margins , dydx(mpg)
reg price c.mpg##c.trunk i.rep78
margins , dydx(mpg)

```

The two agree exactly. Both give $-98.12003$ on `mpg`, 295.0544 on `trunk` and
$-12.23318$ on the interaction, and both average marginal effects are
$-268.4981$ with a standard error of 74.12513. The only difference is that
`xtreg` reports a $z$ statistic and `reg` a $t$, so the confidence intervals
differ in the fourth digit. Nothing here needs care.

## Marginal effects in a non-linear model

In a nonlinear model the two approaches diverge. We now need a genuine panel and
a count outcome, so we simulate one: 250 observations, 50 units observed over 5
periods. Within each unit, $\text{mpg} \sim N(20, 4^2)$ and
$\text{trunk} \sim N(15, 3^2)$ vary period to period, while the unit effect
$\alpha_i \sim N(0,1)$ is drawn once and held fixed. The conditional mean is
$\lambda_{it} = \exp(0.5 + 0.05\,\text{mpg}_{it} - 0.03\,\text{trunk}_{it} + \alpha_i)$
and $\text{accidents}_{it} \sim \text{Poisson}(\lambda_{it})$. The true
coefficients are 0.05 and $-0.03$.

```{r}
#| label: stata-chunk2
#| engine: 'stata'
#| engine.path: '/usr/local/bin/stata'
#| cache: true

clear
set seed 42
set obs 250
gen id = ceil(_n/5)
bys id: gen t = _n
gen mpg = rnormal(20,4)
gen trunk = rnormal(15,3)
gen alpha = rnormal(0,1)
bys id: replace alpha = alpha[1]
gen lambda = exp(0.5 + 0.05*mpg - 0.03*trunk + alpha)
gen accidents = rpoisson(lambda)
xtset id
xtpoisson accidents mpg trunk, fe
margins , dydx(mpg)
margins , dydx(mpg) predict(nu0)
poisson accidents mpg trunk i.id
margins , dydx(mpg)

```

(We use a simulated count outcome `accidents` and a proper panel with 50 units of 5 periods each, rather than `auto`'s continuous `price` and its 5-category `rep78` — a true panel needs many units, and a fixed-effect count model needs an actual count dependent variable.)

In this example, "xtpoisson, fe" and "poisson ... i.id" return the
same coefficients (both give `mpg` = .0487946 with a standard error of .0090524).  Fixed effect Poisson model (sometimes called
conditional fixed effect Poisson) is the same models as a Poisson
model with dummies, just like a linear model (OLS with dummies is the
same as fixed effect OLS). Poisson model and OLS are unique in this
sense that there is no "incidental parameter" problem.

Both recover the truth: 0.0488 against a true 0.05 for `mpg`, and $-0.0316$
against $-0.03$ for `trunk`. Note the sample sizes differ, 245 against 250 —
`xtpoisson, fe` drops the one group whose outcome is zero in every period,
because such a group contributes nothing to the conditional likelihood. It
changes neither coefficient here.

The `margins` calls return different marginal effects even though the models are the same, and the spread is not small:

| call | AME of `mpg` | std. err. |
|---|---|---|
| `margins, dydx(mpg)` after `xtpoisson, fe` | 0.0488 | 0.0091 |
| `margins, dydx(mpg) predict(nu0)` | 0.0850 | 0.0341 |
| `margins, dydx(mpg)` after `poisson ... i.id` | 0.2239 | 0.0421 |

Three answers from one coefficient, the largest 4.6 times the smallest. In the conditional fixed-effect Poisson the fixed effects are conditioned out — they are not estimated. The first row is not really a marginal effect at all: with no $\alpha_i$ to exponentiate against, it returns the coefficient itself. `margins, predict(nu0)` sets the fixed effect to zero, which is arbitrary, and note that the conditional model has no intercept either — it was swept out with the $\alpha_i$. So `nu0` exponentiates $X\beta$ alone, giving a baseline count no unit in the data actually has. The Poisson with dummies estimates the fixed effects, so `margins` uses actual estimates and gets the largest value.

The arithmetic is easy to follow, since for a Poisson $\partial\lambda/\partial x = \beta\lambda$ and therefore every one of these AMEs is $\beta$ times an average $\lambda$. Divide each row by 0.0488 and the implied baselines are 1.00, 1.74 and 4.59. The last matches $E[\lambda] = e^{1.05}E[e^{\alpha}] = 4.71$ from the DGP; the middle matches $e^{X\beta}$ evaluated without intercept or fixed effect, 1.65. The three rows differ only in what they assume about $\alpha_i$, and none of them is a modelling subtlety — it is a choice of baseline. The third is the defensible one, but for the conditional model the safest route is to report the original coefficients and avoid `margins` for marginal effects entirely.

The conditional logit is worse. The fixed effects are not estimated, and unlike Poisson there is an incidental-parameter problem, so logit with dummies is inconsistent unless the panel is deep (roughly 20+ observations per unit). With conditional logit the predicted probability depends on the unestimated $\alpha_i$; `margins` after `clogit` or `xtlogit, fe` offers options like `pu0` (set all fixed effects to zero), but none is defensible.

In a fixed effect logit model,

$$ log(P(y=1)/(1-P(y=1))) = \alpha_i +  \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1*x_2 $$ {#eq-marginal-effects-fe-1}

Here $\alpha_i$ is the unit fixed effect. Therefore

$$ P(y=1) = F(\alpha_i +  \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1*x_2) $$ {#eq-marginal-effects-fe-2}

Without estimating $\alpha_i$ there is no way to predict $P$ in a meaningful way. But the log-odds are linear in the covariates, and the marginal effect of $x_1$ or $x_2$ on the log-odds does not involve $\alpha_i$. So `margins` with the `predict(xb)` option — marginal effects on the log-odds scale — is interpretable.  Note that `xb` here
is $X_{it}\beta$ only — the individual fixed effect $\alpha_i$ is not
estimated in conditional logit and so is not part of the prediction.
The logged-odds *effect* of a covariate ($\partial(X\beta)/\partial x$)
therefore matches the original coefficient exactly and does not depend
on $\alpha_i$, but `xb` itself is a log-odds relative to each
individual's (unidentified) baseline, not the individual's absolute
log-odds. This lets you make linear extrapolations of relative
log-odds across values of the covariates, but not of absolute
predicted probabilities, since those would require $\alpha_i$.

```{r}
#| label: stata-chunk3
#| engine: 'stata'
#| engine.path: '/usr/local/bin/stata'
#| cache: true

clear
webuse union
clogit union c.age##i.south not_smsa grade, group(idcode)
margins, at( age=(15 20 25 30 35 40) south=(0 1)) predict(xb)
marginsplot
```



This fits a conditional logit of union status on age, south, and their interaction, on 12,035 person-years from the `union` panel. The age coefficient is 0.0097 with a standard error of 0.0050, and the south indicator is $-1.382$ with a standard error of 0.277. `margins` with `predict(xb)` then returns predicted log-odds at six ages crossed with south — twelve combinations — but not predicted probabilities, since those require $\alpha_i$.

Read the vertical axis of that plot as a *relative* scale.  Because $\alpha_i$ is
not estimated, `xb` is each individual's log-odds measured from an unidentified
baseline, so the level of any single point is arbitrary; what is interpretable is
how the curves move across age and how far apart the south and non-south curves
sit.  Differences within the plot are the estimable quantities, not heights.
