---
title: "Marginal effects in models with fixed effects"
date: "2019-01-25"
---


## Marginal effects in a linear model

Stata's `margins` command computes predicted means and marginal effects, but care is needed when fixed effects are present. In a linear model the issue is benign: demeaning and dummy-variable approaches return the same coefficients and the same marginal effects.


```{r}
#| label: stata-chunk
#| engine: 'stata'
#| engine.path: '/usr/local/bin/stata'
#| cache: true

clear
sysuse auto
xtset rep78
xtreg price c.mpg##c.trunk, fe
margins , dydx(mpg)
reg price c.mpg##c.trunk i.rep78
margins , dydx(mpg)

```

## Marginal effects in a non-linear model

In a nonlinear model the two approaches diverge:

```{r}
#| label: stata-chunk2
#| engine: 'stata'
#| engine.path: '/usr/local/bin/stata'
#| cache: true

clear
set seed 42
set obs 250
gen id = ceil(_n/5)
bys id: gen t = _n
gen mpg = rnormal(20,4)
gen trunk = rnormal(15,3)
gen alpha = rnormal(0,1)
bys id: replace alpha = alpha[1]
gen lambda = exp(0.5 + 0.05*mpg - 0.03*trunk + alpha)
gen accidents = rpoisson(lambda)
xtset id
xtpoisson accidents mpg trunk, fe
margins , dydx(mpg)
margins , dydx(mpg) predict(nu0)
poisson accidents mpg trunk i.id
margins , dydx(mpg)

```

(We use a simulated count outcome `accidents` and a proper panel with 50 units of 5 periods each, rather than `auto`'s continuous `price` and its 5-category `rep78` — a true panel needs many units, and a fixed-effect count model needs an actual count dependent variable.)

In this example, "xtpoisson, fe" and "poisson ... i.id" return the
same coefficients (both give `mpg` = .0487946 with a standard error of .0090524).  Fixed effect Poisson model (sometimes called
conditional fixed effect Poisson) is the same models as a Poisson
model with dummies, just like a linear model (OLS with dummies is the
same as fixed effect OLS). Poisson model and OLS are unique in this
sense that there is no "incidental parameter" problem.

The `margins` calls return different marginal effects even though the models are the same. In the conditional fixed-effect Poisson the fixed effects are conditioned out — they are not estimated. `margins, predict(nu0)` sets them to zero, which is arbitrary. The Poisson with dummies estimates them, so `margins` uses actual estimates. The second is more defensible, but for the conditional model the safest route is to report the original coefficients and avoid `margins` for marginal effects entirely.

The conditional logit is worse. The fixed effects are not estimated, and unlike Poisson there is an incidental-parameter problem, so logit with dummies is inconsistent unless the panel is deep (roughly 20+ observations per unit). With conditional logit the predicted probability depends on the unestimated $\alpha_i$; `margins` after `clogit` or `xtlogit, fe` offers options like `pu0` (set all fixed effects to zero), but none is defensible.

In a fixed effect logit model,

$$ log(P(y=1)/(1-P(y=1))) = \alpha_i +  \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1*x_2 $$ {#eq-marginal-effects-fe-1}

Here $\alpha_i$ is the unit fixed effect. Therefore

$$ P(y=1) = F(\alpha_i +  \beta_1 x_1 + \beta_2 x_2 + \beta_{12} x_1*x_2) $$ {#eq-marginal-effects-fe-2}

Without estimating $\alpha_i$ there is no way to predict $P$ in a meaningful way. But the log-odds are linear in the covariates, and the marginal effect of $x_1$ or $x_2$ on the log-odds does not involve $\alpha_i$. So `margins` with the `predict(xb)` option — marginal effects on the log-odds scale — is interpretable.  Note that `xb` here
is $X_{it}\beta$ only — the individual fixed effect $\alpha_i$ is not
estimated in conditional logit and so is not part of the prediction.
The logged-odds *effect* of a covariate ($\partial(X\beta)/\partial x$)
therefore matches the original coefficient exactly and does not depend
on $\alpha_i$, but `xb` itself is a log-odds relative to each
individual's (unidentified) baseline, not the individual's absolute
log-odds. This lets you make linear extrapolations of relative
log-odds across values of the covariates, but not of absolute
predicted probabilities, since those would require $\alpha_i$.

```{r}
#| label: stata-chunk3
#| engine: 'stata'
#| engine.path: '/usr/local/bin/stata'
#| cache: true

clear
webuse union
clogit union c.age##i.south not_smsa grade, group(idcode)
margins, at( age=(15 20 25 30 35 40) south=(0 1)) predict(xb)
marginsplot
```



This fits a conditional logit of union status on age, south, and their interaction. `margins` with `predict(xb)` returns predicted log-odds for different age-by-south combinations — but not predicted probabilities, since those require $\alpha_i$.

Read the vertical axis of that plot as a *relative* scale.  Because $\alpha_i$ is
not estimated, `xb` is each individual's log-odds measured from an unidentified
baseline, so the level of any single point is arbitrary; what is interpretable is
how the curves move across age and how far apart the south and non-south curves
sit.  Differences within the plot are the estimable quantities, not heights.
