---
title: "Mediation analysis in R and Stata"
date: "2019-08-06"
---


## Mediation analysis

The traditional mediation model has two equations:

$$ Y = a X + b M + \epsilon_1 $$ {#eq-mediation-analysis-1}
$$ M = c X + \epsilon_2 $$ {#eq-mediation-analysis-2}

The effect of $X$ on $Y$ decomposes into a direct effect and an indirect effect through $M$. The classical approach is [Baron and Kenny's](http://davidakenny.net/cm/mediate.htm) four-step procedure. Modern implementations use SEM to estimate both equations jointly, as in R's `lavaan` and Stata's `sem`.

```{r}
#| echo: true
#| message: false

library(lavaan)
set.seed(1234)
n <- 10000
X <- rnorm(n)
M <- 0.5*X + rnorm(n)
Y <- 0.7*M + 0.3*X + rnorm(n)
Data <- data.frame(X = X, Y = Y, M = M)
model <- '   Y ~ a*X + b*M
             M ~ c*X
           # direct effect (a)
           # indirect effect (b*c)
             bc := b*c
           # total effect
             total := a + (b*c)
         '
fit <- sem(model, data = Data)
summary(fit)

```
We simulate $n = 10{,}000$ observations with $X \sim N(0,1)$,
$M = 0.5X + \varepsilon_2$ and $Y = 0.7M + 0.3X + \varepsilon_1$, both errors
standard normal. So the true direct effect is $a = 0.3$, the mediator
coefficient is $b = 0.7$, the treatment-to-mediator path is $c = 0.5$, the
indirect effect is $bc = 0.35$ and the total is 0.65. There is no confounding of
any kind.

`lavaan` recovers all of it: $a = 0.286$, $b = 0.699$, $c = 0.511$, giving
$bc = 0.357$ and a total of 0.643. The model is saturated (0 degrees of
freedom), so the perfect fit statistic carries no information.

The same analysis in Stata:

```{r}
#| echo: false
#| message: false

library(Statamarkdown)
```
```{stata}
*| cache: true

set obs 10000
gen x= rnormal()
gen m= .5*x + rnormal()
gen y= .7*m + .3*x +rnormal()
sem (y <- x m) (m <- x)
estat teffects

```

The above examples should have direct effect of .3 and indirect effect of .35, and total effect of .65.

## Causal Mediation analysis

The traditional mediation analysis has been criticized for the lack of causal interpretation.  Without manipulation of the mediator, it is hard to interpret the effects causally, because even if the treatment is from random experiments, the mediator is often not.  Therefore there could be an unmeasured confounder that is causing both $M$ and $Y$.

R's "mediation" package is for causal mediation analysis.  It uses simulation to estimate the causal effects of treatment, under assumptions of sequential ignorability.

::: {.callout-note}
## Identifying assumptions for natural (in)direct effects

Before reading the `mediate()` output causally, the following must hold (this is "sequential ignorability" plus the standard support and link conditions):

1. **Treatment ignorability** — no unmeasured treatment–outcome confounding given covariates $W$.
2. **Mediator ignorability** — no unmeasured mediator–outcome confounding given $W$ and treatment.
3. **No treatment-induced confounding** — no variable affected by the treatment confounds the mediator–outcome relationship (this is the strong, often-violated condition).
4. **Positivity** — both treatment and mediator have positive probability across $W$.
5. **Consistency** — observed outcomes equal the corresponding potential outcomes.
6. **Cross-world independence** — the condition underlying natural effects, which no single experiment can test.

Mediator ignorability and the no-treatment-induced-confounding condition are the ones that usually break, because the mediator is rarely randomized.
:::

It estimates the following quantities:

$$\tau_i = Y_i(1, M_i(1)) - Y_i(0, M_i(0))$$ {#eq-mediation-analysis-3}

This is the total treatment effect, which is to say, what's the change in $Y$ if we change each unit from control to treated, hypothetically?

Then this can be decomposed into the causal mediation effects:

$$\delta_i(t) = Y_i(t, M_i(1)) - Y_i(t, M_i(0))$$ {#eq-mediation-analysis-4}

for treatment status $t=0,1$.  This is to say, given treatment status, what's the mediation effect?

$$\eta_i(t) = Y_i(1, M_i(t)) - Y_i(0, M_i(t)) $$ {#eq-mediation-analysis-5}

for treatment status $t=0,1$.  This is to say, given mediator status for each treatment status, what's the direct effect?

Therefore there are four quantities estimated, direct and mediation effect for treated and control.

R's "mediation" needs users to feed two models, outcome model and mediation model.

If we study the same data, we would expect it returns the same estimates as the traditional methods.  However, the causal mediation models can be much more flexible in outcome and mediation models.

```{r}
#| echo: true
#| message: false

library(mediation)
med.fit <- lm(M ~ X, data=Data)
summary(med.fit)
out.fit <- lm(Y ~ X + M,data=Data)
summary(out.fit)
set.seed(123)
med.out <- mediate(med.fit, out.fit, treat = "X", mediator = "M", sims = 100)
summary(med.out)

```

And it does. The two component regressions give $c = 0.510954$ and, in the
outcome equation, $a = 0.285582$ and $b = 0.699006$ — the same numbers `lavaan`
produced. `mediate()` then reports ACME 0.35684 (against a true 0.35), ADE
0.28540 (against 0.30) and a total effect of 0.64224 (against 0.65), with the
proportion mediated at 0.555. The simulation-based intervals all exclude zero.

The point is that on this data the causal machinery adds nothing — which is the
right outcome for a DGP with no confounding, a linear outcome model and no
treatment-mediator interaction. Its value appears when those conditions fail.

## Binary outcome

For example, in the case of binary outcome, the traditional approach will have difficulties.  We can estimate the outcome model and mediator model jointly, but the total effects are not easy to decompose into direct and indirect effect (see Imai et al, page 320 https://imai.fas.harvard.edu/research/files/BaronKenny.pdf).

The causal mediation analysis framework is much more general.


```{r}
#| echo: true
#| message: false

library(mediation)
data(framing)
med.fit <- lm(emo ~ treat + age + educ + gender + income, data = framing)
out.fit <- glm(cong_mesg ~ emo + treat + age + educ + gender + income,
               data = framing, family = binomial("probit"))
set.seed(123)
med.out <- mediate(med.fit, out.fit, treat = "treat", mediator = "emo",
                     robustSE = TRUE, sims = 100)
summary(med.out)

```

Here the decomposition does something the linear example could not. The
`framing` data has a binary outcome (`cong_mesg`, whether the subject sent a
message to Congress) modelled by probit, with anxiety (`emo`) as the mediator.
ACME comes out at 0.0813 under control and 0.0824 under treatment, both
significant, while the direct effects are 0.0183 and 0.0194 with $p = 0.72$ and
the total effect is 0.1007 with $p = 0.24$.

Read that pattern carefully: the *total* effect is not distinguishable from
zero, and neither is the direct effect, yet the mediated effect clearly is. A
study reporting only the total effect would have concluded the framing treatment
did nothing. The proportion mediated is reported as 0.63, but note its interval
runs from $-13.7$ to $4.3$ — a ratio whose denominator is indistinguishable from
zero is not a usable quantity, and this is the standard reason not to lean on
"proportion mediated".

"mediation" package has more functionalities, such as multilevel, interaction of treatment and mediator, etc.

Stata's sem and gsem commands can model different situations, but the direct effect and indirect effects are not easy to compute, especially when you have binary outcome, or other non-continuous outcome situations.  They are not designed for causal mediation analysis.

## Going further: formal causal mediation

The `mediation` package above implements sequential ignorability — a specific
set of identifying assumptions. The [causal mediation chapter](causal-mediation.qmd)
covers the assumptions more formally, introduces the controlled direct effect
(CDE) and natural direct/indirect effects (NDE/NIE) using potential-outcome
notation, and discusses identification with unmeasured mediator–outcome
confounding. If your setting involves a non-binary outcome, treatment–mediator
interaction, or you need a doubly-robust estimator, start there.

---

<!-- see-also-footer -->
*Systematic treatment: [R](https://xiangao.github.io/causal_econometrics_guide/mediation.html) · [Julia](https://xiangao.github.io/causal_econometrics_julia/mediation.html).*
