---
title: "Correlated Random Effects (CRE) in Linear and Nonlinear Models"
author: "Xiang Ao"
date: "2025-11-11"
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(fixest)
library(bacondecomp)
library(lme4)
library(marginaleffects)
```

## Background: Unobserved Heterogeneity and CRE

In panel data models with unobserved individual heterogeneity $v_i$:

$$y_{it} = \alpha + X_{it}\beta + v_i + \epsilon_{it}$$

Standard Random Effects (RE) assumes that $v_i$ is independent of the covariates ($E[v_i \mid X_i] = 0$). If $v_i$ correlates with $X_{it}$, RE estimates are biased and inconsistent. Conversely, Fixed Effects (FE) conditions out $v_i$ via demeaning or first-differencing, but cannot estimate coefficients on time-invariant covariates and does not easily extend to nonlinear models.

The **Correlated Random Effects (CRE)** framework (Mundlak, 1978; Chamberlain, 1982; Wooldridge, 2010) parameterizes the dependence between $v_i$ and the history of $X_i$:

$$v_i = \psi + \bar{X}_i \xi + a_i, \qquad E[a_i \mid X_i] = 0$$

where $\bar{X}_i = \frac{1}{T}\sum_{t=1}^T X_{it}$ is the time-average of time-varying covariates for unit $i$. Substituting this projection into the model yields:

$$y_{it} = (\alpha + \psi) + X_{it}\beta + \bar{X}_i \xi + a_i + \epsilon_{it}$$

Adding unit means $\bar{X}_i$ directly controls for the correlation between $X_{it}$ and the unobserved effect $v_i$.

## Linear Models: Equivalence to Fixed Effects

In a balanced linear panel, estimating the model with group means $\bar{X}_i$ via Pooled OLS (POLS) or Random Effects (GLS) produces point estimates on $X_{it}$ that are **algebraically identical** to the Within (FE) estimator.

### Empirical Demonstration: `castle` Data

Using the balanced `castle` panel (50 states over 11 years; $N = 550$):
- Outcome: log homicide rate (`l_homicide`)
- Covariates: poverty rate (`poverty`) and log police employment (`l_police`)

```{r}
#| label: fix1
#| message: false
#| warning: false

data("castle")

# 1. Two-way Fixed Effects
fe_model <- feols(l_homicide ~ poverty + l_police | state + year, data = castle)

# 2. Standard Random Effects (assumes Cov(X, v) = 0)
re_model <- lmer(l_homicide ~ poverty + l_police + factor(year) + (1 | state), data = castle)

# 3. CRE Pooled OLS (Mundlak device)
castle2 <- castle |>
  group_by(state) |>
  mutate(poverty_mean = mean(poverty, na.rm = TRUE),
         l_police_mean = mean(l_police, na.rm = TRUE))

cre_pols <- feols(l_homicide ~ poverty + poverty_mean + l_police + l_police_mean + factor(year), 
                  data = castle2, cluster = ~state)

# 4. CRE Random Effects
cre_re <- lmer(l_homicide ~ poverty + poverty_mean + l_police + l_police_mean + factor(year) + (1 | state), 
               data = castle2)

# Compare estimates on poverty and l_police
res_table <- data.frame(
  Estimator = c("Two-way Fixed Effects", "Standard Random Effects", "CRE Pooled OLS", "CRE Random Effects"),
  beta_poverty = c(coef(fe_model)["poverty"], fixef(re_model)["poverty"], 
                   coef(cre_pols)["poverty"], fixef(cre_re)["poverty"]),
  beta_police  = c(coef(fe_model)["l_police"], fixef(re_model)["l_police"], 
                   coef(cre_pols)["l_police"], fixef(cre_re)["l_police"])
)
res_table[, 2:3] <- round(res_table[, 2:3], 6)
knitr::kable(res_table, caption = "Comparison of linear panel estimators.")
```

### Key Takeaways in the Linear Case

1. **Numerical Equivalence:** The Two-way FE, CRE Pooled OLS, and CRE RE models yield identical coefficients: $\beta_{\text{poverty}} = -0.027068$ and $\beta_{\text{police}} = 0.066033$.
2. **Failure of Standard RE:** Without the group means, standard RE yields $\beta_{\text{poverty}} = -0.000799$, severely attenuated toward zero due to correlation with unobserved state effects.
3. **Inference:** Because POLS ignores within-unit error correlation, cluster-robust standard errors clustered at the unit level (`cluster = ~state`) are required to obtain valid standard errors matching FE.

## Nonlinear Panel Models: Binary Outcomes

For a binary outcome $y_{it} \in \{0, 1\}$, the latent model is:

$$P(y_{it} = 1 \mid X_{it}, v_i) = \Phi(X_{it}\beta + v_i)$$

In nonlinear models, the algebraic equivalence between FE and CRE breaks down:
1. **Limitations of Conditional (Fixed Effects) Logit:**
   - Requires strict conditional independence of $y_{it}$ across time given $X_i$ and $v_i$. Any serial correlation in the idiosyncratic errors causes inconsistency.
   - Because $v_i$ is conditioned out rather than estimated, partial effects (Average Marginal Effects) cannot be computed.
2. **Advantages of CRE Pooled Probit:**
   - Incorporates the Mundlak device by including $\bar{X}_i$ in a pooled probit.
   - Robust to arbitrary serial correlation when paired with cluster-robust standard errors (Lin and Wooldridge, 2019).
   - Allows straightforward computation of Average Marginal Effects (AMEs).

### Empirical Demonstration: Binary Outcome

We define a binary outcome `high_homicide` indicating homicide rates above the sample mean:

```{r}
#| label: fix2
#| message: false
#| warning: false

castle2 <- castle2 |>
  mutate(high_homicide = ifelse(l_homicide > mean(l_homicide, na.rm = TRUE), 1, 0))

# 1. CRE Pooled Probit
cre_probit <- glm(high_homicide ~ poverty + poverty_mean + l_police + l_police_mean + factor(year), 
                  data = castle2, family = binomial(link = "probit"))

# 2. Linear Probability Model with CRE
cre_lpm <- lm(high_homicide ~ poverty + poverty_mean + l_police + l_police_mean + factor(year), 
              data = castle2)

# 3. CRE Random Effects Probit
cre_re_probit <- glmer(high_homicide ~ poverty + poverty_mean + l_police + l_police_mean + factor(year) + (1 | state), 
                       data = castle2, family = binomial(link = "probit"))

# Compute Average Marginal Effects (AMEs)
ame_probit <- avg_slopes(cre_probit, variables = c("poverty", "l_police"))
ame_lpm    <- avg_slopes(cre_lpm, variables = c("poverty", "l_police"))
ame_re     <- avg_slopes(cre_re_probit, variables = c("poverty", "l_police"))

ame_comp <- data.frame(
  Model = c("CRE Pooled Probit", "CRE Linear Probability Model", "CRE Random Effects Probit"),
  AME_poverty = c(ame_probit$estimate[ame_probit$term == "poverty"],
                  ame_lpm$estimate[ame_lpm$term == "poverty"],
                  ame_re$estimate[ame_re$term == "poverty"]),
  SE_poverty  = c(ame_probit$std.error[ame_probit$term == "poverty"],
                  ame_lpm$std.error[ame_lpm$term == "poverty"],
                  ame_re$std.error[ame_re$term == "poverty"])
)
ame_comp[, 2:3] <- round(ame_comp[, 2:3], 5)
knitr::kable(ame_comp, caption = "Average Marginal Effects across binary CRE models.")
```

Across all three specifications, the estimated average marginal effect of poverty is approximately $-0.064$ with standard error $\approx 0.034$. The CRE pooled probit provides a robust, computationally simple estimator that allows partial effect estimation without relying on strict serial independence.

## Summary

1. **Linear Panel Data:** Adding unit averages $\bar{X}_i$ allows Pooled OLS and Random Effects to recover the exact Fixed Effects estimates while retaining cluster-robust inference.
2. **Nonlinear Models:** In binary and count models, the CRE approach avoids the incidental parameters problem and allows calculation of Average Marginal Effects.
3. **Robustness:** Pooled estimation with cluster-robust standard errors is preferred over nonlinear random effects models due to robustness against arbitrary serial correlation.
