---
title: "More on Correlated Random Effect (CRE)"
author: "Xiang Ao"
date: "2025-11-11"
---

Jeff Wooldridge suggested using Correlated Random Effect (CRE) for binary outcomes in this twitter post:  https://x.com/jmwooldridge/status/1986100627454206220


Here I summarize what I learned so far:

## linear model

For panel data, the usual set up is:

$$y_{it} = \alpha + X_{it} \beta + v_i + \epsilon_{it} $$

Wooldridge (2021) paper on DiD shows that these models are the same:
1. Two way fixed effect OLS. Dummy variable approach is the same too.
2. Mundlak's CRE approach.  This includes a pooled OLS with Mundlak device (means of $X$'s by group), and CRE with random effects.

The idea of Chamberlain's device is that since we don't observe $v_i$, we project $v_i$ onto the history of $X_i$: $v_i = \psi + \bar X_i \xi + a_i$, and replace $v_i$ with that projection.  Mundlak's device is a special case, where we put equal weight on all $X_{it}$, so the projection of $v_i$ onto $X_i$ reduces to the mean $\bar X_i = \frac{1}{T}\sum_t X_{it}$.  Replace $v_i$ with that projection, and what is left ($a_i$) is by construction uncorrelated with $X_{it}$.


Here is an example:


### example 1


```{r}
#| label: fix1
#| cache: true
#| warning: false
#| message: false
library(dplyr)
library(fixest)
library(bacondecomp)
data("castle")

# Fixed-effects model with individual and time fixed effects
fe_model_twoway <- feols(l_homicide ~ poverty + l_police | state + year, data = castle)
summary(fe_model_twoway)

# random effect
library(lme4)
re_model <- lmer(l_homicide ~ poverty + l_police +  factor(year) +  (1 | state), data = castle)
summary(re_model)

# Mundlak
castle2 <- castle |>
  group_by(state) |>
  mutate(poverty_mean = mean(poverty, na.rm = TRUE), l_police_mean= mean(l_police, na.rm = TRUE))
cre_model <- feols(l_homicide ~ poverty + poverty_mean + l_police + l_police_mean + factor(year), data = castle2)
summary(cre_model)

# Mundlak
cre_model2 <- lmer(l_homicide ~ poverty + poverty_mean + l_police + l_police_mean + factor(year) +  (1 | state), data = castle2)
summary(cre_model2)
```

In this example, fixed effect model, CRE1 and CRE2 all give the same coefficient on poverty.

As we know, RE model assumes $v_i$ is uncorrelated with $X_{it}$, while CRE and FE allow for correlation between $v_i$ and $X_{it}$.

## nonlinear model

What I did not realize before is that this is only valid for linear case.   For OLS to be consistent, we only need contemporaneous exogeneity, which means $E[\epsilon_{it} | X_{it}, v_i] = 0$.  This is not the case for nonlinear cases, such as logit or probit models.

### Binary outcome

For binary outcomes, the model is:

$$\Pr(y_{it}=1 \mid X_{it}, v_i) = \Phi(\alpha + X_{it} \beta + v_i) $$
I used to think it's the best to use conditional logit.  However, it needs "conditional independence" (serial independence could be a better name) to be consistent.  That is, we have to assume the series of $y_{it}$ is conditionally independent of each other given $X_{it}$ and $v_i$.  This is a strong assumption.  If we think we have serial correlation in the error term, then this would not hold.  The other disadvantage of conditional logit (also called fixed effect logit) is that $v_i$'s are not estimated, it is just a nuisance parameter.  Any partial effects cannot be calculated, because any partial effects are functions of $v_i$'s.

Instead, Wooldridge suggests using CRE probit.  Basically adding Mundlak device (group means of $X$'s) to the model, and then use pooled probit.  Note that we do not use RE probit or logit, since that would need conditional independence too.

For nonlinear panel data with unobserved heterogeneity (meaning there is $v_i$ that we don't observe), we have a few assumptions.

First, strict exogeneity,
$$ D(y_{it} | X_{i1}, ..., X_{iT}, v_i) = D(y_{it} | X_{it}, v_i) $$
This means that the distribution of $y_{it}$ only depends on $X_{it}$, given $v_i$.

Second, conditional independence,
$$ D(y_{i1}, y_{i2}, ..., y_{iT} | X_i, v_i) = \prod_{t=1}^T D(y_{it} | X_i, v_i) $$

This means that the distribution of joint distribution of $y_{it}$'s can be modeled by each $y_{it}$ independently, given $X_i$ and $v_i$.  We need this in most nonlinear cases, but not in linear case.

Then we need to specify $D(v_i) | X_i$.  The random effect assumption is saying $$ D(v_i | X_i) = D(v_i) $$, that is, $v_i$ is independent of $X_i$.  This is a strong assumption.  We should try to avoid this assumption.

Now, most nonlinear models make the second assumption, in the panel setting, when we have $v_i$.  That includes RE model, conditional logit.  However, that does not include pooled probit/logit, or CRE with pooled logit/probit.  That is why Wooldridge recommend using CRE with pooled probit.


### Example 2

```{r}
#| label: fix2
#| cache: true
#| warning: false
#| message: false
# generate a binary outcome, high_homocide

castle2 <- castle2 |>
  mutate(high_homicide = ifelse(l_homicide > mean(l_homicide, na.rm = TRUE), 1, 0))

# CRE pooled probit
cre_probit <- glm(high_homicide ~ poverty + poverty_mean + l_police + l_police_mean + factor(year), data = castle2, family = binomial(link = "probit"))
# Compute Average Marginal Effects (AMEs)
library(marginaleffects)
ame_cre_probit <- avg_slopes(cre_probit)
ame_cre_probit


# linear model
cre_lm <- lm(high_homicide ~ poverty + poverty_mean + l_police + l_police_mean + factor(year), data = castle2)
ame_lm <- avg_slopes(cre_lm)
ame_lm

# CRE RE probit
cre_re_probit <- glmer(high_homicide ~ poverty + poverty_mean + l_police + l_police_mean + factor(year) +  (1 | state), data = castle2, family = binomial(link = "probit"))
ame_cre_re_probit <- avg_slopes(cre_re_probit)
ame_cre_re_probit
```

Note on the random-effects probit AMEs: by default, `avg_slopes` (and `marginaleffects` generally) evaluates the random effect at 0, which gives the *conditional* AME (the average marginal effect for a "typical", $v_i=0$, unit). This differs from the *population-averaged* AME, which would integrate the marginal effect over the estimated distribution of $v_i$. When the random-effect variance is large, the two can diverge substantially in a non-linear model like probit, since $\Phi(\cdot)$ is nonlinear in $v_i$. If the population-averaged effect is of interest, it must be computed by simulating from (or numerically integrating over) the estimated random-effects distribution rather than fixing $v_i=0$.

Ultimately we are interested in the partial effect, therefore we can compare across models.  In this case, we should prefer average marginal effect by CRE pooled probit.
