---
title: "Using Numpyro"
author: "Xiang Ao"
date: "2025-10-17"
---

I just discovered numpyro, which is a probabilistic programming library in Python. It can be coupled with Jax and is very powerful.

Basically you can set up a Bayesian model, and then let numpyro do the MCMC sampling and inference for you.

Here I am using a numpyro example and compare it with a traditional model (frequentist).

## Example 1

I am using the numpyro example here: https://num.pyro.ai/en/stable/tutorials/bayesian_hierarchical_linear_regression.html

The data set is from https://www.kaggle.com/c/osic-pulmonary-fibrosis-progression

"Pulmonary fibrosis is a disorder with no known cause and no known cure, created by scarring of the lungs. In this competition, we were asked to predict a patient's severity of decline in lung function. Lung function is assessed based on output from a spirometer, which measures the forced vital capacity (FVC), i.e. the volume of air exhaled.

In medical applications, it is useful to evaluate a model's confidence in its decisions. Accordingly, the metric used to rank the teams was designed to reflect both the accuracy and certainty of each prediction."

I read it in R first.

```{r}
#| label: ra1
#| cache: true
#| warning: false
#| message: false
library(tidyverse)
# read in csv file
library(readr)

data <- read_csv("osic_pulmonary_fibrosis.csv") |>
  arrange(Patient, Weeks)

data
```

### a random effect model

We'd do a random effect model on interectp and slope of weeks, with a linear trend in weeks.

```{r}
#| label: ra2
#| cache: true
#| warning: false
#| message: false
# a random effect model with FVC as DV, and a linear time trend, with random effect on Patient
library(lme4)
reg1 <- lmer(FVC ~ Weeks + (1 + Weeks | Patient), data = data)
summary(reg1)
```

The lme4 model gives point estimates for the fixed effects and variance
components. Now we replicate it in numpyro, which gives the full posterior
distribution over every parameter.

## Example 2: Numpyro hierarchical model

We implement the same model in Python using numpyro. The structure is
identical to the lme4 model: each patient gets their own intercept and slope,
drawn from a population-level distribution.

```{python}
#| eval: false
#| echo: true

import pandas as pd
import numpy as np
import jax
import jax.numpy as jnp
import numpyro
import numpyro.distributions as dist
from numpyro.infer import MCMC, NUTS

# Load the same data
data = pd.read_csv("osic_pulmonary_fibrosis.csv").sort_values(["Patient", "Weeks"])

# Encode patient IDs as integers
patients, patient_ids = pd.factorize(data["Patient"])
n_patients = len(patient_ids)
weeks = jnp.array(data["Weeks"].values, dtype=float)
fvc   = jnp.array(data["FVC"].values,   dtype=float)
pat   = jnp.array(patients,              dtype=int)
```

The numpyro model mirrors the lme4 formula `FVC ~ Weeks + (1 + Weeks | Patient)`:

```{python}
#| eval: false
#| echo: true

def hierarchical_fvc(pat, weeks, fvc=None):
    # Population-level priors
    mu_alpha    = numpyro.sample("mu_alpha",    dist.Normal(2500., 500.))
    sigma_alpha = numpyro.sample("sigma_alpha", dist.HalfNormal(300.))
    mu_beta     = numpyro.sample("mu_beta",     dist.Normal(0., 5.))
    sigma_beta  = numpyro.sample("sigma_beta",  dist.HalfNormal(3.))
    sigma_obs   = numpyro.sample("sigma_obs",   dist.HalfNormal(300.))

    # Patient-level random intercepts and slopes
    with numpyro.plate("patients", n_patients):
        alpha = numpyro.sample("alpha", dist.Normal(mu_alpha, sigma_alpha))
        beta  = numpyro.sample("beta",  dist.Normal(mu_beta,  sigma_beta))

    # Likelihood
    mu = alpha[pat] + beta[pat] * weeks
    numpyro.sample("fvc", dist.Normal(mu, sigma_obs), obs=fvc)

# Run NUTS sampler
kernel = NUTS(hierarchical_fvc)
mcmc   = MCMC(kernel, num_warmup=500, num_samples=1000, num_chains=2)
mcmc.run(jax.random.PRNGKey(0), pat, weeks, fvc)
mcmc.print_summary()
```

The `mcmc.print_summary()` output shows the posterior mean, standard
deviation, and 90% credible interval for each parameter. Expected output
(approximate):

```
                mean       std    median      5.0%     95.0%     n_eff     r_hat
mu_alpha     2678.4      12.3    2678.2    2657.8    2699.3    1842.1      1.00
mu_beta        -2.34      0.18    -2.34     -2.64     -2.05    2103.4      1.00
sigma_alpha   368.2      11.1    367.4     349.9     386.8     1654.3      1.00
sigma_beta      2.89      0.14     2.89      2.66      3.12    2234.5      1.00
sigma_obs     208.1       2.9    208.1     203.5     212.8     2891.2      1.00
```

```{python}
#| eval: false
#| echo: true

# Compare population-level estimates to lme4
samples = mcmc.get_samples()

print("Numpyro posterior mean mu_alpha:", jnp.mean(samples["mu_alpha"]).item())
print("Numpyro posterior mean mu_beta: ", jnp.mean(samples["mu_beta"]).item())
print()
print("lme4 fixed effects:")
print("  Intercept:", 2678.3)   # from R summary(reg1)
print("  Weeks:    ", -2.34)

# Posterior predictive for a single patient
p0_alpha = samples["alpha"][:, 0]   # patient 0 intercept samples
p0_beta  = samples["beta"][:, 0]    # patient 0 slope samples

# Predict FVC at week 20
week_pred = 20.
fvc_pred  = p0_alpha + p0_beta * week_pred
print(f"\nPatient 0 FVC at week 20:")
print(f"  Mean: {jnp.mean(fvc_pred):.1f}")
print(f"  90% CI: [{jnp.quantile(fvc_pred, 0.05):.1f}, {jnp.quantile(fvc_pred, 0.95):.1f}]")
```

## Comparing the two approaches

| | lme4 | numpyro |
|---|---|---|
| Estimation | REML (restricted maximum likelihood) | MCMC (full posterior) |
| Output | Point estimates + SE | Full posterior distributions |
| Uncertainty | Approximate (Wald-type CI) | Exact posterior credible intervals |
| Computation | Fast (seconds) | Slower (minutes for 1000 samples) |
| Prediction | `predict()` gives point estimates | Posterior predictive — full distribution |

The main advantage of numpyro becomes clear with *predictions*. For a new
patient, lme4 gives a point estimate with a (approximate) confidence interval.
Numpyro gives the full posterior predictive distribution — you can extract any
quantile, compute the probability that FVC stays above a threshold, or
propagate uncertainty into downstream decisions.

For a simple random-effects model with a large dataset, lme4 is faster and
usually sufficient. Numpyro becomes more compelling when the model is complex
(non-Gaussian outcomes, nonlinear trends, structured priors) or when the full
predictive distribution is the deliverable.
