---
title: "Using Numpyro"
author: "Xiang Ao"
date: "2025-10-17"
---

I just discovered numpyro, which is a probabilistic programming library in Python. It can be coupled with Jax and is very powerful.

Basically you can set up a Bayesian model, and then let numpyro do the MCMC sampling and inference for you.

Here I am using a numpyro example and compare it with a traditional model (frequentist).

## Example 1

I am using the numpyro example here: https://num.pyro.ai/en/stable/tutorials/bayesian_hierarchical_linear_regression.html

The data set is from https://www.kaggle.com/c/osic-pulmonary-fibrosis-progression

"Pulmonary fibrosis is a disorder with no known cause and no known cure, created by scarring of the lungs. In this competition, we were asked to predict a patient's severity of decline in lung function. Lung function is assessed based on output from a spirometer, which measures the forced vital capacity (FVC), i.e. the volume of air exhaled.

In medical applications, it is useful to evaluate a model's confidence in its decisions. Accordingly, the metric used to rank the teams was designed to reflect both the accuracy and certainty of each prediction."

I read it in R first.

```{r}
#| label: ra1
#| cache: true
#| warning: false
#| message: false
library(tidyverse)
# read in csv file
library(readr)

data <- read_csv("osic_pulmonary_fibrosis.csv") |>
  arrange(Patient, Weeks)

data
```

### a random effect model

We'd do a random effect model on intercept and slope of weeks, with a linear trend in weeks.

```{r}
#| label: ra2
#| cache: true
#| warning: false
#| message: false
# a random effect model with FVC as DV, and a linear time trend, with random effect on Patient
library(lme4)
reg1 <- lmer(FVC ~ Weeks + (1 + Weeks | Patient), data = data)
summary(reg1)
```

The data are the OSIC pulmonary-fibrosis panel: 1,549 spirometry readings from
176 patients, with forced vital capacity (`FVC`) as the outcome and `Weeks`
since baseline as the time variable. The model gives each patient a random
intercept and slope.

The fixed effects are an intercept of 2810.3 (SE 62.9) and a slope of $-4.263$
(SE 0.438): average FVC declines by about 4.3 units per week, and the $t$ of
$-9.7$ leaves little doubt about the direction. The variance components show
where the heterogeneity lives — a patient-intercept standard deviation of 828.7
against a residual of 136.4, so patients differ far more in level than any single
measurement deviates from that patient's own line. The slope SD of 5.09 against
a mean slope of $-4.26$ means the rate of decline itself varies substantially
across patients, with some improving.

The `lme4` model gives point estimates for the fixed effects and variance
components. Now do the same model in `numpyro`, where the output is a posterior
distribution for the parameters.

## Example 2: Numpyro hierarchical model

The Python model has the same structure as the `lme4` model: each patient gets
an intercept and slope, drawn from a population-level distribution.

> The NumPyro/Python chunks in this chapter are shown for reading but not
> executed (`eval: false`), because the render environment does not have
> `numpyro`/`jax` installed. Run them in a Python environment with those
> packages to reproduce the posterior summaries described in the text.

```{python}
#| eval: false
#| echo: true

import pandas as pd
import numpy as np
import jax
import jax.numpy as jnp
import numpyro
import numpyro.distributions as dist
from numpyro.infer import MCMC, NUTS

# Load the same data
data = pd.read_csv("osic_pulmonary_fibrosis.csv").sort_values(["Patient", "Weeks"])

# Encode patient IDs as integers
patients, patient_ids = pd.factorize(data["Patient"])
n_patients = len(patient_ids)
weeks = jnp.array(data["Weeks"].values, dtype=float)
fvc   = jnp.array(data["FVC"].values,   dtype=float)
pat   = jnp.array(patients,              dtype=int)
```

Drawing `alpha` and `beta` from independent Normal priors, as below, gives
patient-level intercepts and slopes with *zero* covariance by construction --
that corresponds to `FVC ~ Weeks + (1|Patient) + (0+Weeks|Patient)`, not
`(1+Weeks|Patient)`. `lme4`'s `(1+Weeks|Patient)` estimates a full
(unstructured, correlated) 2x2 covariance matrix for the intercept and
slope. To match that in `numpyro`, draw `(alpha, beta)` jointly from a
`MultivariateNormal` with an LKJ-Cholesky prior on the correlation matrix:

```{python}
#| eval: false
#| echo: true

def hierarchical_fvc(pat, weeks, fvc=None):
    # Population-level priors
    mu_alpha    = numpyro.sample("mu_alpha",    dist.Normal(2500., 500.))
    mu_beta     = numpyro.sample("mu_beta",     dist.Normal(0., 5.))
    sigma_obs   = numpyro.sample("sigma_obs",   dist.HalfNormal(300.))

    # Correlated patient-level random intercepts and slopes: LKJ prior on
    # the correlation matrix, HalfNormal priors on the per-parameter scales,
    # combined into a Cholesky-factorized covariance -- this is the numpyro
    # analogue of lme4's unstructured (1 + Weeks | Patient).
    sigma_re = numpyro.sample("sigma_re", dist.HalfNormal(jnp.array([300., 3.])))
    L_corr   = numpyro.sample("L_corr", dist.LKJCholesky(2, concentration=2.0))
    L_cov    = jnp.diag(sigma_re) @ L_corr

    mu_re = jnp.stack([mu_alpha, mu_beta])
    with numpyro.plate("patients", n_patients):
        re = numpyro.sample("re", dist.MultivariateNormal(mu_re, scale_tril=L_cov))
    alpha, beta = re[:, 0], re[:, 1]
    # Record these as sites, or they are only local variables and
    # mcmc.get_samples() will not contain "alpha" or "beta".
    numpyro.deterministic("alpha", alpha)
    numpyro.deterministic("beta", beta)

    # Likelihood
    mu = alpha[pat] + beta[pat] * weeks
    numpyro.sample("fvc", dist.Normal(mu, sigma_obs), obs=fvc)

# Run NUTS sampler
kernel = NUTS(hierarchical_fvc)
mcmc   = MCMC(kernel, num_warmup=500, num_samples=1000, num_chains=2)
mcmc.run(jax.random.PRNGKey(0), pat, weeks, fvc)
mcmc.print_summary()
```

`mcmc.print_summary()` reports posterior means, standard deviations and credible
intervals for each site: `mu_alpha`, `mu_beta`, the two-vector `sigma_re`,
`L_corr`, and `sigma_obs`.

No output is shown here, and that is deliberate. This chapter's Python is not
executed during the render, so any table printed at this point would be written
by hand rather than produced by the model. An earlier version of this section did
exactly that, and the invented numbers drifted away from the model in two visible
ways: they reported `sigma_alpha` and `sigma_beta`, which this parameterization
does not create -- it has the vector site `sigma_re` -- and they disagreed with
the `lme4` fixed effects fitted above. Treat the code below as a sketch of the
comparison, not as a record of one that was run.

```{python}
#| eval: false
#| echo: true

# Compare population-level estimates to lme4
samples = mcmc.get_samples()

print("Numpyro posterior mean mu_alpha:", jnp.mean(samples["mu_alpha"]).item())
print("Numpyro posterior mean mu_beta: ", jnp.mean(samples["mu_beta"]).item())
print()
print("lme4 fixed effects (from summary(reg1) above):")
print("  Intercept:", 2810.3)
print("  Weeks:    ", -4.263)

# Posterior predictive for a single patient
p0_alpha = samples["alpha"][:, 0]   # patient 0 intercept samples
p0_beta  = samples["beta"][:, 0]    # patient 0 slope samples

# Predict FVC at week 20 -- posterior predictive (add observation noise)
week_pred = 20.
mu_pred   = p0_alpha + p0_beta * week_pred
fvc_pred  = dist.Normal(mu_pred, samples["sigma_obs"]).sample(jax.random.PRNGKey(1))
print(f"\nPatient 0 FVC at week 20:")
print(f"  Mean: {jnp.mean(fvc_pred):.1f}")
print(f"  90% CI: [{jnp.quantile(fvc_pred, 0.05):.1f}, {jnp.quantile(fvc_pred, 0.95):.1f}]")
```

## Comparing the two approaches

| | lme4 | numpyro |
|---|---|---|
| Estimation | REML (restricted maximum likelihood) | MCMC (full posterior) |
| Output | Point estimates + SE | Full posterior distributions |
| Uncertainty | Approximate (Wald-type CI) | Simulation-based posterior credible intervals |
| Computation | Fast (seconds) | Slower (minutes for 1000 samples) |
| Prediction | `predict()` gives point estimates | Posterior predictive — full distribution |

The main advantage of `numpyro` shows up in prediction. For a new patient,
`lme4` gives a point prediction and an approximate interval. `numpyro` gives a
posterior predictive distribution. Then we can look at any quantile or compute
probabilities such as whether FVC stays above a threshold.

For a simple random-effects model with a large data set, `lme4` is much faster
and usually enough. `numpyro` becomes more useful when the model is more
complicated, or when the full predictive distribution is what we need.

---

<!-- see-also-footer -->
*Systematic treatment: [R](https://xiangao.github.io/causal_econometrics_guide/bayesian-causal.html) · [Julia](https://xiangao.github.io/causal_econometrics_julia/bayesian-causal.html).*
