---
title: "Partial Interference"
date: "2026-03-21"
---

Most causal inference methods assume SUTVA — the Stable Unit Treatment Value Assumption — which says that one unit's treatment doesn't affect another unit's outcome.  In practice, this is often violated.  Vaccinating your neighbor reduces *your* risk of infection.  A job training program for one worker may crowd out another's employment prospects.  A student's test score depends on whether classmates received tutoring.

**Partial interference** (Sobel, 2006; Hudgens & Halloran, 2008) relaxes SUTVA in a structured way: units are organized into clusters, interference happens *within* clusters but not *between* them.  This is weaker than full interference (where everyone affects everyone) but still much harder than the standard no-interference setting.

Qu et al. (2021) develop efficient estimators — clustered IPW and AIPW — for heterogeneous treatment effects under partial interference.  The `CausalModel` R package implements these estimators alongside standard observational and experimental methods.

# The Setup

## Clusters and Groups

Each unit $i$ belongs to a **cluster** $c$ (e.g., a household, classroom, or village).  Within each cluster, units are organized into **groups** indexed by $j = 0, 1, \ldots, J-1$.  The **group structure** $\mathbf{s} = (s_0, s_1, \ldots, s_{J-1})$ gives the number of units per group.

For example, with `group_struct = c(2, 3)`:

- Group 0 has 2 units
- Group 1 has 3 units
- Each cluster has 5 units total

Groups can represent substantively different roles (e.g., parents vs children, different job categories) or simply partition units for modeling flexibility.

## The Neighborhood Configuration G

The key object is the **neighborhood configuration** $G_i = (g_0, g_1, \ldots, g_{J-1})$, where $g_j$ counts how many *other* units in group $j$ (within unit $i$'s cluster) are treated.  "Other" means excluding unit $i$ itself.

With `group_struct = c(2, 3)`:

- If unit $i$ is in group 0, then $g_0 \in \{0, 1\}$ (the other unit in group 0 is treated or not) and $g_1 \in \{0, 1, 2, 3\}$ (0 to 3 units in group 1 are treated)
- $G$ encodes the full treatment environment that unit $i$ experiences

## Heterogeneous Treatment Effects β(G)

Under partial interference, the treatment effect for unit $i$ depends on its neighborhood $G_i$:

$$\beta(G) = E[Y_i(1, G) - Y_i(0, G)]$$

where $Y_i(z, G)$ is unit $i$'s potential outcome when its own treatment is $z$ and its neighborhood configuration is $G$.

A natural model for the treatment effect is **linear in the neighborhood**:

$$\beta(G) = \tau + \sum_{j=0}^{J-1} \gamma_j \cdot g_j$$

where:

- $\tau$ is the **direct effect** — the effect of treatment when no neighbors are treated
- $\gamma_j$ is the **spillover effect** from group $j$ — the additional effect per treated neighbor in group $j$

### Example

With `group_struct = c(2, 3)`, $\tau = 1$, $\gamma_0 = 0.5$, $\gamma_1 = 0.1$:

| Neighborhood $G$ | Formula | $\beta(G)$ |
|---|---|---|
| $(0, 0)$ | $1 + 0 \times 0.5 + 0 \times 0.1$ | 1.0 |
| $(1, 0)$ | $1 + 1 \times 0.5 + 0 \times 0.1$ | 1.5 |
| $(0, 1)$ | $1 + 0 \times 0.5 + 1 \times 0.1$ | 1.1 |
| $(1, 1)$ | $1 + 1 \times 0.5 + 1 \times 0.1$ | 1.6 |

Having one treated neighbor in group 0 adds 0.5 to the treatment effect; having one in group 1 adds only 0.1.

# Estimation

## Two Propensity Models

Standard IPW uses one propensity score: $P(Z_i = 1 | X_i)$.  Under interference, we need two:

1. **Individual propensity**: $P(Z_i = 1 | X_i)$ — probability of being treated given covariates (logistic regression)
2. **Neighborhood propensity**: $P(G_i = g | X_i)$ — probability of having neighborhood configuration $g$ given covariates (multinomial logistic regression)

The neighborhood propensity is the new ingredient.  It captures how likely a particular treatment environment is, conditional on covariates.

## Clustered IPW

The clustered IPW estimator for $\beta(g)$ reweights outcomes by both propensity scores:

$$\hat{\beta}_{\text{IPW}}(g) = \frac{1}{N} \sum_i \left[ \frac{\mathbf{1}(Z_i=1, G_i=g)}{P(Z_i=1|X_i) \cdot P(G_i=g|X_i)} - \frac{\mathbf{1}(Z_i=0, G_i=g)}{P(Z_i=0|X_i) \cdot P(G_i=g|X_i)} \right] Y_i$$

This gives a separate treatment effect estimate at each neighborhood configuration $g$.

## Clustered AIPW (Doubly Robust)

The AIPW version adds outcome modeling to gain robustness:

- Fit separate outcome models $E[Y | X, Z, G=g]$ for each configuration $g$
- Combine with propensity weighting so the estimator is consistent if *either* the outcome model *or* both propensity models are correct

This is the recommended estimator — it inherits the double robustness property from the standard AIPW framework.

## Variance Estimation

Standard errors are computed via a matching-based variance estimator.  Within each group and neighborhood configuration, units are matched on standardized covariates, and the variance is estimated from the matched pairs.  This avoids the need for bootstrap (which is expensive with clusters) or asymptotic formulas that may not apply at moderate sample sizes.

# Using CausalModel

```{r}
#| label: setup
#| warning: false
#| cache: true
#| message: false
library(CausalModel)
library(knitr)
set.seed(42)
```

## Generate Clustered Data

The `generate_fixed_cluster()` function creates synthetic data with known interference structure.

```{r}
#| label: gen_data
#| warning: false
#| cache: true
#| message: false
cdata <- generate_fixed_cluster(
  clusters = 200,
  group_struct = c(2, 3),   # 2 units in group 0, 3 in group 1
  tau = 1,                   # direct effect
  gamma = c(0.5, 0.1)        # spillover: 0.5 per treated in group 0, 0.1 in group 1
)

cat("Total units:", length(cdata$Y), "\n")
cat("Units per cluster:", sum(c(2, 3)), "\n")
cat("Treatment prevalence:", round(mean(cdata$Z), 3), "\n")
```

Each unit has:

- `Y` — observed outcome
- `Z` — treatment indicator (0/1)
- `X` — individual covariates
- `cluster_labels` — which cluster the unit belongs to
- `group_labels` — which group within the cluster (0-indexed)
- `ingroup_labels` — position within the group (0-indexed)

## Construct the Clustered Object

```{r}
#| label: clustered_obj
#| warning: false
#| cache: true
#| message: false
cl_obj <- clustered(
  Y = cdata$Y,
  Z = cdata$Z,
  X = cdata$X,
  cluster_labels = cdata$cluster_labels,
  group_labels = cdata$group_labels,
  ingroup_labels = cdata$ingroup_labels,
  n_matches = 50
)
```

The `clustered()` constructor does several things internally:

1. Augments individual covariates with cluster-level aggregates (mean of neighbors' covariates)
2. Fits the individual propensity model (logistic regression)
3. Fits the neighborhood propensity model (multinomial logistic regression)
4. Computes $G$ for each unit — the number of treated neighbors per group

## Estimate Treatment Effects

### Clustered AIPW

```{r}
#| label: aipw_est
#| warning: false
#| cache: true
#| message: false
#| results: hide
result_aipw <- est_via_aipw(cl_obj)
```

The result is a list with one element per group.  Each element contains `beta_g` (treatment effects at each neighborhood configuration) and `se` (standard errors):

```{r}
#| label: aipw_show
#| warning: false
#| cache: true
#| message: false
for (j in seq_along(result_aipw)) {
  cat(sprintf("Group %d:\n", j - 1))
  valid <- !is.nan(result_aipw[[j]]$beta_g) & result_aipw[[j]]$se > 0
  df <- data.frame(
    g_index = which(valid) - 1,
    beta_g = round(result_aipw[[j]]$beta_g[valid], 3),
    se = round(result_aipw[[j]]$se[valid], 3)
  )
  print(df, row.names = FALSE)
  cat("\n")
}
```

### Clustered IPW

```{r}
#| label: ipw_est
#| warning: false
#| cache: true
#| message: false
#| results: hide
result_ipw <- est_via_ipw(cl_obj)
```

```{r}
#| label: ipw_show
#| warning: false
#| cache: true
#| message: false
cat("Clustered IPW beta_g (group 0):\n")
valid <- !is.nan(result_ipw[[1]]$beta_g) & result_ipw[[1]]$se > 0
round(result_ipw[[1]]$beta_g[valid], 3)
```

## Interpreting the Results

The neighborhood configurations are encoded as integers.  For `group_struct = c(2, 3)`, the encoding for group 0 is:

$$G_{\text{encoded}} = g_0 + g_1 \times (s_0 + 1)$$

where $s_0 = 2$ is the size of group 0.  The stride for the second dimension is $s_0 + 1 = 3$ because the package allocates slots for $g_0 \in \{0, 1, 2\}$ (even though $g_0 = 2$ is impossible for units in group 0, since there's only one *other* unit in that group).

| $g_0$ | $g_1$ | Encoded | True $\beta$ |
|---|---|---|---|
| 0 | 0 | 0 | 1.0 |
| 1 | 0 | 1 | 1.5 |
| 0 | 1 | 3 | 1.1 |
| 1 | 1 | 4 | 1.6 |
| 0 | 2 | 6 | 1.2 |
| 1 | 2 | 7 | 1.7 |

Not all configurations may have enough observations for reliable estimation — the `se > 0` filter removes degenerate cases.

# Monte Carlo Validation

To verify that the estimators work correctly, we run a Monte Carlo study: generate data 100 times, estimate $\beta(G)$ each time, and check bias, SE calibration, and coverage.

```{r}
#| label: mc_setup
#| warning: false
#| cache: true
#| message: false
n_clusters <- 200
group_struct <- c(2, 3)
tau_int <- 1
gamma_int <- c(0.5, 0.1)
n_reps <- 100

# True beta(g) function
true_beta <- function(g) tau_int + sum(g * gamma_int)

# Track results for group 0 at three neighborhood configurations
g_indices <- list(c(0, 0), c(1, 0), c(0, 1))
g_encoded <- sapply(g_indices, function(g) {
  g[1] + g[2] * (group_struct[1] + 1) + 1  # 1-indexed; stride = max_g0 + 1
})
true_values <- sapply(g_indices, true_beta)

int_results <- array(NA, dim = c(n_reps, length(g_indices), 2),
                      dimnames = list(NULL, NULL, c("beta", "se")))
```

```{r}
#| label: mc_run
#| warning: false
#| cache: true
#| message: false
#| results: hide
set.seed(2024)
for (i in seq_len(n_reps)) {
  cdata <- generate_fixed_cluster(
    clusters = n_clusters, group_struct = group_struct,
    tau = tau_int, gamma = gamma_int
  )
  cl_obj <- clustered(
    Y = cdata$Y, Z = cdata$Z, X = cdata$X,
    cluster_labels = cdata$cluster_labels,
    group_labels = cdata$group_labels,
    ingroup_labels = cdata$ingroup_labels,
    n_matches = 50
  )
  res <- est_via_aipw(cl_obj)

  for (j in seq_along(g_indices)) {
    idx <- g_encoded[j]
    int_results[i, j, "beta"] <- res[[1]]$beta_g[idx]
    int_results[i, j, "se"] <- res[[1]]$se[idx]
  }
}
```

```{r}
#| label: mc_table
#| warning: false
#| cache: true
#| message: false
int_rows <- list()
for (j in seq_along(g_indices)) {
  betas <- int_results[, j, "beta"]
  ses <- int_results[, j, "se"]

  valid <- !is.nan(betas) & !is.nan(ses) & ses > 0
  betas <- betas[valid]
  ses <- ses[valid]
  tv <- true_values[j]

  ci_lo <- betas - 1.96 * ses
  ci_hi <- betas + 1.96 * ses

  int_rows[[j]] <- data.frame(
    g = paste0("(", paste(g_indices[[j]], collapse = ","), ")"),
    True_beta = tv,
    Mean_est = mean(betas),
    Bias = mean(betas) - tv,
    Emp_SE = sd(betas),
    Mean_SE = mean(ses),
    Coverage = mean(ci_lo <= tv & tv <= ci_hi),
    N_valid = length(betas)
  )
}
tbl_int <- do.call(rbind, int_rows)
kable(tbl_int, digits = 3, row.names = FALSE,
      caption = sprintf("Clustered AIPW (group 0, %d clusters, %d reps)",
                        n_clusters, n_reps))
```

```{r}
#| label: mc_hist
#| warning: false
#| cache: true
#| message: false
#| fig-cap: "Studentized statistics under clustered AIPW"
#| fig-height: 4
par(mfrow = c(1, length(g_indices)), mar = c(3, 3, 2, 1))
for (j in seq_along(g_indices)) {
  betas <- int_results[, j, "beta"]
  ses <- int_results[, j, "se"]
  valid <- !is.nan(betas) & !is.nan(ses) & ses > 0
  t_stats <- (betas[valid] - true_values[j]) / ses[valid]
  g_label <- paste0("(", paste(g_indices[[j]], collapse = ","), ")")
  hist(t_stats, breaks = 20, freq = FALSE, col = "lightblue",
       main = sprintf("g = %s", g_label),
       xlab = "", xlim = c(-4, 4))
  curve(dnorm(x), add = TRUE, col = "red", lwd = 2)
  abline(v = 0, lty = 2)
}
```

The studentized statistics should approximately follow $N(0,1)$ if the estimator and SE are both consistent.

# What If You Ignore Interference?

What happens if you use a standard estimator (say, AIPW without the interference structure) on data with spillovers?  You get a single number that averages over all neighborhood configurations — and it doesn't correspond to any well-defined causal estimand.

```{r}
#| label: ignore_interference
#| warning: false
#| cache: true
#| message: false
# Generate one dataset with interference
set.seed(42)
cdata <- generate_fixed_cluster(
  clusters = 200,
  group_struct = c(2, 3),
  tau = 1,
  gamma = c(0.5, 0.1)
)

# Naive: ignore clusters, treat as standard observational data
obs_naive <- observational(cdata$Y, cdata$Z, cdata$X)
naive_ate <- est_via_aipw(obs_naive)

cat("Naive AIPW (ignoring interference):", round(naive_ate$ate, 3), "\n")
cat("SE:", round(naive_ate$se, 3), "\n\n")

# Correct: use clustered estimator
cl_obj <- clustered(
  Y = cdata$Y, Z = cdata$Z, X = cdata$X,
  cluster_labels = cdata$cluster_labels,
  group_labels = cdata$group_labels,
  ingroup_labels = cdata$ingroup_labels,
  n_matches = 50
)
result <- est_via_aipw(cl_obj)

cat("Clustered AIPW beta_g (group 0):\n")
valid <- !is.nan(result[[1]]$beta_g) & result[[1]]$se > 0
df <- data.frame(
  g_index = which(valid) - 1,
  beta_g = round(result[[1]]$beta_g[valid], 3),
  se = round(result[[1]]$se[valid], 3)
)
print(df, row.names = FALSE)
```

The naive estimate gives you *some* weighted average of the $\beta(g)$ values, but you can't tell how much is direct effect and how much is spillover.  The clustered estimator decomposes the total effect into interpretable components — exactly the information you need for policy design.  If the spillover is large, expanding treatment to clusters (rather than individuals) may be much more effective.

# Summary

- **Partial interference** relaxes SUTVA within clusters: a unit's outcome depends on neighbors' treatment, not just its own
- The **neighborhood configuration** $G = (g_0, g_1, \ldots)$ encodes the treatment environment — how many neighbors in each group are treated
- Treatment effects $\beta(G)$ vary by neighborhood: $\beta(G) = \tau + \sum \gamma_j g_j$ (direct effect plus spillovers)
- **Clustered IPW/AIPW** estimates $\beta(G)$ at each configuration, requiring two propensity models (individual + neighborhood)
- **Clustered AIPW** is doubly robust — consistent if either the outcome model or both propensity models are correct
- Ignoring interference gives a single number that mixes direct and spillover effects — not useful for policy
- R package: `CausalModel`, an R port based on Qu et al. (2021), "Efficient Treatment Effect Estimation in Observational Studies under Heterogeneous Partial Interference" (arXiv:2107.12420)
