---
title: "Poisson model with exposure"
date: "2024-04-20"
---

## Poisson with exposure or Poisson rate

Poisson models for count data often arise when units are observed for different
lengths of time. A hospital admission count for a patient followed for 30 days
is not directly comparable to a count for a patient followed for 365 days. The
natural quantity of interest is the *rate* — admissions per unit time — not
the raw count.

There are three related but distinct specifications, each making a different
assumption:

**Model 1 — Poisson with offset (rate model):**

$$\log\!\bigl(E[Y]/t\bigr) = \beta_0 + \beta_1 X$$

Equivalently, $\log E[Y] = \beta_0 + \beta_1 X + \log t$, where $\log t$ is
included as a *fixed* offset (coefficient constrained to 1). This is the
standard exposure model: the expected count is proportional to exposure time.

**Model 2 — Poisson on the ratio $Y/t$:**

$$\log\!\bigl(E[Y/t]\bigr) = \beta_0 + \beta_1 X$$

This looks similar, but it is not equivalent to the offset model.  Because $t$ is
a fixed constant for each unit, $E[Y/t] = E[Y]/t$ exactly (linearity of
expectation), so the difference is not from Jensen's inequality.  Rather, the log
link is applied to the ratio rather than the count, and the variance of $Y/t$ is
$\lambda/t$ (it scales with $1/t$), so the Poisson variance assumption no longer
holds.  In practice the two can give similar point estimates, but the standard
errors differ because the variance structure changes when you divide by $t$.

**Model 3 — Unconstrained log-exposure:**

$$\log E[Y] = \beta_0 + \beta_1 X + \gamma \log t$$

Here $\gamma$ is estimated from the data. If $\hat\gamma \approx 1$, the
proportionality assumption is supported. If $\hat\gamma < 1$, counts grow
sub-linearly with exposure — a testable, substantive claim.

## Simulation

```{r}
#| label: sim-data
#| cache: true
#| warning: false
#| message: false

library(tidyverse)
set.seed(42)

n <- 800

# Simulate a health study: count of adverse events
# Covariate: age group (0 = younger, 1 = older)
age_old <- rbinom(n, 1, 0.45)

# Exposure: days under observation (varies 7–365)
exposure <- sample(seq(7, 365, by = 7), n, replace = TRUE)

# True model: rate = exp(-3 + 1.2 * age_old)
# Count ~ Poisson(rate * exposure)
true_rate <- exp(-3 + 1.2 * age_old)
count <- rpois(n, lambda = true_rate * exposure)

df <- data.frame(count, age_old, exposure, log_exposure = log(exposure))
summary(df[, c("count", "exposure")])
```

The data has a wide range of exposure times, so ignoring them would badly
confound the comparison between age groups.

## Fitting the three models

```{r}
#| label: fit-models
#| cache: true
#| warning: false

# Model 1: offset — coefficient on log(exposure) constrained to 1
m1 <- glm(count ~ age_old + offset(log(exposure)),
          family = poisson, data = df)

# Model 2: Poisson on the rate Y/t (different response)
df$rate <- df$count / df$exposure
m2 <- glm(rate ~ age_old,
          family = poisson, data = df)

# Model 3: log(exposure) as a free covariate
m3 <- glm(count ~ age_old + log_exposure,
          family = poisson, data = df)

# Collect coefficients
results <- data.frame(
  Model = c("1. Offset (constrained γ=1)",
            "2. Rate Y/t as response",
            "3. Unconstrained log(t)"),
  Intercept = c(coef(m1)[1], coef(m2)[1], coef(m3)[1]),
  age_old   = c(coef(m1)[2], coef(m2)[2], coef(m3)[2]),
  gamma_log_t = c(1, NA, coef(m3)[3])
)

knitr::kable(results, digits = 3,
  caption = "Coefficient estimates: true intercept = −3, true age effect = 1.2, true γ = 1")
```

**Reading the table:** Model 1 and Model 3 give nearly identical age-group
coefficients because the true $\gamma = 1$ (counts are indeed proportional to
exposure time). Model 2 differs slightly in the intercept because the response
variable changes. The estimated $\hat\gamma$ from Model 3 should be close to 1.

## When $\gamma \ne 1$

To see why the unconstrained model matters, repeat the simulation with a
sub-linear count process ($\gamma = 0.6$: an initial burst of events that
doesn't grow proportionally with long follow-up):

```{r}
#| label: nonlinear-exposure
#| cache: true
#| warning: false

count2 <- rpois(n, lambda = true_rate * exposure^0.6)
df2 <- data.frame(count2, age_old, exposure, log_exposure = log(exposure))

m1b <- glm(count2 ~ age_old + offset(log(exposure)),
           family = poisson, data = df2)
m3b <- glm(count2 ~ age_old + log_exposure,
           family = poisson, data = df2)

cat("Offset model age coefficient:      ", round(coef(m1b)[2], 3), "\n")
cat("Unconstrained model age coefficient:", round(coef(m3b)[2], 3), "\n")
cat("Estimated gamma (log-exposure):    ", round(coef(m3b)[3], 3),
    " (true = 0.6)\n")
```

When $\gamma \ne 1$, forcing the offset constraint biases the treatment
coefficient. The unconstrained model recovers both the true age effect and
the true $\gamma$.

## Stata equivalent

```stata
* Model 1: offset
poisson count age_old, exposure(exposure)

* Model 2: rate as response
gen rate = count / exposure
poisson rate age_old

* Model 3: unconstrained log(t)
gen log_exposure = log(exposure)
poisson count age_old log_exposure
```

## Which model to use?

| Situation | Recommended model |
|---|---|
| Counts grow proportionally with time | Model 1 (offset) |
| Uncertain about proportionality | Model 3 (unconstrained), test $\hat\gamma = 1$ |
| Rate is the natural estimand | Model 1 or Model 2; both are valid |
| Long follow-up with declining event rate | Model 3 or negative binomial with offset |

In most health and social-science applications, Model 1 (Poisson with offset)
is the default. Model 3 is the diagnostic check: fit it, look at $\hat\gamma$,
and return to Model 1 if the constraint is supported by the data.

---

<!-- see-also-footer -->
*Systematic treatment: [R](https://xiangao.github.io/causal_econometrics_guide/poisson-iv.html) · [Julia](https://xiangao.github.io/causal_econometrics_julia/poisson-iv.html).*
