---
title: "IV in Fixed effect Poisson with many fixed effects"
date: "2024-08-29"
---


## Poisson regression with many fixed effects

I have a few posts about Poisson regression for non-negative outcomes, including continuous variable, count variable, or corner solution (such as non-negative outcome with many zeros).  The benefits include that we can use QMLE Poisson, which means we only need to have the mean of outcome specified correctly, as an exponentiated function of covariates.  There is no assumption that the distribution needs to be a Poisson distribution.  When we want to model non-negative outcome with fixed effects, fixed effect Poisson should be our first choice also, because fixed effect Poisson is shown to be consistent without the assumption of the distribution of the error term, and it does not have the "incidental parameter" problem.

Stata has "xtpoisson" to do fixed effect Poisson. However, we would recommend "ppmlhdfe".  "ppmlhdfe is a Stata package that implements Poisson pseudo-maximum likelihood regressions (PPML) with multi-way fixed effects".  It is faster than xtpoisson.  More importantly, it does report the predicted value correctly.  I mentioned in past post that we need to be cautious when we need to use predicted values after "xtpoisson, fe", or "xtreg, fe".  The reason is that these are models estimated after "absorbing" the fixed effects, meaning fixed effects are not actually estimated.  Therefore the predicted values are without fixed effects.  But in "ppmlhdfe" we get the correct predicted values with fixed effect.

On a side note, the same author has "reghdfe" as a replacement for "areg" in Stata.  "reghdfe" is faster than "areg", or "xtreg, fe".  It also allows for many levels of fixed effects.  As "ppmlhdfe", it also provides correct predicted values with fixed effects.


## IV poisson with many fixed effects

Stata has a "ivpoisson" function with handles IV in a Poisson regression setting.  However, it does not have the option to include fixed effects.

This estimator has a published treatment.  Lin and Wooldridge (2019), "Testing and correcting for endogeneity in nonlinear unobserved effects models," treats exactly this setup: their **Procedure 3** is a reduced-form residual added to an FE Poisson, with a robust Wald test on the residual's coefficient serving as a test of endogeneity.  Several points below come from it.

One warning about that test, since it is easy to misuse.  It is tempting to run it and then decide which model to report: if the residual's coefficient is insignificant, fall back on the plain fixed-effects estimate.  Do not do this.  Guggenberger (2010) studies exactly this two-stage procedure and finds that its asymptotic size **equals one** --- the two-stage test rejects with probability approaching one --- because the pretest lacks power against correlations local to zero, so you frequently fail to reject when the regressor is in fact endogenous and then report the non-robust estimate.  He examines both the linear IV version, where the pretest is a test of exogeneity of a regressor, and the random-versus-fixed-effects panel version, and recommends against the two-stage procedure in both.  Both of his results are for linear models, so the transfer to a nonlinear control function is an analogy rather than a theorem, but the mechanism does not depend on linearity.  Report the coefficient on the residual as an estimate of the degree of endogeneity, and report both models side by side; do not let the test choose for you.

In the case that the endogenous variable is continuous and we can use a linear model for the first stage, Wooldridge 2010 textbook (chapter 18) suggested using control function approach, which is very simple:

1. Run the first stage regression (the reduced form for the endogenous variable), and get the residual.  This is the first-stage **control-function** residual.  Note in the many fixed effects case, we need to include all fixed effects in the first stage regression.  In the case of stata, we would use "reghdfe" here.

    A terminology note, because two different objects are easy to conflate.  The **Mundlak device** means adding group *means* of covariates as regressors, as in the [Mundlak chapter](mundlak-device.qmd).  The **control-function residual** here is the residual from a reduced form.  They are different constructions.  But a third object borrows the name, and it matters below: Lin and Wooldridge (2019) call the residual from a reduced form *that itself includes the Mundlak means* --- from regressing $y_{it2}$ on $z_{it}$ and $\bar z_i$ --- the **Mundlak residual**, as against the **FE residual** obtained by within-transforming.  So in that paper "Mundlak residual" denotes a control-function residual, not the device itself.

2. There is a useful result about which of those two residuals to use.  Lin and Wooldridge (2019, appendices A.1 and A.2) show that in FE Poisson they give **numerically identical** estimates of both the structural coefficient and the coefficient on the control function.  The reason is that the two residuals differ only by a term constant within unit, and the FE Poisson score is unaffected by adding anything unit-constant.

    This has a practical consequence.  It means the first stage does not need the thousands of unit dummies: the unit means will do, at no cost in the estimate.  That is what makes a **nonlinear** first stage practical --- you cannot put 2,000 dummies into a logit without an incidental-parameters problem, but you can put in 2,000 unit means.  Note the result as stated is for a *linear* reduced form; see the caveat on discrete endogenous variables below.

3. Include both the endogenous variable and the control-function residual as additional regressors in the second stage, the original Poisson regression.  Use "ppmlhdfe" here.

Note that the standard errors would be incorrect if we use the standard errors from the second stage regression.  We need to use the bootstrap method to get the correct standard errors.  This is not merely a precaution: Lin and Wooldridge (2019, p. 30) recommend exactly this, namely a panel bootstrap in which both estimating steps are done with each bootstrap sample.  The reason is that the control-function residual is generated from the first stage, so we need to bootstrap the whole process to get the correct standard errors.  In the case of clustered standard errors, we need to bootstrap the clusters.



## Example


```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#| include: false

library(Statamarkdown)
stataexe <- find_stata()
#stataexe <- "/usr/local/bin/stata"
knitr::opts_chunk$set(engine.path=list(stata=stataexe))
```

```{stata}
*| label: stata1
*| echo: true
*| collectcode: true

* control function approach
webuse website
reghdfe time phone frfam, absorb(ad female) resid
predict double u2h_fe, resid
ppmlhdfe visits time u2h_fe frfam, absorb(ad female)

```

In this example, "time" is the endogenous variable, and "phone" is the instrument.  We first run the first stage regression with "reghdfe", and get the control-function residual "u2h_fe".  Then we include "time" and "u2h_fe" in the second stage regression with "ppmlhdfe".  "ad" and "female" are two fixed effects.  The standard errors are incorrect, we need to use bootstrap to get the correct standard errors.

To verify our control function approach, we do the same with a linear model.  A word of caution on "it just works": for the *linear* model, inserting the first-stage residual is the standard 2SLS-equivalent control function.  For the *Poisson* model it is not automatic — residual inclusion is valid here under the specific structure Wooldridge (2010, ch. 18) assumes: a **continuous** endogenous variable, a **linear** first stage, and an exponential conditional mean with the control-function residual entering the index (an additive-error / Mundlak-type assumption).  Outside that structure (e.g. a discrete endogenous regressor, or a different error structure) plugging in a linear residual is not generally consistent, and even when it is, the second-stage standard errors must be bootstrapped because the residual is generated (as noted above).  There is a package "ivreghdfe" we use to verify our two step process.


```{stata}
*| label: stata2
*| echo: true
*| collectcode: true

clear all
webuse website
ivreghdfe visits frfam (time=phone), absorb(ad female)
* now the control function apparoach
reghdfe time phone frfam , absorb(ad female) resid
predict u2h_fe, resid
reghdfe visits time u2h_fe frfam , absorb(ad female)

```

The two-step control-function estimate of the `time` coefficient is 0.5643 with
a standard error of 0.2124, against a naive FE Poisson value of 0.0448 that is
not even significant ($p = 0.32$). We can see that "ivreghdfe" gives the same result as our two step process.  The standard errors are wrong.  Now we do bootstrapping.

## Bootstrapping standard errors

Suppose we need clustered standard errors on "ad".


```{stata}
*| label: stata3
*| echo: true
*| collectcode: true

* bootstrap, suppose we need to cluster by "ad"
* Note: bootstrap's idcluster(newid) below gives each resampled COPY of a
* cluster its own unique id -- necessary because a cluster can be drawn more
* than once within a replicate, and those duplicates must be treated as
* distinct groups. The regressions inside this program must absorb "newid",
* not the original "ad": absorbing "ad" would pool resampled duplicates of
* the same cluster back into a single fixed effect, which violates the
* cluster-bootstrap logic (verified empirically: within a replicate, the
* number of distinct "ad" values is consistently below the true cluster
* count, while "newid" always has exactly that many distinct values).
clear all
capture program drop ppmlhdfe_cf
 program ppmlhdfe_cf, rclass
  reghdfe time phone frfam , absorb(newid female) resid
  predict double u2h_fe, resid
  ppmlhdfe visits time u2h_fe frfam , absorb(newid female)
  return scalar b_time = _b[time]
  drop u2h_fe
  xtset, clear
 end

webuse website

xtset, clear
bootstrap r(b_time) , reps(100) seed(123) cluster(ad) idcluster(newid) : ppmlhdfe_cf

```

A troubleshooting note: we need to run `xtset, clear` before each regression inside the bootstrap. The reason is that Stata's cluster bootstrap generates a new id variable for the resampled clusters and leaves the panel `xtset` to that id; if it is not cleared, a later regression can pick up the stale panel setting even when it does not use the id. Clearing it each iteration avoids this.

## How to do it with R

R has similar packages https://cran.r-project.org/web/packages/fixest/vignettes/fixest_walkthrough.html.

We can use "fixest" and "fepois" to do the same thing as "reghdfe" and "ppmlhdfe".  Again, at the moment, we need to manually bootstrap the two stages for correct standard errors.

## What if the endogenous variable is binary?

Wooldridge treats this case only in a section on extensions and future directions, and declines to vouch
for the approximation, so it is the least settled material in this chapter.  Both procedures above assume
otherwise: Procedure 3's first stage is *linear*, and the companion probit procedure in the same
paper states outright that the endogenous variable is assumed continuous.  The binary case appears only
in that paper's final section, on extensions and future directions.

What is offered there is the **generalized residual** from a reduced-form probit in
$w_{it} = (1, z_{it}, \bar z_i)$ --- the instrument, the time-varying exogenous controls, and their unit
means:

$$ \widehat{gr}_{it2} = y_{it2}\,\lambda(w_{it}\hat\theta_2) - (1-y_{it2})\,\lambda(-w_{it}\hat\theta_2) $$ {#eq-poisson-iv-fe-1}

where $\lambda$ is the inverse Mills ratio.  For a **logit** first stage the generalized residual
collapses algebraically to the plain response residual $y_{it2} - \hat p_{it2}$, so the familiar
"predicted probability, then subtract" recipe is the generalized residual, not an approximation to it.

The reason for that collapse is worth knowing, because it tells you when the shortcut stops working.  The
generalized residual is the score of the log-likelihood with respect to the index.  For a logit that score
*is* $y_{it2} - \hat p_{it2}$.  For a probit it is the inverse Mills expression above, which is a
genuinely different variable: in a simulated sample of 5,000 the two differ by as much as 1.59, while
correlating at 0.999.  So the subtraction recipe is exact for a logit and simply wrong for a probit, and
the high correlation means the mistake will not announce itself.

One practical warning follows.  What you want is the **response** residual --- predict the probability,
then subtract --- and `predict` on its own returns a probability, not a residual.  The other residuals on
offer after a binary model are not the generalized residual either: Pearson residuals divide through by
$\sqrt{\hat p (1-\hat p)}$, and deviance residuals are a different transformation again.  In the same
simulation those correlate with the response residual at 0.991 and 0.995 --- close enough to look correct
if you glance at them, and the wrong variable to put in the second stage.

A note on the name, because it misleads.  This construction is usually filed under **correlated random
effects** (CRE), and that is how to find it in the literature --- see the [CRE chapter](more-cre.qmd).
The name describes the *assumption*: heterogeneity is modelled as a function of the group means and
allowed to correlate with the regressors, instead of being treated as a free parameter.  It does not
describe the estimator.  The first stage below is a **pooled logit** with the group means added as
regressors.  No random effect is estimated, no variance component appears, and nothing is integrated
out.  Below it is labelled "CRE (Mundlak device)" so that both names stay visible: the first tells you
where to look it up, the second tells you what is actually computed.

Two caveats should be stated plainly.  The construction is introduced as "a simple example", and is immediately followed
by: "The Mundlak device, or Chamberlain's version of it, might work reasonably well, but neither might be
flexible enough.  We leave investigations into the quality of CF approximations in discrete cases to
future research."  So this is the construction to use, but it is not a settled result.  If you are worried
the specification is too rigid, the same paper suggests adding nonlinear functions of the conditioning
set --- in particular interactions of the covariates with the unit means, and with the control function
itself --- or letting the variance depend on the unit means via a heteroskedastic probit.

### Several fixed effects in the first stage

The expression $w_{it} = (1, z_{it}, \bar z_i)$ above is written for a single unobserved effect.  If the
second stage absorbs more than one set of fixed effects --- units, years, and something else --- the
first stage needs the treatment set out in the [Mundlak chapter](mundlak-device.qmd), where the CRE
construction is the Mundlak device: take means over the
*one* dimension you are eliminating, of everything else in the model, retained dummies included.  In
practice that means unit means of the instrument and the time-varying controls, plus unit means of the
year and other dummies, with those dummies themselves kept in levels.  You do not want one set of means
per dimension; that construction is exact only for orthogonal designs, which observational panels do not
have.

### Why not just put the dummies in the first stage?

Because a logit with many thin groups is badly biased, and this is the whole reason the Mundlak route is
needed.  The size of the problem depends on observations per group, not on the number of groups.  In a
simulation with the total sample held near 12,000 and a true slope of 1, a logit with group dummies
returns about 2.0 at two observations per group --- matching the known result that the fixed-effects
logit slope is inflated by a factor of two in that case --- about 1.37 at four observations per group,
1.15 at six, 1.04 at twenty-five, and essentially the truth by a hundred.  A panel with a few thousand
units and four or five observations each therefore sits in the worst part of that range, while year
effects with hundreds of observations apiece are harmless and can simply be estimated.

The CRE (Mundlak device) first stage is better but not unbiased: in the same simulation it ran roughly ten per cent low at
small group sizes, because integrating a logit over normal heterogeneity does not give a logit.  That is
the same approximation error the previous section's caveat refers to, arriving from a different
direction.

### A worked example

The `website` data used earlier does not have the shape that causes trouble --- its fixed effects have 13
and 2 levels, with roughly 90 observations each, so a logit with dummies would be perfectly well behaved
there.  This simulates something more awkward instead: a count outcome, a binary endogenous regressor, an
instrument, and two thousand units with between two and six observations each.

```{stata}
*| label: binary_cf
*| echo: true
*| collectcode: true

preserve
clear
set seed 20260731
set obs 2000
gen i = _n
gen double ci = rnormal()*0.5
gen T = 2 + int(runiform()*5)        /* unbalanced: 2-6 observations per unit */
expand T
bysort i: gen t = _n

gen double z = rnormal()             /* the instrument */
gen double e = rnormal()
gen double u = 0.8*e + rnormal()*0.6 /* correlated with e, hence the endogeneity */
gen byte   d = (-0.2 + 0.9*z + ci + e > 0)
gen double mu = exp(0.3*d + ci + 0.5*u)
gen y = rpoisson(mu)                 /* true coefficient on d is 0.3 */

* 1. naive FE Poisson, ignoring the endogeneity
quietly ppmlhdfe y d, absorb(i)
scalar b_naive = _b[d]

* 2. control function with a CRE (Mundlak device) logit first stage: a POOLED
*    logit with the group mean of the instrument added.  No random effect is
*    estimated.  For a logit the generalized residual is just d - phat.
bysort i: egen double z_i = mean(z)
quietly logit d z z_i
predict double phat, pr
gen double gr = d - phat
quietly ppmlhdfe y d gr, absorb(i)
scalar b_cre = _b[d]

* 3. control function with a linear first stage, unit FE absorbed exactly
quietly reghdfe d z, absorb(i) resid
predict double v_lin, resid
quietly ppmlhdfe y d v_lin, absorb(i)
scalar b_lin = _b[d]

di "true coefficient on d           = 0.300000"
di "naive FE Poisson                = " %8.6f b_naive
di "CF, CRE (Mundlak) logit         = " %8.6f b_cre
di "CF, linear FE first stage       = " %8.6f b_lin
restore
```

Against a true coefficient of 0.300000, the naive FE Poisson gives 0.747206,
the CRE/Mundlak logit first stage gives 0.293829, and the linear FE first stage
gives 0.277644.

The naive estimate is inflated by a factor of about two and a half; both control-function versions
recover something close to the truth, and they agree with each other — within 0.016 of one another, and both within 0.023 of the truth.  That agreement is the point.  When
the endogenous variable is binary I would report both: the linear first stage rests on the exact result
in point 2 above and drops no observations, while the logit version is better specified for a binary
variable but rests on the unsettled argument in this section.  If they diverge materially, that is the
most important thing you have learned, and it belongs in the write-up.

The standard errors in the chunk above are wrong, for the reason given earlier --- the residual is
generated.  Bootstrap both stages together, as in the earlier section.

---

<!-- see-also-footer -->
*Systematic treatment: [R](https://xiangao.github.io/causal_econometrics_guide/poisson-iv.html) · [Julia](https://xiangao.github.io/causal_econometrics_julia/poisson-iv.html).*
