---
title: "Instrumental Variables in Fixed-Effects Poisson Models"
date: "2024-08-29"
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE)
library(Statamarkdown)
stataexe <- find_stata()
knitr::opts_chunk$set(engine.path = list(stata = stataexe))
```

## The Problem: Endogeneity with Exponential Conditional Means

In panel count data models, the standard conditional mean specification is:

$$E[y_{it} \mid x_{it}, c_i] = c_i \exp(x_{it} \beta)$$

When an explanatory variable in $x_{it}$ is endogenous, standard linear Two-Stage Least Squares (2SLS) cannot be applied because the conditional mean is exponential rather than additive. Meanwhile, nonlinear IV estimators (such as `ivpoisson` in Stata) cannot absorb high-dimensional individual and time fixed effects.

To estimate an endogenous Poisson model with multiple fixed effects, Lin and Wooldridge (2019) establish a two-step **control function** approach.

## The Two-Step Control Function Recipe

For a continuous endogenous regressor $w_{it}$ and excluded instruments $z_{it}$:

1. **First Stage (Reduced Form):** Regress the endogenous variable on the instruments and exogenous controls, absorbing all fixed effects:
   $$w_{it} = z_{it}\pi + x_{it1}\gamma + c_i + f_t + v_{it}$$
   Obtain the OLS residual $\hat{v}_{it}$. In Stata, use `reghdfe ..., resid`; in R, use `fixest::feols()`.

2. **Second Stage (Structural Poisson):** Estimate the fixed-effects Poisson model including both $w_{it}$ and the first-stage residual $\hat{v}_{it}$, absorbing the same fixed effects:
   $$E[y_{it} \mid w_{it}, x_{it1}, c_i, f_t, \hat{v}_{it}] = c_i f_t \exp(w_{it}\beta_1 + x_{it1}\beta_2 + \rho \hat{v}_{it})$$
   In Stata, use `ppmlhdfe`; in R, use `fixest::fepois()`.

3. **Inference & Testing:** 
   - A $t$-test of $H_0: \rho = 0$ serves as a direct test of exogeneity.
   - Because $\hat{v}_{it}$ is a generated regressor, reported second-stage standard errors are invalid. Standard errors must be computed via a panel/cluster bootstrap re-estimating both stages in each replication.

## Continuous Endogenous Variable: Application in Stata

Using Stata's `website.dta` ($N = 500$):
- Outcome: `visits` (count)
- Endogenous regressor: `time` (minutes on site)
- Instrument: `phone` (support calls)
- Exogenous control: `frfam` (referrals)
- Fixed effects: `ad` (13 campaigns) and `female` (gender)

```{stata}
*| label: stata1
*| echo: true
*| collectcode: true

webuse website, clear

* Step 1: Linear first stage absorbing both fixed effects
reghdfe time phone frfam, absorb(ad female) resid
predict double u2h_fe, resid

* Step 2: FE Poisson with control function residual
ppmlhdfe visits time u2h_fe frfam, absorb(ad female)
```

The coefficient on `time` is $0.0448$ ($p = 0.318$). The residual coefficient is $\hat{\rho} = 0.0765$ ($p = 0.096$), indicating weak positive selection.

### Comparison with Linear Models

In a linear model, the control function estimator is algebraically identical to 2SLS. Comparing linear specifications verifies that the two stages are wired correctly:

```{stata}
*| label: stata2
*| echo: true
*| collectcode: true

clear all
webuse website, clear

* 1. Linear 2SLS
ivreghdfe visits frfam (time=phone), absorb(ad female)

* 2. Linear control function
reghdfe time phone frfam, absorb(ad female) resid
predict double u2h_fe, resid
reghdfe visits time u2h_fe frfam, absorb(ad female)

* 3. Naive models (ignoring endogeneity)
reghdfe visits time frfam, absorb(ad female)
ppmlhdfe visits time frfam, absorb(ad female)
```

| Model | Estimator | `time` Coef. | Std. Err. | $p$-value |
|:---|:---|:---:|:---:|:---:|
| Naive Linear FE | `reghdfe` | 0.7402 | 0.0401 | $<0.001$ |
| Linear 2SLS | `ivreghdfe` | 0.5643 | 0.2124 | 0.008 |
| Linear Control Function | Manual two-step | 0.5643 | 0.2083 | 0.007 |
| Naive FE Poisson | `ppmlhdfe` (no residual) | 0.1188 | 0.0096 | $<0.001$ |
| Control-Function FE Poisson | `ppmlhdfe` (with $\hat{v}$) | 0.0448 | 0.0449 | 0.318 |

In linear models, manual control function reproduces 2SLS point estimates exactly. In the Poisson model, addressing endogeneity drops the estimated effect of `time` from $0.1188$ to $0.0448$.

### Cluster Bootstrap in Stata

Because clusters may be sampled multiple times during replacement, each resampled cluster copy must receive a unique identifier via `idcluster()` to prevent duplicates from being pooled into a single fixed effect:

```{stata}
*| label: stata3
*| echo: true
*| collectcode: true

clear all
capture program drop ppmlhdfe_cf
program ppmlhdfe_cf, rclass
  reghdfe time phone frfam, absorb(newid female) resid
  predict double u2h_fe, resid
  ppmlhdfe visits time u2h_fe frfam, absorb(newid female)
  return scalar b_time = _b[time]
  return scalar b_u2h  = _b[u2h_fe]
  drop u2h_fe
  xtset, clear
end

webuse website, clear
xtset, clear
bootstrap r(b_time) r(b_u2h), reps(1000) seed(123) cluster(ad) idcluster(newid) nodots: ppmlhdfe_cf
```

The bootstrap standard error on `time` is $0.0477$ (compared to the unadjusted analytic standard error of $0.0449$).

## Implementation in R: `fixest`

In R, the same workflow is implemented using `fixest::feols()` and `fixest::fepois()`. Note that `feols()` drops singleton clusters (here `ad == 10`), so the residual vector must be aligned using `obs()`:

```{r}
#| label: r-demo
#| echo: true
#| message: false
#| warning: false

library(fixest)
library(haven)

df <- as.data.frame(read_dta("https://www.stata-press.com/data/r18/website.dta"))
df$ad     <- as.factor(df$ad)
df$female <- as.factor(df$female)

# Step 1: Linear first stage
fs <- feols(time ~ phone + frfam | ad + female, data = df)

# Align rows (feols drops singletons)
d <- df[obs(fs), ]
d$u2h_fe <- residuals(fs)

# Step 2: FE Poisson with and without control function
m_naive <- fepois(visits ~ time + frfam          | ad + female, data = d, notes = FALSE)
m_cf    <- fepois(visits ~ time + u2h_fe + frfam | ad + female, data = d, notes = FALSE)

etable(m_naive, m_cf, headers = c("Naive FE Poisson", "Control Function"), se.below = TRUE)
```

The point estimates match Stata to seven digits: $0.1187612$ (naive) and $0.0448103$ (control function).

### Cluster Bootstrap in R

```{r}
#| label: r-bootstrap
#| echo: true
#| message: false
#| warning: false

set.seed(123)
B <- 1000
clusters   <- unique(d$ad)
boot_coefs <- matrix(NA_real_, nrow = B, ncol = 2, dimnames = list(NULL, c("time", "resid")))

for (b in seq_len(B)) {
  drawn <- sample(clusters, length(clusters), replace = TRUE)
  boot_data <- do.call(rbind, lapply(seq_along(drawn), function(k) {
    sub <- d[d$ad == drawn[k], ]
    sub$boot_id <- k
    sub
  }))
  boot_data$boot_id <- as.factor(boot_data$boot_id)

  fs_b <- feols(time ~ phone + frfam | boot_id + female, data = boot_data, notes = FALSE)
  bb   <- boot_data[obs(fs_b), ]
  bb$u_b <- residuals(fs_b)

  m_b <- try(fepois(visits ~ time + u_b + frfam | boot_id + female, data = bb, notes = FALSE), silent = TRUE)
  if (!inherits(m_b, "try-error")) boot_coefs[b, ] <- coef(m_b)[c("time", "u_b")]
}

cat("Completed replications :", sum(complete.cases(boot_coefs)), "\n")
cat("Bootstrap SE, time     :", round(sd(boot_coefs[, "time"],  na.rm = TRUE), 4), "\n")
cat("Bootstrap SE, residual :", round(sd(boot_coefs[, "resid"], na.rm = TRUE), 4), "\n")
```

The R bootstrap yields $\text{SE} = 0.0450$. With only 12 clusters in `ad`, small differences across software packages ($0.0477$ in Stata vs. $0.0450$ in R) reflect finite-sample sensitivity in how rare degenerate resamples and singletons are handled.

## Extension: Binary Endogenous Regressor

When the endogenous regressor $D_{it}$ is binary (e.g., participation in a program), a linear first stage is technically misspecified because the error cannot satisfy conditional independence.

### The CRE Logit Control Function

1. **Avoid Dummy Variable Incidental Parameter Trap:** Estimating unit dummies in a first-stage logit with small $T$ causes severe bias (e.g., slopes inflated by $2\times$ when $T = 2$).
2. **Correlated Random Effects (Mundlak Device):** Instead of unit dummies, include unit-level averages of time-varying covariates ($\bar{z}_i$) in a pooled logit:
   $$P(D_{it} = 1 \mid z_{it}, \bar{z}_i) = \Lambda(z_{it}\gamma + \bar{z}_i \pi)$$
3. **Generalized Residual:** For a logit model, the generalized residual simplifies directly to the response residual:
   $$\hat{r}_{it} = D_{it} - \hat{p}_{it}$$
4. **Second Stage:** Include $\hat{r}_{it}$ directly in the fixed-effects Poisson model.

### Simulation Demonstration

We simulate $N = 2,000$ units with unbalanced panels ($T_i \in [2, 6]$), a binary endogenous treatment $D$, and true treatment parameter $\beta = 0.30$:

```{stata}
*| label: binary_cf
*| echo: true
*| collectcode: true

preserve
clear
set seed 20260731
set obs 2000
gen i = _n
gen double ci = rnormal()*0.5
gen T = 2 + int(runiform()*5)        /* 2-6 observations per unit */
expand T
bysort i: gen t = _n

gen double z = rnormal()             /* instrument */
gen double e = rnormal()
gen double u = 0.8*e + rnormal()*0.6 /* correlated with e -> endogeneity */
gen byte   d = (-0.2 + 0.9*z + ci + e > 0)
gen double mu = exp(0.3*d + ci + 0.5*u)
gen y = rpoisson(mu)                 /* true beta = 0.3 */

* 1. Naive FE Poisson (ignores endogeneity)
quietly ppmlhdfe y d, absorb(i)
scalar b_naive = _b[d]

* 2. CF with CRE (Mundlak) logit first stage
bysort i: egen double z_i = mean(z)
quietly logit d z z_i
predict double phat, pr
gen double gr = d - phat
quietly ppmlhdfe y d gr, absorb(i)
scalar b_cre = _b[d]

* 3. CF with linear first stage
quietly reghdfe d z, absorb(i) resid
predict double v_lin, resid
quietly ppmlhdfe y d v_lin, absorb(i)
scalar b_lin = _b[d]

di "True coefficient on d       = 0.300000"
di "Naive FE Poisson            = " %8.6f b_naive
di "CF, CRE (Mundlak) Logit     = " %8.6f b_cre
di "CF, Linear First Stage      = " %8.6f b_lin
restore
```

Against the true $\beta = 0.30$:
- Naive FE Poisson estimates $0.747$ (biased by over $2.4\times$).
- The CRE logit control function recovers $0.294$.
- The linear first stage control function recovers $0.278$.

In applied practice, reporting both the CRE logit and linear first-stage control functions provides a robust sensitivity check.

## Summary

1. **Mechanism:** Control function augments the second-stage exponential regression with the first-stage reduced-form residual, effectively absorbing endogeneity.
2. **Commands:** Use `reghdfe` + `ppmlhdfe` in Stata, or `fixest::feols()` + `fixest::fepois()` in R.
3. **Inference:** Standard errors must be cluster-bootstrapped over both stages; analytic second-stage standard errors ignore first-stage estimation uncertainty.
4. **Binary Regressors:** Use a CRE logit first stage with group means $\bar{z}_i$ and add the generalized residual $\hat{r} = d - \hat{p}$ to avoid the incidental parameters problem of unit dummies.
