---
title: "Poisson model with exposure"
date: "2024-04-20"
---

## Poisson with exposure or Poisson rate

Poisson models for count data often arise when units are observed for different
lengths of time. A hospital admission count for a patient followed for 30 days
is not directly comparable to a count for a patient followed for 365 days. The
natural quantity of interest is the *rate* — admissions per unit time — not
the raw count.

There are three related but distinct specifications, each making a different
assumption:

**Model 1 — Poisson with offset (rate model):**

$$\log\!\bigl(E[Y]/t\bigr) = \beta_0 + \beta_1 X$$ {#eq-poisson-exposure-1}

Equivalently, $\log E[Y] = \beta_0 + \beta_1 X + \log t$, where $\log t$ is
included as a *fixed* offset (coefficient constrained to 1). This is the
standard exposure model: the expected count is proportional to exposure time.

**Model 2 — Poisson on the ratio $Y/t$:**

$$\log\!\bigl(E[Y/t]\bigr) = \beta_0 + \beta_1 X$$ {#eq-poisson-exposure-2}

This looks similar, but it is not equivalent to the offset model, and strictly
speaking $Y/t$ is **not a Poisson outcome at all**: a Poisson likelihood is
defined for non-negative integer counts, and the continuous ratio $Y/t$ is not
one. Because $t$ is a fixed constant for each unit, $E[Y/t] = E[Y]/t$ exactly
(linearity of expectation), so the *conditional mean* is still log-linear in
$X$; that is why the point estimates are often similar. But the Poisson variance
assumption fails: $\mathrm{Var}(Y/t) = \lambda/t^2$, whereas the mean is
$E[Y/t] = \lambda/t$, so the variance is $1/t$ times the mean rather than equal to
it. (Writing it as $\lambda/t$ would make variance equal mean and quietly restore
equidispersion, which is precisely what fails here.) So fitting `glm(rate ~ x, family = poisson)` is a quasi-likelihood
*mean model* rather than a correct Poisson likelihood. If you use it, treat it
as a workaround: weight by exposure (`weights = t`) so the implied variance is
right, and/or use robust standard errors. The offset model (Model 1) avoids the
issue entirely by keeping the integer count as the outcome.

**Model 3 — Unconstrained log-exposure:**

$$\log E[Y] = \beta_0 + \beta_1 X + \gamma \log t$$ {#eq-poisson-exposure-3}

Here $\gamma$ is estimated from the data. If $\hat\gamma \approx 1$, the
proportionality assumption is supported. If $\hat\gamma < 1$, counts grow
sub-linearly with exposure — a testable, substantive claim.

## Simulation

```{r}
#| label: sim-data
#| cache: true
#| warning: false
#| message: false

library(tidyverse)
set.seed(42)

n <- 800

# Simulate a health study: count of adverse events
# Covariate: age group (0 = younger, 1 = older)
age_old <- rbinom(n, 1, 0.45)

# Exposure: days under observation (varies 7–365)
exposure <- sample(seq(7, 365, by = 7), n, replace = TRUE)

# True model: rate = exp(-3 + 1.2 * age_old)
# Count ~ Poisson(rate * exposure)
true_rate <- exp(-3 + 1.2 * age_old)
count <- rpois(n, lambda = true_rate * exposure)

df <- data.frame(count, age_old, exposure, log_exposure = log(exposure))
summary(df[, c("count", "exposure")])
```

We simulate $n = 800$ patients. Age group is
$\text{age\_old} \sim \text{Bernoulli}(0.45)$, and exposure is days under
observation, drawn uniformly from the weekly grid 7 to 364 and independently of
age. The event rate is $\exp(-3 + 1.2\,\text{age\_old})$ per day and the count
is $\text{Poisson}(\text{rate} \times \text{exposure})$, so the true intercept
is $-3$, the true age effect is 1.2, and the true exposure elasticity is
$\gamma = 1$ by construction.

The data has a wide range of exposure times --- 7 to 364 days, median 196, with
counts from 0 to 71 and a mean of 19 --- so ignoring them would badly
confound the comparison between age groups.

## Fitting the three models

```{r}
#| label: fit-models
#| cache: true
#| warning: false

# Model 1: offset — coefficient on log(exposure) constrained to 1
m1 <- glm(count ~ age_old + offset(log(exposure)),
          family = poisson, data = df)

# Model 2: Poisson on the rate Y/t (different response)
df$rate <- df$count / df$exposure
m2 <- glm(rate ~ age_old,
          family = poisson, data = df)

# Model 3: log(exposure) as a free covariate
m3 <- glm(count ~ age_old + log_exposure,
          family = poisson, data = df)

# Collect coefficients
results <- data.frame(
  Model = c("1. Offset (constrained γ=1)",
            "2. Rate Y/t as response",
            "3. Unconstrained log(t)"),
  Intercept = c(coef(m1)[1], coef(m2)[1], coef(m3)[1]),
  age_old   = c(coef(m1)[2], coef(m2)[2], coef(m3)[2]),
  gamma_log_t = c(1, NA, coef(m3)[3])
)

knitr::kable(results, digits = 3,
  caption = "Coefficient estimates: true intercept = −3, true age effect = 1.2, true γ = 1")
```

All three recover the truth. The offset model gives an intercept of $-3.009$ and
an age effect of 1.212 against true values of $-3$ and 1.2; the rate model gives
$-2.980$ and 1.176; the unconstrained model gives $-2.801$, 1.212 and
$\hat\gamma = 0.962$ against a true $\gamma$ of 1.

**Reading the table:** Model 1 and Model 3 give nearly identical age-group
coefficients because the true $\gamma = 1$ (counts are indeed proportional to
exposure time). Model 2 differs slightly in the intercept because the response
variable changes. The estimated $\hat\gamma$ from Model 3 should be close to 1.

## When $\gamma \ne 1$

To see why the unconstrained model matters, repeat the simulation with a
sub-linear count process ($\gamma = 0.6$: an initial burst of events that
doesn't grow proportionally with long follow-up):

```{r}
#| label: nonlinear-exposure
#| cache: true
#| warning: false

count2 <- rpois(n, lambda = true_rate * exposure^0.6)
df2 <- data.frame(count2, age_old, exposure, log_exposure = log(exposure))

m1b <- glm(count2 ~ age_old + offset(log(exposure)),
           family = poisson, data = df2)
m3b <- glm(count2 ~ age_old + log_exposure,
           family = poisson, data = df2)

cat("Offset model age coefficient:      ", round(coef(m1b)[2], 3), "\n")
cat("Unconstrained model age coefficient:", round(coef(m3b)[2], 3), "\n")
cat("Estimated gamma (log-exposure):    ", round(coef(m3b)[3], 3),
    " (true = 0.6)\n")
```

Two things to read off this, and the first is not what one might expect.

The offset model's age coefficient (1.171) and the unconstrained model's (1.168)
are essentially the same, and both sit just below the true 1.2. Forcing
$\gamma = 1$ when the truth is 0.6 does **not** bias the treatment coefficient
here --- and the reason is structural: `exposure` is drawn by `sample(...)`
independently of `age_old`, so the misspecified exposure elasticity is absorbed by
the intercept and cannot contaminate a coefficient on a covariate orthogonal to
it. The constraint would bite if exposure were *correlated* with the covariate of
interest --- for instance if older patients were systematically followed longer,
which is common in real cohort data and is the case worth worrying about.

What the unconstrained model does buy, unambiguously, is $\gamma$ itself: it
recovers $\hat\gamma = 0.566$ against a truth of 0.6, whereas the offset model
cannot report it at all because it fixes it at 1. So Model 3's role is diagnostic
--- it tells you whether the proportionality assumption holds --- rather than
protective of the treatment estimate.

## Stata equivalent

```stata
* Model 1: offset
poisson count age_old, exposure(exposure)

* Model 2: rate as response
gen rate = count / exposure
* Stata will warn that the dependent variable is non-integer (poisson expects
* integer counts by default) -- this is expected and can be ignored here,
* since Poisson QMLE is valid for continuous non-negative outcomes too.
poisson rate age_old

* Model 3: unconstrained log(t)
gen log_exposure = log(exposure)
poisson count age_old log_exposure
```

## Which model to use?

| Situation | Recommended model |
|---|---|
| Counts grow proportionally with time | Model 1 (offset) |
| Uncertain about proportionality | Model 3 (unconstrained), test $\hat\gamma = 1$ |
| Rate is the natural estimand | Model 1 (offset) — it estimates a rate directly; Model 2 only as a weighted/robust quasi-model |
| Long follow-up with declining event rate | Model 3 or negative binomial with offset |

In most health and social-science applications, Model 1 (Poisson with offset)
is the default. Model 3 is the diagnostic check: fit it, look at $\hat\gamma$,
and return to Model 1 if the constraint is supported by the data.

---

<!-- see-also-footer -->
*Systematic treatment: [R](https://xiangao.github.io/causal_econometrics_guide/poisson-iv.html) · [Julia](https://xiangao.github.io/causal_econometrics_julia/poisson-iv.html).*
