---
title: "What model to use for rare events"
date: "2017-10-26"
---


## Introduction

In empirical studies, people are worried about rare event situation.  That is, when you have, for example, lots of 0's and only a few 1's, or vice versa.  Do you run a logit model, or do you use a "rare event logit"?  When should you use either approach?  Or there is a third approach?

Paul Allison said in his blog (https://statisticalhorizons.com/logistic-regression-for-rare-events):

"Prompted by a 2001 article by King and Zeng, many researchers worry about whether they can legitimately use conventional logistic regression for data in which events are rare. Although King and Zeng accurately described the problem and proposed an appropriate solution, there are still a lot of misconceptions about this issue.

The problem is not specifically the rarity of events, but rather the possibility of a small number of cases on the rarer of the two outcomes.  If you have a sample size of 1000 but only 20 events, you have a problem. If you have a sample size of 10,000 with 200 events, you may be OK. If your sample has 100,000 cases with 2000 events, you're golden."

In general I agree with him.  However, when exactly should we use King and Zeng's "relogit"?

Allison also mentioned two other methods.  One is called the Firth method, a penalized likelihood approach.  The other one is the exact logistic regression, which is for small samples.

In this simulation exercise, I compare logit against bias-reduced logistic
regression (`brglm2` with `method="brglmFit"`, `type="AS_mean"`) and the Firth
model (`logistf`).

One thing to be clear about before reading any comparison of the latter two: for a
logistic regression **they are the same estimator.** Firth's penalized likelihood
(penalizing by the Jeffreys prior) coincides with mean bias reduction for GLMs
with a canonical link, and logit is canonical for the binomial. So `brglm2`'s
`AS_mean` and `logistf` solve the same estimating equations and differ only in
numerical tolerance and in how each package builds standard errors and intervals
(`logistf` uses penalized profile likelihood). A demonstration:

```{r}
#| label: firth-equivalence
#| message: false
#| warning: false

library(brglm2); library(logistf)
set.seed(11)
n <- 500; x <- rnorm(n); y <- rbinom(n, 1, plogis(-3.5 + 0.8 * x))
d <- data.frame(y, x)

rbind(
  `plain logit`     = coef(glm(y ~ x, data = d, family = binomial)),
  `brglm2 AS_mean`  = coef(glm(y ~ x, data = d, family = binomial,
                               method = "brglmFit", type = "AS_mean")),
  `logistf (Firth)` = coef(logistf(y ~ x, data = d))
)
```

They agree to about $10^{-7}$. This matters for reading the findings below: what
the simulation can tell us is how **logit** compares with **Firth-type bias
reduction**, not which of two implementations of the same estimator is better. If
a genuinely distinct third method is wanted, `brglm2`'s `type = "AS_median"`
(median bias reduction) is one, and King and Zeng's `relogit` correction is
another.

## Simulation

Here I have some code for using multiple cores to run these three models.  The bias-reduced logistic regression is implemented in R via the `brglm2` package, the Firth method is implemented in `logistf`.

```{r}
#| label: rare-events-sim
#| eval: false

library(brglm2)
library(logistf)
require(snowfall)
set.seed(666)

# initialize parallel cores.
sfInit(parallel=TRUE, cpus=16)

gen.sim <- function(df){
  x <- rnorm(df['nobs'], 0, 1)

  # generate binary data
  p <- df['p']
  alpha <- -log((1-p)/p)
  z = alpha + 2*x
  pr = 1/(1+exp(-z))
  y = rbinom(df['nobs'], 1, pr)
  df = data.frame(y=y, x=x)

  # With small nobs and small p (e.g. nobs=10, p=.01), y can easily come out
  # all-zero (>90% chance in that cell), which makes glm/logistf unable to
  # estimate a coefficient on x. Guard against that rather than letting a
  # single degenerate draw crash the whole parallel simulation loop.
  if (var(y) == 0) {
    return(c(logit=NA, brglm2=NA, logistf=NA))
  }

  # logit
  m1 <- tryCatch(glm(y ~ x, family='binomial'), error = function(e) NULL)
  m1.x <- if (!is.null(m1)) summary(m1)$coefficients['x','Estimate'] - 2 else NA

  # bias-reduced logistic regression (brglm2)
  m2 <- tryCatch(
    glm(y ~ x, family = binomial(link = "logit"), data = df,
        method = "brglmFit", type = "AS_mean"),
    error = function(e) NULL
  )
  m2.x <- if (!is.null(m2)) coef(m2)['x'] - 2 else NA

  # logistf (Firth)
  m3 <- tryCatch(logistf(y ~ x, data=df), error = function(e) NULL)
  m3.x <- if (!is.null(m3)) coef(m3)['x'] - 2 else NA

  return(c(logit=m1.x, brglm2=m2.x, logistf=m3.x))
}

# set parameter space
sim.grid = seq(1, 100, 1)
p.grid = c(.01, .05, .1)
nobs.grid = c(10, 30, 50, 100, 200, 500, 1000, 10000)

data.grid <- expand.grid(nobs.grid, sim.grid, p.grid)
names(data.grid) <- c('nobs', 'nsim', 'p')

# export functions and libraries to parallel workers
sfExport(list=list("gen.sim"))
sfLibrary(brglm2)
sfLibrary(logistf)

results <- data.frame(t(sfApply(data.grid, 1, gen.sim)))

# stop the cluster
sfStop()

forshiny <- cbind(data.grid, results)
write.csv(forshiny, 'results.csv')
```

We simulate 100 times with sample size from 10 to 10000, event probability .01, .05, and .1.

Since there are many simulations, we used the `snowfall` library to speed things up.

(The original post plotted bias and MSE by sample size and event probability from the `results.csv` output above; those figures are not reproduced here — the findings are summarized in prose below.)

In the case of small sample and rare event (for example, any situation that the product of sample size and probability is less than 5), none of these three models perform well.  This is understandable, after all, the rarer of the two groups has only less than 5 observations.  When the product is more than 50, there is not much difference between these three models.  For the situations in between, that is, the product of sample size and probability is greater than 5 but less than 50, we found that Firth-type bias reduction --- whether computed by `brglm2` or by `logistf` --- performs better than plain logit.  Any apparent ranking *between* those two should be discounted, since as shown above they are the same estimator; differences at that level reflect numerical tolerance or differing standard-error conventions, not a substantive advantage.

In the small sample situation, maybe it's better to use the exact logistic regression.
