---
title: "Sensitivity analysis of OLS with sensemakr"
date: "2025-04-03"
---


## OVB

Consider a true model,

$$ Y = \delta D + X \beta + \gamma Z + \epsilon $$ {#eq-sensitivity-analysis-1}

If $Z$ is unobserved we estimate

$$ Y = \delta D + X \beta + u $$ {#eq-sensitivity-analysis-2}

and face omitted variable bias (OVB). Cinelli and Hazlett (2020) show that

$$ |\widehat{\text{bias}}| = \sqrt{\frac{R^2_{Y \sim Z|D,X} R^2_{D \sim Z|X}}{1-R^2_{D \sim Z|X}}} (\frac{\hat \sigma_{y.dx}}{\hat \sigma_{d.x}})$$ {#eq-sensitivity-analysis-3}

Bias is large only when the omitted variable is strongly associated with both $Y$ and $D$. The Cinelli and Hazlett framework yields three key quantities.

## Partial $R^2$ of treatment on outcome

This measures how much variation in the outcome the treatment explains after partialling out $X$. In the extreme case where the confounder explains all remaining outcome variance ($R^2_{Y \sim Z|D,X} = 1$), the treatment effect goes to zero only if $R^2_{D \sim Z|X} \ge R^2_{Y \sim D|X}$. In other words, the confounder's association with treatment must exceed the treatment's partial association with the outcome.

## Robust value

$RV_{q,\alpha}$ is the minimum strength of association the confounder must have with both treatment and outcome for a $(1-\alpha)$ confidence interval to include a $q\%$ change from the current estimate. With $q=1$ this gives the threshold for the effect to include zero. Both the $Z$-$D$ and $Z$-$Y$ associations must exceed this threshold simultaneously.


## Bounds using observed covariate(s)

The first two quantities are dimensionless but hard to interpret without context. Benchmarking against an observed covariate helps. If a covariate $X_j$ is known to be an important predictor, we can ask how much stronger the confounder must be relative to $X_j$ to drive the effect to zero.

$$ k_D := \frac{R^2_{D \sim Z|X_{-j}}}{R^2_{D \sim X_j|X_{-j}}} $$ {#eq-sensitivity-analysis-4}

where $X_{-j}$ represents all covariates other than the $j$th we pick as the benchmark covariate.

$$ k_Y := \frac{R^2_{Y \sim Z|D,X_{-j}}}{R^2_{Y \sim X_j|D,X_{-j}}} $$ {#eq-sensitivity-analysis-5}


## Example

Here is an example from sensemakr documentation.

```{r}

# loads package
library(sensemakr)

# loads dataset
data("darfur")

# runs regression model
model <- lm(peacefactor ~ directlyharmed + age + farmer_dar + herder_dar +
                         pastvoted + hhsize_darfur + female + village, data = darfur)

# runs sensemakr for sensitivity analysis
sensitivity <- sensemakr(model = model, 
                         treatment = "directlyharmed",
                         benchmark_covariates = "female",
                         kd = 1:3)

# short description of results
sensitivity

# long description of results
summary(sensitivity)

```
```{r}
#| results: 'asis'

# minimal reporting
ovb_minimal_reporting(sensitivity, format = "html")

```

Let's see how to interpret this.  

$RV_{q=1} = 13.9 \%$.  This means that unobserved confounders that explain 13.9% of the residual variance both of the treatment and of the outcome are sufficiently strong to explain away all the observed effect. On the other hand, unobserved confounders that do not explain at least 13.9% of the residual variance both of the treatment and of the outcome are not sufficiently strong to do so.    

The robustness value for testing the null hypothesis that the coefficient of "directlyharmed" is zero ($RV_{q=1,\alpha=0.05}$) is 7.6%. This means that unobserved confounders that explain 7.6% of the residual variance both of the treatment and of the outcome are sufficiently strong to bring the lower bound of the confidence interval to zero (at the chosen significance level of 5%).
 
 
The partial $R^2$ of `directlyharmed` with `peacefactor` is 2.2%. This means that, in an extreme scenario in which we assume that unobserved confounders explain all of the left-out variance of the outcome, these unobserved confounders would need to explain at least 2.2% of the residual variance of the treatment to fully explain away the observed effect.

The above numbers are for the absolute strength the confounder needs to be to explain away the treatment effect.   The authors believe in this context, "female" is an important factor.  Suppose the confounder is as important as "female", what the partial $R^2$ would be?  It is shown that $R^2_{Y \sim Z|X,D} = 12.5 \%$ and $R^2_{D \sim Z|X} = 0.9 \%$, so a confounder as strong as "female" is not strong enough to explain away the treatment effect.

It is worth being careful about *why*, because comparing each number to $RV_{q=1} = 13.9\%$ one at a time invites the wrong reading. The $RV$ is defined for a confounder of *equal* strength on both sides, so it is not a threshold that each partial $R^2$ must separately clear. Here the outcome-side association (12.5%) is in fact close to the $RV$, which on its own might look like a near miss --- but the treatment-side association is only 0.9%, more than an order of magnitude short. It is the treatment side that does the work: "female" predicts the outcome nearly well enough to matter, and barely predicts the treatment at all. The contour plot below is the honest check, since it evaluates the pair jointly rather than either coordinate alone.  


```{r}

# plot bias contour of point estimate
plot(sensitivity)

```

This plot shows that even 3 times as strong as "female", the confounder is not strong enough to explain away the treatment effect.  

## sensitivity analysis for IV

Even with an IV model, we need sensitivity analysis.  The worry would be there could be an unobserved confounder that affects both the instrument and the outcome.  

Cinelli and Hazlett have a package "iv.sensemakr" for this.  The usage is similar to "sensemakr".

The example is Card's returns-to-schooling data, 3,010 men from the 1976 NLS.
The outcome is log wage, the treatment is years of education, and the instrument
is `nearc4`, an indicator for growing up near a four-year college. The covariates
are experience and its square, race, region, and urban indicators. The concern
the sensitivity analysis addresses is a confounder affecting both proximity to a
college and wages.

```{r}

# loads package
library(iv.sensemakr)

# loads dataset
data("card")

# prepares data
y <- card$lwage  # outcome
d <- card$educ   # treatment
z <- card$nearc4 # instrument
x <- model.matrix( ~ exper + expersq + black + south + smsa + reg661 + reg662 +
                     reg663 + reg664 + reg665+ reg666 + reg667 + reg668 + smsa66,
                   data = card) # covariates
# fits IV model
card.fit <- iv_fit(y,d,z,x)

# see results
card.fit

# runs sensitivity analysis
card.sens <- sensemakr(card.fit, benchmark_covariates = c("black", "smsa"))

# see results
card.sens

summary(card.sens)

```

The basic idea is to run an IV regression, then apply sensemakr to that model.  In this case, the authors use "black" and "smsa" as benchmark covariates.  The authors believe these are important factors.

The IV estimate is 0.132 with $t = 2.33$, $p = 0.02$, and an Anderson-Rubin
confidence interval of $[0.025, 0.285]$ that excludes zero. Read the sensitivity
statistics next to that, and the picture is not reassuring. The Robustness Value
is 0.00667 — two-thirds of one percent. Compare the OLS example earlier in this
chapter, where $RV_{q=1}$ was 13.9%. This IV result is roughly twenty times more
fragile than that one.

The benchmark bounds make it concrete, and they point the other way from what one
might hope. A confounder as strong as `black` explains 0.22% of the residual
variance of the instrument and 7.5% of the outcome, and under it the adjusted
interval becomes $[-0.021, 0.402]$. For `smsa` the figures are 0.64% and 2.0%,
giving $[-0.019, 0.396]$. **Both intervals contain zero.** A confounder no
stronger than either benchmark covariate is enough to remove the significance of
the estimate.

That `smsa` number is worth a second look on its own terms: its instrument-side
partial $R^2$ of 0.00639 sits just under the Robustness Value of 0.00667, so a
confounder of that strength lands essentially exactly on the tipping point. The
two diagnostics agree.

None of this says the Card estimate is wrong. It says the identification rests on
the instrument being cleaner than `black` or `smsa` are as predictors, which is
an assumption about college proximity that the data cannot check. We can also plot a contour plot to see how strong the confounder needs to be to explain away the treatment effect.

```{r}
#| results: 'asis'

plot(card.sens, parm = "rf", sensitivity.of = "t-value", lim = 0.08, kz = 1, ky = 1)
```

Unfortunately the package currently only allows for one instrument.  Should not be too hard to modify it to be more general.

---

<!-- see-also-footer -->
*Systematic treatment: [R](https://xiangao.github.io/causal_econometrics_guide/sensitivity-analysis.html) · [Julia](https://xiangao.github.io/causal_econometrics_julia/sensitivity-analysis.html).*
