---
title: "Sensitivity analysis of OLS with sensemakr"
date: "2025-04-03"
---


# OVB

Suppose we have a true model,

$$ Y = \delta D + X \beta + \gamma Z + \epsilon $$

However, we only observe $X$, not $Z$.  We estimate the model

$$ Y = \delta D + X \beta + u $$

In this case, we have omitted variable bias (OVB).  

Cinelli and Hazlett (2020) shows that 

$$ |\hat bias| = \sqrt{\frac{R^2_{Y \sim Z|D,X} R^2_{D \sim Z|X}}{1-R^2_{D \sim Z|X}}} (\frac{\hat \sigma_{y.dx}}{\hat \sigma_{d.x}})$$

We can see that when we have high $R^2$ between $Y$ and $Z$, bias is high.  When we have high $R^2$ between $D$ and $Z$, bias is high.  In other words, the omitted variable needs to be somewhat highly correlated with both $Y$ and $D$ to have a high bias.

Cinalli and Hazlett paper has many stats, but we focus on three things we can report:

# Partial $R^2$ of treatment on outcome

The original purpose of this stat is to see how much variation treatment explains outcome after partialling out the covariates.  

Also, this can be used for an extreme situation.  Suppose the unobserved confounder $Z$ explains all residual variance of the outcome, that is, $R^2_{Y \sim Z|D,X} = 1$.  What kind of situation we would have 0 effect?  

It is shown if  $R^2_{Y \sim Z|D,X} = 1$, then $R^2_{D \sim Z|X} \ge R^2_{Y \sim D|X}$ to make the effect go to 0.  Roughly speaking, the correlation between omitted variable and outcome needs to be higher than the correlation between treatment and outcome to make the treatment effect go to 0.

# Robust value

The robustness value $RV_{q,α}$ quantifies the minimal strength of association that the confounder needs to have, both with the treatment and with the outcome, so that a confidence interval of level $1 − \alpha$ includes a change of $q \%$ of the current estimated value.  $q$ is often 1.  That is, for the effect to go to 0, that robust value is the minimum value of the correlation between the omitted variable and the treatment and the correlation between the omitted variable and the outcome.  Note this is about both the correlation between $Z$ and $D$ and the correlation between $Z$ and $Y$.  They need to be both high to make the effect go to 0.


# Bounds using observed covariate(s)

For the first two values, they are absolute values.  You don't know whether those are large or small, without context.  They cannot be compared across studies.  

Suppose there is a covariate $x$ that you know from your domain knowledge that it is an important factor in explaining treatment or/and outcome.  Then you can compare how bad the confounder needs to be compare to this covariate. This will give us some sense how much a confounder needs to be comparing to a strong covariate, in driving the effect to 0.

$$ k_D := \frac{R^2_{D \sim Z|X_{-j}}}{R^2_{D \sim X_j|X_{-j}}} $$

where $X_{-j}$ represents all covariates other than the $j$th we pick as the benchmark covariate.

$$ k_Y := \frac{R^2_{Y \sim Z|D,X_{-j}}}{R^2_{Y \sim X_j|D,X_{-j}}} $$


# Example

Here is an example from sensemakr documentation.

```{r}

# loads package
library(sensemakr)

# loads dataset
data("darfur")

# runs regression model
model <- lm(peacefactor ~ directlyharmed + age + farmer_dar + herder_dar +
                         pastvoted + hhsize_darfur + female + village, data = darfur)

# runs sensemakr for sensitivity analysis
sensitivity <- sensemakr(model = model, 
                         treatment = "directlyharmed",
                         benchmark_covariates = "female",
                         kd = 1:3)

# short description of results
sensitivity

# long description of results
summary(sensitivity)

```
```{r}
#| results: 'asis'

# minimal reporting
ovb_minimal_reporting(sensitivity, format = "html")

```

Let's see how to interpret this.  

$RV_{q=1} = 13.9 \%$.  This means that unobserved confounders that explain 13.9% of the residual variance both of the treatment and of the outcome are sufficiently strong to explain away all the observed effect. On the other hand, unobserved confounders that do not explain at least 13.9% of the residual variance both of the treatment and of the outcome are not sufficiently strong to do so.    

The robustness value for testing the null hypothesis that the coefficient of "directlyharmed" is zero ($RV_{q=1,\alpha=0.05}$) is 7.6%. This means that unobserved confounders that explain 7.6% of the residual variance both of the treatment and of the outcome are sufficiently strong to bring the lower bound of the confidence interval to zero (at the chosen significance level of 5%).
 
 
The partial $R^2$  of directlyharmed with peacefactor means that, in an extreme scenario, in which we assume that unobserved confounders explain all of the left out variance of the outcome, these unobserved confounders would need to explain at least 2.2% of the residual variance of the treatment to fully explain away the observed effect.

The above numbers are for the absolute strength the confounder needs to be to explain away the treatment effect.   The authors believe in this context, "female" is an important factor.  Suppose the confounder is as important as "female", what the partial $R^2$ would be?  It is shown that $R^2_{Y \sim Z|X,D} = 12.5 \%$ and $R^2_{D \sim Z|X} = 0.9 \%$.  These two are both below the $RV$ values.  That means even if the confounder is as strong as "female", it is not strong enough to explain away the treatment effect.  


```{r}

# plot bias contour of point estimate
plot(sensitivity)

```

This plot shows that even 3 times as strong as "female", the confounder is not strong enough to explain away the treatment effect.  

# sensitivity analysis for IV

Even with an IV model, we need sensitivity analysis.  The worry would be there could be an unobserved confounder that affects both the instrument and the outcome.  

Cinelli and Hazlett have a package "iv.sensemakr" for this.  The usage is similar to "sensemakr". 

```{r}

# loads package
library(iv.sensemakr)

# loads dataset
data("card")

# prepares data
y <- card$lwage  # outcome
d <- card$educ   # treatment
z <- card$nearc4 # instrument
x <- model.matrix( ~ exper + expersq + black + south + smsa + reg661 + reg662 +
                     reg663 + reg664 + reg665+ reg666 + reg667 + reg668 + smsa66,
                   data = card) # covariates
# fits IV model
card.fit <- iv_fit(y,d,z,x)

# see results
card.fit

# runs sensitivity analysis
card.sens <- sensemakr(card.fit, benchmark_covariates = c("black", "smsa"))

# see results
card.sens

summary(card.sens)

```

The basic idea is to run an IV regression, then apply sensmakr to that model.  In this case, the authors use "black" and "smsa" as benchmark covariates.  The authors believe these are important factors.  A quick look at the result we can see that a confounder as strong as "black" or "smsa" is not strong enough to explain away the treatment effect.  We can also plot a contour plot to see how strong the confounder needs to be to explain away the treatment effect.

```{r}
#| results: 'asis'

plot(card.sens, parm = "rf", sensitivity.of = "t-value", lim = 0.08, kz = 1, ky = 1)
```

Unfortunately the package currently only allows for one instrument.  Should not be too hard to modify it to be more general.

