---
title: "Using machine learning for causal effect in observational study"
date: "2017-09-21"
---



## A simulation for an OLS model

In an observational study, we need to assume we have the functional form to get causal effect estimated correctly, in addtion to the assumption of treatment being exogenous.  

```{r}
#| cache: true
#| warning: false


library(MASS)
library(ggplot2)
library(dplyr)
library(tmle)
library(glmnet)
set.seed(366)

nobs <- 2000
xw <- .8
xz <- .5
zw <- .6
nrow <- 3
ncol <- 3
covarMat = matrix( c(1^2, xz^2, xw^2, xz^2, 1^2, zw^2,  xw^2, zw^2, 1^2 ) , nrow=ncol , ncol=ncol )

mu <- rep(0,3)
rawvars <- mvrnorm(n=nobs, mu=mu, Sigma=covarMat)
df <- as_tibble(rawvars, .name_repair = "minimal")
names(df) <- c('x','z','w')
df <- df %>%
    mutate(log.x=log(x^2), log.z=log(z^2), log.w=log(w^2), z.sqr=z^2, w.sqr=w^2) %>%
    mutate(g.var= log.w  + rnorm(nobs)) %>%
    mutate(A = rbinom(nobs, 1, 1/(1+exp((g.var))))) %>%
    mutate(y0=rnorm(nobs) + log.x) %>%
    mutate(tau.true = 2  + rnorm(nobs), y1=y0+tau.true, treat=A, y = treat*y1 + (1-treat)*y0)
lm1 <- lm(y ~ A + log.w + log.x , data=df)
summary(lm1)
lm2 <- lm(y ~ A , data=df)
summary(lm2)
lm3 <- lm(y ~ A + w, data=df)
summary(lm3)
lm4 <- lm(y ~ A + w + x, data=df)
summary(lm4)

```

In this example, treatment assignment process is determined by logged
w, and outcome is dtermined by logged x and treatment.  However, what
we observe is w and x.  In observational studies, this happens all the
time.  In fact, this is an ideal situation, that we observe variables
that are determinants of outcome, although we are not sure about the
functional form that determines the outcome.  However, this example
shows that unless we have observed exactly the factors themselves (in
this case logged x, w, which determines the DGP), we have biased
estimates of the true treatment effect.

Model 1 is the only model with reasonable estimate of treatment effect
(which is 2 in this case).  Model 2 is a model with endogeneity: A is
correlated with the missing variabel logged x.  Model 3 and 4 we have
x and w, but not logged, therefore still biased.

The lesson here is the functional form does matter.  However, we have
no way of knowing the functional form.  What can we do here?  


```{r}
#| cache: true
#| warning: false


Q.SL.library <- c("SL.glmnet","SL.glm","SL.glm.interaction", "SL.rpart","SL.bayesglm","SL.step","SL.mean")
g.SL.library <- c("SL.glmnet","SL.glm","SL.glm.interaction", "SL.rpart","SL.bayesglm","SL.step","SL.mean")

# this one is good since both Q and g are correct (including z in it)
tmle1 <- tmle(Y = df$y, A = df$treat, W = df[,c('x','w')], g.SL.library = g.SL.library , Q.SL.library = Q.SL.library)
tmle1
tmle2 <- tmle(Y = df$y, A = df$treat, W = df[,c('x','w', 'z')], g.SL.library = g.SL.library , Q.SL.library = Q.SL.library)
tmle2


```


We use [Mark van der Laan's TMLE method]
(http://biostats.bepress.com/ucbbiostat/paper275/).  It uses
[SuperLearner] (http://biostats.bepress.com/ucbbiostat/paper222/) as
the initial estimator.  It's an ensemble of mulitple machine learning
algorithms.  Therefore it does not need to assume the functional form
of the DGP.  Even if we don't have the variables that determines the
DGP of outcome, if we observe some functions (even nonlinear
functions) of these variables, we can still get reasonable estimates
of the treatment effect.

In this example, we used multiple popular machine learning algorithms
in modeling both treatment assingment process and the outcome process.
The first TMLE model is with x and w (note not the logged x and w
which are in the true DGP), the second one with an additional variable
z.

It seems that TMLE results are less biased than the linear models with
x and w.  It may not be better than the linear model with logged x and
w, but in empirical studies, we often cannot assume we have the
variables in the DGP, but only some proxy of the variables in the DGP.
I'll do more simulations to see whether TMLE does perform better in
the situation that we are not sure about the functional form.  We
should expect that is the case.

So far TMLE can only be used when treatment is binary variable.  

It's about time we embrace machine learning techniques into studies of
caual effect in observational studies.




