---
title: "Uplift Modeling and CATE"
author: "Xiang Ao"
date: "2025-10-10"
---

Uplift modeling estimates the incremental impact of a treatment on an
individual — who will convert *because of* the treatment, as opposed to who
would have converted anyway (or never converts regardless). It is widely used
in marketing and customer targeting.

This is the same estimand as the **conditional average treatment effect**
(CATE): $\tau(x) = E[Y(1) - Y(0) \mid X = x]$. The term "uplift" comes from
the industry; "CATE" comes from the causal inference literature. They are the
same thing.

## Python: causalml uplift trees

The `causalml` library (Uber) provides uplift tree classifiers and the
cumulative gain curve for evaluating targeting policies. The code below
follows the library's tutorial with synthetic data.

I am copying this code from https://github.com/uber/causalml/blob/master/docs/examples/uplift_trees_with_synthetic_data.ipynb.

```{python}
#| eval: false

import numpy as np
import pandas as pd

from causalml.dataset import make_uplift_classification
from causalml.inference.tree import UpliftRandomForestClassifier
from causalml.metrics import plot_gain

from sklearn.model_selection import train_test_split
import importlib
print(importlib.metadata.version('causalml') )

df, x_names = make_uplift_classification()
df.head()

# Look at the conversion rate and sample size in each group
df.pivot_table(values='conversion',
               index='treatment_group_key',
               aggfunc=[np.mean, np.size],
               margins=True)

# Split data to training and testing samples for model validation (next section)
df_train, df_test = train_test_split(df, test_size=0.2, random_state=111)

from causalml.inference.tree import UpliftTreeClassifier

clf = UpliftTreeClassifier(control_name='control')
clf.fit(df_train[x_names].values,
         treatment=df_train['treatment_group_key'].values,
         y=df_train['conversion'].values)
p = clf.predict(df_test[x_names].values)

df_res = pd.DataFrame(p, columns=clf.classes_)
df_res.head()

uplift_model = UpliftRandomForestClassifier(control_name='control')

uplift_model.fit(df_train[x_names].values,
                 treatment=df_train['treatment_group_key'].values,
                 y=df_train['conversion'].values)

df_res = uplift_model.predict(df_test[x_names].values, full_output=True)
print(df_res.shape)
df_res.head()

y_pred = uplift_model.predict(df_test[x_names].values)

# Put the predictions to a DataFrame for a neater presentation
# The output of `predict()` is a numpy array with the shape of [n_sample, n_treatment] excluding the
# predictions for the control group.
result = pd.DataFrame(y_pred,
                      columns=uplift_model.classes_[1:])
result.head()

# If all deltas are negative, assing to control; otherwise assign to the treatment
# with the highest delta
best_treatment = np.where((result < 0).all(axis=1),
                           'control',
                           result.idxmax(axis=1))

# Create indicator variables for whether a unit happened to have the
# recommended treatment or was in the control group
actual_is_best = np.where(df_test['treatment_group_key'] == best_treatment, 1, 0)
actual_is_control = np.where(df_test['treatment_group_key'] == 'control', 1, 0)

synthetic = (actual_is_best == 1) | (actual_is_control == 1)
synth = result[synthetic]

auuc_metrics = (synth.assign(is_treated = 1 - actual_is_control[synthetic],
                             conversion = df_test.loc[synthetic, 'conversion'].values,
                             uplift_tree = synth.max(axis=1))
                     .drop(columns=list(uplift_model.classes_[1:])))

plot_gain(auuc_metrics, outcome_col='conversion', treatment_col='is_treated')
```

## R: causal forest for CATE

In R, the `grf` package implements **causal forests** — the standard
nonparametric CATE estimator in the econometrics literature. The setup is
simpler than uplift trees: provide outcomes, treatments, and covariates; the
forest handles nuisance estimation internally via honest splitting.

```{r}
#| eval: false
#| echo: true

library(grf)
library(ggplot2)

# Simulate a marketing intervention
# True CATE: treatment helps high-engagement users more
set.seed(123)
n  <- 5000
X  <- matrix(rnorm(n * 5), n, 5)
colnames(X) <- paste0("x", 1:5)
W  <- rbinom(n, 1, 0.5)                          # random treatment
tau_true <- 0.3 + 0.5 * X[, 1]                   # CATE depends on x1
Y  <- tau_true * W + X[, 2] + rnorm(n, sd = 0.5) # observed outcome

# Fit causal forest
cf <- causal_forest(X, Y, W, num.trees = 2000, seed = 42)

# Estimate CATE for each unit
tau_hat <- predict(cf, estimate.variance = TRUE)
cate    <- tau_hat$predictions
cate_se <- sqrt(tau_hat$variance.estimates)

cat("Mean estimated CATE:", round(mean(cate), 3), "\n")
cat("True mean CATE:     ", round(mean(tau_true), 3), "\n")
```

```{r}
#| eval: false
#| echo: true

# Who benefits? Targeting policy
# Sort by predicted CATE, target top 30%
threshold <- quantile(cate, 0.70)
targeted  <- cate >= threshold

cat("Average CATE in targeted group:", round(mean(cate[targeted]), 3), "\n")
cat("Average CATE in non-targeted:  ", round(mean(cate[!targeted]), 3), "\n")

# Variable importance: which covariates drive heterogeneity?
imp <- variable_importance(cf)
barplot(imp, names.arg = colnames(X),
        main = "Variable importance for CATE",
        ylab = "Importance")
```

```{r}
#| eval: false
#| echo: true

# Omnibus test: is there significant CATE heterogeneity?
test_calibration(cf)
```

The `test_calibration()` output tests whether the causal forest's predictions
are meaningful — whether high-predicted-CATE units actually show higher
effects. A significant coefficient on `mean.forest.prediction` confirms that
the CATE estimates carry signal.

## Uplift vs CATE: the same thing with different vocabulary

| Uplift term | Causal inference term |
|---|---|
| Uplift score | $\hat\tau(x)$ — estimated CATE |
| Persuadables | Units with $\tau(x) > 0$ |
| Do-not-disturbs | Units with $\tau(x) < 0$ |
| Cumulative gain curve | Policy value curve / AUTOC |
| Uplift tree | Causal tree / honest tree |
| Uplift forest | Causal forest (grf) |

The Python causalml library originated in the marketing/tech industry; `grf`
originates in econometrics. They estimate the same quantity. `grf` provides
better-calibrated confidence intervals and formal tests for heterogeneity
(which are important for publication); causalml provides richer
policy-evaluation tools (gain curves, Qini coefficients) that matter for
business decisions.
