---
title: "Uplift Modeling and CATE"
author: "Xiang Ao"
date: "2025-10-10"
---

Uplift modeling is about estimating who changes because of treatment. In a
marketing example, this means separating people who buy because of the campaign
from people who would have bought anyway.

This is the same object as the conditional average treatment effect (CATE),
$$\tau(x) = E[Y(1) - Y(0) \mid X = x].$$
The word "uplift" is more common in industry. The word "CATE" is more common in
the causal inference literature.

## Python: causalml uplift trees

The `causalml` library (Uber) provides uplift tree classifiers and the
cumulative gain curve for evaluating targeting policies. The code below
follows the library's tutorial with synthetic data.

I am copying this code from https://github.com/uber/causalml/blob/master/docs/examples/uplift_trees_with_synthetic_data.ipynb.

```{python}
#| eval: false

import numpy as np
import pandas as pd

from causalml.dataset import make_uplift_classification
from causalml.inference.tree import UpliftRandomForestClassifier
from causalml.metrics import plot_gain

from sklearn.model_selection import train_test_split
import importlib
print(importlib.metadata.version('causalml') )

df, x_names = make_uplift_classification()
df.head()

# Look at the conversion rate and sample size in each group
df.pivot_table(values='conversion',
               index='treatment_group_key',
               aggfunc=[np.mean, np.size],
               margins=True)

# Split data to training and testing samples for model validation (next section)
df_train, df_test = train_test_split(df, test_size=0.2, random_state=111)

from causalml.inference.tree import UpliftTreeClassifier

clf = UpliftTreeClassifier(control_name='control')
clf.fit(df_train[x_names].values,
         treatment=df_train['treatment_group_key'].values,
         y=df_train['conversion'].values)
p = clf.predict(df_test[x_names].values)

df_res = pd.DataFrame(p, columns=clf.classes_)
df_res.head()

uplift_model = UpliftRandomForestClassifier(control_name='control')

uplift_model.fit(df_train[x_names].values,
                 treatment=df_train['treatment_group_key'].values,
                 y=df_train['conversion'].values)

df_res = uplift_model.predict(df_test[x_names].values, full_output=True)
print(df_res.shape)
df_res.head()

y_pred = uplift_model.predict(df_test[x_names].values)

# Put the predictions to a DataFrame for a neater presentation
# The output of `predict()` is a numpy array with the shape of [n_sample, n_treatment] excluding the
# predictions for the control group.
result = pd.DataFrame(y_pred,
                      columns=uplift_model.classes_[1:])
result.head()

# If all deltas are negative, assing to control; otherwise assign to the treatment
# with the highest delta
best_treatment = np.where((result < 0).all(axis=1),
                           'control',
                           result.idxmax(axis=1))

# Create indicator variables for whether a unit happened to have the
# recommended treatment or was in the control group
actual_is_best = np.where(df_test['treatment_group_key'] == best_treatment, 1, 0)
actual_is_control = np.where(df_test['treatment_group_key'] == 'control', 1, 0)

synthetic = (actual_is_best == 1) | (actual_is_control == 1)
synth = result[synthetic]

auuc_metrics = (synth.assign(is_treated = 1 - actual_is_control[synthetic],
                             conversion = df_test.loc[synthetic, 'conversion'].values,
                             uplift_tree = synth.max(axis=1))
                     .drop(columns=list(uplift_model.classes_[1:])))

plot_gain(auuc_metrics, outcome_col='conversion', treatment_col='is_treated')
```

## R: causal forest for CATE

In R, the `grf` package implements causal forests. The setup is simpler than
the uplift tree example: provide outcomes, treatments, and covariates, and the
forest estimates CATEs using honest splitting.

```{r}
#| eval: false
#| echo: true

library(grf)
library(ggplot2)

# Simulate a marketing intervention
# True CATE: treatment helps high-engagement users more
set.seed(123)
n  <- 5000
X  <- matrix(rnorm(n * 5), n, 5)
colnames(X) <- paste0("x", 1:5)
W  <- rbinom(n, 1, 0.5)                          # random treatment
tau_true <- 0.3 + 0.5 * X[, 1]                   # CATE depends on x1
Y  <- tau_true * W + X[, 2] + rnorm(n, sd = 0.5) # observed outcome

# Fit causal forest
cf <- causal_forest(X, Y, W, num.trees = 2000, seed = 42)

# Estimate CATE for each unit
tau_hat <- predict(cf, estimate.variance = TRUE)
cate    <- tau_hat$predictions
cate_se <- sqrt(tau_hat$variance.estimates)

cat("Mean estimated CATE:", round(mean(cate), 3), "\n")
cat("True mean CATE:     ", round(mean(tau_true), 3), "\n")
```

```{r}
#| eval: false
#| echo: true

# Who benefits? Targeting policy
# Sort by predicted CATE, target top 30%
threshold <- quantile(cate, 0.70)
targeted  <- cate >= threshold

cat("Average CATE in targeted group:", round(mean(cate[targeted]), 3), "\n")
cat("Average CATE in non-targeted:  ", round(mean(cate[!targeted]), 3), "\n")

# Variable importance: which covariates drive heterogeneity?
imp <- variable_importance(cf)
barplot(imp, names.arg = colnames(X),
        main = "Variable importance for CATE",
        ylab = "Importance")
```

```{r}
#| eval: false
#| echo: true

# Omnibus test: is there significant CATE heterogeneity?
test_calibration(cf)
```

The `test_calibration()` output is a useful diagnostic. It checks whether units
with higher predicted CATEs also appear to have higher treatment effects.

## Uplift vs CATE: the same thing with different vocabulary

| Uplift term | Causal inference term |
|---|---|
| Uplift score | $\hat\tau(x)$ — estimated CATE |
| Persuadables | Units with $\tau(x) > 0$ |
| Do-not-disturbs | Units with $\tau(x) < 0$ |
| Cumulative gain curve | Policy value curve / AUTOC |
| Uplift tree | Causal tree / honest tree |
| Uplift forest | Causal forest (grf) |

The Python `causalml` package comes from the marketing/tech vocabulary. `grf`
comes from the econometrics vocabulary. They are trying to estimate the same
quantity. For publication, I would usually want the inference and diagnostics
from `grf`. For a targeting exercise, the gain-curve tools in `causalml` are
also useful.
