---
title: "Matching and Weighting - Part 2: weighting"
date: "2025-05-31"
---


This is based on R's MatchIt and WeightIt packages and their documentations.  All credits go to Noah Greifer and coauthors.  I write to summarize what I have learned.

## Assumptions

Matching or weighting is usually used in observational studies.  We have the following assumptions before using matching or weighting to make a causal claim:

1. SUTVA
2. ignorability (unconfoundedness, or no unobserved confounders)
3. overlap (common support, or positivity)

With these assumptions, we can try to identify the treatment effect in an observational study.

## Problem

The second assumption is saying controlling for $X$, the treatment assignment $D$ is independent of the potential outcomes $Y(0)$ and $Y(1)$, i.e., we have a pseudo-randomized experiment.

In reality, it's rare we have a balanced distribution of $X$ for treated and control.  That's when we need to use matching or weighting to balance the distribution of $X$ between treatment and control groups.

## weighting methods

The idea of weighting is to use weights to creat a pseudo population which has a balanced distribution of $X$ for treated and control groups.  The most popular weighting method is inverse probability of treatment weighting (IPTW).  IPTW uses the propensity score to create weights.

There are a few other weighting methods that are gaining popularity.

### Inverse probability of treatment weighting (IPTW)

IPTW is the most common weighting method.  It uses the inverse of the propensity score as weights.  For the **ATE**, weight each treated unit by $1/p$ and each control unit by $1/(1-p)$, where $p$ is the propensity score.  For the **ATT** --- which is what the examples below request via `estimand = "ATT"` --- the treated units all get weight $1$ and the controls get $p/(1-p)$; you can see this in the weight summaries, where the treated range is exactly $1$ to $1$.  Either way the point is the same: create a pseudo population in which the distribution of $X$ is balanced between the groups.

### Covariate balancing propensity score (CBPS)

The same idea, but with a different way to estimate the propensity score.  CBPS is a form of logistic regression where balance constraints are incorporated to a generalized method of moments estimation of the model coefficients.

### Entropy balancing

Entropy balancing involves the specification of an optimization problem, the solution to which is then used to compute the weights. The constraints of the primal optimization problem correspond to covariate balance on the means (for binary and multi-category treatments) or treatment-covariate covariances (for continuous treatments), positivity of the weights, and that the weights sum to a certain value.  Entropy balancing is doubly robust (for the ATT) in the sense that it is consistent either when the true propensity score model is a logistic regression of the treatment on the covariates or when the true outcome model for the control units is a linear regression of the outcome on the covariates, and it attains a semi-parametric efficiency bound when both are true. Entropy balancing will always yield exact mean balance on the included terms (not necessarily on the distribution of the covariates).

### Energy balancing

Energy balancing is a method of estimating weights using optimization without a propensity score.

The primary benefit of energy balancing is that all features of the covariate distribution are balanced, not just means, as with other optimization-based methods like entropy balancing.

### SuperLearner

SuperLearner works by fitting several machine learning models to the treatment and covariates and then taking a weighted combination of the generated predicted values to use as the propensity scores, which are then used to construct weights.


## Example

Here is an example with Lalonde data, we are interested in the effect of treatment on "re78" (real earnings 1978).

### IPW
```{r}
#| include: true

library("WeightIt")
library(cobalt)
data(lalonde)
w.out1 <- weightit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                  data = lalonde, estimand = "ATT", method = "glm")
w.out1 #print the output
summary(w.out1)
```
```{r}
#| include: true

library(cobalt)
bal.tab(w.out1, stats = c("m", "v"), thresholds = c(m = .05))
```

### entropy balancing

```{r}
#| include: true

w.out2 <- weightit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                  data = lalonde, estimand = "ATT", method = "ebal")
summary(w.out2)
bal.tab(w.out2, stats = c("m", "v"), thresholds = c(m = .05))
```
Suppose we are satisfied with the balance, we can use the weights to estimate the treatment effect.  We can use the "marginaleffects" package to do that.

```{r}

# Fit outcome model
fit <- lm_weightit(re78 ~ treat * (age + educ + race + married +
                                     nodegree + re74 + re75),
                   data = lalonde, weightit = w.out2)
# G-computation for the treatment effect
library("marginaleffects")
avg_comparisons(fit, variables = "treat",
                newdata = subset(lalonde, treat == 1))
```

### energy balancing

```{r}
#| include: true

w.out3 <- tryCatch(
  weightit(treat ~ age + educ + race + married +
                  nodegree + re74 + re75, data = lalonde,
                method = "energy", estimand = "ATE"),
  error = function(e) { message("Energy balancing failed: ", e$message); NULL }
)
if (!is.null(w.out3)) summary(w.out3)
```

```{r}
#| include: true

if (!is.null(w.out3)) {
  fit <- lm_weightit(re78 ~ treat * (age + educ + race + married +
                                       nodegree + re74 + re75),
                     data = lalonde, weightit = w.out3)
  # G-computation for the treatment effect
  avg_comparisons(fit, variables = "treat")
}

```

Two things are worth noticing in that output rather than passing over it. First,
the estimand changed: `w.out3` requests the **ATE**, while the entropy-balancing
fit above targeted the **ATT**, so the two numbers ($-73$ against $1273$) are not
competing estimates of the same quantity. Second, look at the weight summary.
Entropy balancing gives every treated unit a weight of exactly $1$, as the ATT
requires. Energy balancing spreads the treated weights from $0$ to $14.0$ --- one
treated unit is effectively discarded and another counts as fourteen --- and the
resulting standard error ($1085$) is much larger than entropy balancing's ($770$).
Balancing the whole covariate *distribution* rather than just its means is a
genuine advantage, but it is bought with weight variability, and that shows up
directly in the precision of the estimate.

## comparing matching and weighting

Matching and weighting are popular in different fields.  We used to say matching involves discarding data therefore changing the estimand, but now with full matching and other methods, matching can be done without losing data.  We used to believe weighting can be unstable, as weights can be highly variable, but with CBPS, entropy balancing and energy balancing, weighting can perform much better.

So, in practice, we can try different methods, either matching or weighting, and see which one gives us better balance. Then use "marginaleffects" to get the treatment effect.

---

<!-- see-also-footer -->
*Systematic treatment: [R](https://xiangao.github.io/causal_econometrics_guide/matching.html) · [Julia](https://xiangao.github.io/causal_econometrics_julia/matching.html).*
