---
title: "Weights in OLS in causal inference"
date: "2025-01-16"
---


## Unit level weights for OLS

This chapter follows Hazlett and Shinkre, “Demystifying and Avoiding the OLS 'Weighting Problem': Unmodeled Heterogeneity and Straightforward Solutions.”

Under unconfoundedness, the identifying assumption holds at each level of $X$. When $X$ is discrete and low-dimensional, we can stratify and compute the ATE by the g-formula. In practice we usually run a linear regression

$$ Y = \alpha + \beta D + \gamma X + \epsilon $$ {#eq-weights-ols-1}

and treat $\hat\beta$ as the treatment effect, relying on linearity to span the covariate space. The question is what this estimator is actually weighting.

Suppose we are interested in the ATE of $D$ on $Y$.  Say we have discrete variables $X$.  Under unconfoundedness, the ATE is


$$
\begin{aligned}
 \tau_{ATE} &= \sum E[Y(1) - Y(0) | X = x] P(X = x) \\
            &= \sum (E[Y|D=1, X=x]- E[Y|D=0, X=x]) P(X=x) \\
            &= \sum DIM_x P(X=x) \\
\end{aligned}
$$ {#eq-weights-ols-2}


This is basically the g-formula, or regression adjustment formula.  We compute the DIM (difference in means) between treated and control for each strata $X=x$, then the weighted average is the ATE, with $\hat P(X=x)$ as weights.

If we have high dimensional $X$, we usually do
$$ Y = \beta_0 + \beta_1 X + \delta_{reg} D + \epsilon $$ {#eq-weights-ols-3}

We would hope $\delta_{reg}$ is close to $\delta_{ATE}$, but we are not sure.

OLS coefficients are:

$$ \hat \beta = (Z'Z)^{-1} Z'y $$ {#eq-weights-ols-4}

Here I use $Z$ to represent all regressors.  Each coefficient is a weighted sum of the outcome,

$$ \hat \tau_{reg} = \sum \omega_i Y_i $$ {#eq-weights-ols-5}

By FWL theorem,

$$ \hat \tau_{reg} = \frac{\sum Y_i (D_i - \hat d(X_i)) }{\sum (D_i - \hat d(X_i))^2} $$ {#eq-weights-ols-6}

Here $\hat d(X_i) = \hat p(D_i=1|X_i)$ which is the predicted value of linear regression of $D$ on $X$.  The numerator is the covariance between $Y$ and residualized $D$, the denominator is the variance of the residualized $D$.  Note that the original FWL would have residualized $Y$ here, but it turns out it works the same with the original $Y$.

Therefore the weights are


$$
\begin{aligned}
\omega_i &= \frac{D_i - \hat d(X_i)}{\sum (D_i - \hat d(X_i))^2} \\
\end{aligned}
$$ {#eq-weights-ols-7}



Note for the denominator, it's a constant.  It does *not* scale the weights to sum to one --- they sum to **zero**, because $D_i - \hat d(X_i)$ is an OLS residual from a regression that includes an intercept.  What the denominator normalizes is the contrast: since $\sum D_i(D_i - \hat d(X_i)) = \sum (D_i - \hat d(X_i))^2$ (the residual is orthogonal to the fitted values), we get $\sum_i \omega_i D_i = 1$, i.e. the weights on the treated sum to $+1$ and those on the control sum to $-1$.  That is what makes $\sum_i \omega_i Y_i$ a weighted *difference* of means.  The numerator for treated units are $1-\hat d(X_i)$, for control units it's $-\hat d(X_i)$ (negative).  It is similar to IPW weights, which are $1/\hat d(X_i)$ for treated and $1/(1-\hat d(X_i))$ for control.  These two sets of weights go the same direction.  For the treated, the higher probability of being treated, the lower weight; for the control, the higher probability of being treated, the larger the magnitude of the (negative) weight.

If these weights are the same as the IPW weights, then we'll get $\tau_{reg}=\tau_{ATE}$.  However, they are usually different.

Another way to write this as weighted difference in means:

$$ \hat \tau_{wdim} = \sum \omega_i Y_i = \sum_{i:D=1} \tilde \omega_i Y_i + \sum_{i:D=0} \tilde \omega_i  Y_i $$ {#eq-weights-ols-8}

where, with $c = \sum (D_i - \hat d(X_i))^2$, the per-arm weights are $\tilde \omega_i = (1-\hat d(X_i))/c$ for treated units ($D_i=1$) and $\tilde \omega_i = -\hat d(X_i)/c$ for control units ($D_i=0$)

Because $\hat d(X_i)$ is from a linear probability model, it's possible that it's not between 0 and 1.  In that case, we can have negative weights.  The negative weights could mean the observations do not have a match in the other group.  For example, $\hat d(X_i) > 1$ will cause the weight $(1-\hat d(X_i))/c$ to flip sign and become negative for a treated unit.  That means this unit is more like treated units than any control units.  In other words, there is no match in the control group.  The mirror image happens for a control unit with $\hat d(X_i) < 0$: its weight $-\hat d(X_i)/c$ flips to *positive*, so it enters the contrast on the same side as the treated.  Either way the unit contributes with the wrong sign.

The negative weights are invisible unless we examine $\hat d(X_i)$ directly — which is to say, unless we look at the propensity score. A linear regression of $Y$ on $D$ and $X$ hides the extrapolation. So $\hat\tau_{\text{reg}}$ is a weighted average of outcomes with weights that differ from the ATE weights, and some of those weights may carry the wrong sign without the researcher knowing.

##  Strata level weights

Angrist ([1998](https://doi.org/10.2307/2999578), *Econometrica*) has the strata-wise weights for OLS:

$$ \hat \tau_{reg} = \frac{\sum_x \hat \tau_x \hat d(x) (1-\hat d(x)) \hat P(X=x)}{\sum_x  \hat d(x) (1-\hat d(x)) \hat P(X=x)} $$ {#eq-weights-ols-9}

this shows that different from ATE, these weights are $\hat d(x) (1-\hat d(x)) \hat P(X=x)$.  The stratas are weighted not only by $P(X_i=x)$, but also by the variance of treatment status.  When $\hat d(X_i)$ is close to .5, the weight is the largest.

If there is no treatment effect heterogeneity, then it does not matter how you weight.  When there is treatment effect heterogeneity, this estimator could be very different from ATE.

Hazlett and Shinkre show that Angrist's formula is a special case: it is exact when the probability of treatment is linear in $X$ (as in a saturated model with discrete $X$). They derive a more general formula, but the Angrist version suffices under linearity of $d(X)$.


## Recommended estimation approaches

Under constant effects and linearity, $E[Y \mid X, D] = \beta_0 + \beta_1 X + \beta_2 D$, a regression of $Y$ on $D$ and $X$ recovers the ATE without further adjustment. When the linearity or constant-effects assumption fails, Hazlett and Shinkre suggest relaxing to *separate* linearity — allowing different intercepts for $Y(0)$ and $Y(1)$. Under that weaker assumption, several estimators are unbiased for the ATE.

The interaction (Lin [2013](https://doi.org/10.1214/12-AOAS583)) estimator regresses $Y$ on $D$, $\tilde X = X - \bar X$, and $D \tilde X$; the coefficient on $D$ is the ATE. The imputation estimator fits separate regressions for treated and control, predicts counterfactual outcomes for the other group, and takes the average difference — algebraically, it returns the same estimate. Stratification and g-computation do the same for discrete $X$: compute the stratum-level difference in means and weight by the stratum probability.

All of these still rely on linearity within each arm. When that fails, matching (e.g. mean-balancing) or nonparametric estimation is needed.

## A worked example

We make the regression weights $\omega_i$ concrete and compare them to IPW weights $1/\hat d(X_i)$ (treated) and $1/(1-\hat d(X_i))$ (control).

```{r}
#| label: fwl-weights-example
#| cache: true
#| warning: false
#| message: false
library(dplyr)

set.seed(1)
n <- 500
X <- rnorm(n)
d_prob <- plogis(0.5 * X)
D <- rbinom(n, 1, d_prob)
Y <- 1 + 2 * D + 1.5 * X + rnorm(n)
dat <- data.frame(Y, D, X)

# Regression-implied weights via the FWL residualization of D on X.
d_hat <- fitted(lm(D ~ X, data = dat))
resid_D <- dat$D - d_hat
omega_reg <- resid_D / sum(resid_D^2)

# IPW weights (Horvitz-Thompson), using the same linear-probability d_hat
# purely for comparability with the regression weights above.
ipw <- ifelse(dat$D == 1, 1 / d_hat, 1 / (1 - d_hat))
# rescale by a single constant so the treated weights average 1; this only
# fixes the axis scale of the plot below and does not change any comparison
ipw <- ipw / sum(ipw[dat$D == 1]) * sum(dat$D == 1)

tau_reg <- sum(omega_reg * dat$Y)
tau_lm  <- coef(lm(Y ~ D + X, data = dat))["D"]

data.frame(tau_reg_from_weights = tau_reg, tau_from_lm_coef = tau_lm)
```

The two are numerically identical (up to floating point), confirming $\hat\tau_{reg} = \sum \omega_i Y_i$. Plotting `omega_reg` against `ipw` for treated and control units separately shows the two weighting schemes move in the same direction but are not proportional — confirming that $\hat\tau_{reg} \neq \hat\tau_{ATE}$ in general, as discussed above.

```{r}
#| label: fwl-weights-plot
#| cache: true
#| warning: false
#| message: false
par(mfrow = c(1, 2))
plot(ipw[dat$D == 1], omega_reg[dat$D == 1],
     xlab = "IPW weight", ylab = "Regression (FWL) weight",
     main = "Treated units")
plot(ipw[dat$D == 0], omega_reg[dat$D == 0],
     xlab = "IPW weight", ylab = "Regression (FWL) weight",
     main = "Control units")
par(mfrow = c(1, 1))
```

Note also what this example does *not* show. Here $\hat d(X_i)$ stays inside
$[0.06, 0.99]$, so every weight keeps its expected sign and the negative-weight
problem discussed above never materializes. That is a property of this particular
design, not a reassurance. Strengthen the linear index and the linear probability
model starts predicting outside the unit interval:

```{r}
#| label: fwl-negative-weights
#| cache: true
#| warning: false
#| message: false
set.seed(2)
X2 <- rnorm(n)
D2 <- rbinom(n, 1, plogis(3 * X2))       # much stronger selection on X
dat2 <- data.frame(D = D2, X = X2)
d_hat2 <- fitted(lm(D ~ X, data = dat2))
omega2 <- (dat2$D - d_hat2) / sum((dat2$D - d_hat2)^2)

data.frame(
  dhat_below_0        = sum(d_hat2 < 0),
  dhat_above_1        = sum(d_hat2 > 1),
  treated_wrong_sign  = sum(dat2$D == 1 & omega2 < 0),
  control_wrong_sign  = sum(dat2$D == 0 & omega2 > 0)
)
```

Those units are the ones with no counterpart in the other arm, and a plain
regression of $Y$ on $D$ and $X$ gives you no indication that they are there.











